---
title: "Spread"
output:
  html_document: default
  pdf_document: default
date: "2026-03-19"
editor_options:
  markdown:
    wrap: 72
---

# SPREAD

## 1. Data setup

```{r setup, include=FALSE}
getwd()
library(urca)
library(tseries)
library(quantmod)
library(forecast)
library(rugarch)
library(FinTS)
library(ggplot2)
library(dplyr)
library(httpgd)
library(R.utils)
library(xts)
library(readxl)
library(fBasics)
library(lmtest)
library(sandwich)
library(fUnitRoots)
library(stats)
library(graphics)
library(MTS)
library(MASS)
library(zoo)
library(forecast)
library(seastests)
library(strucchange)


```

```{r, eval=FALSE}
# ============================================================
# MACOS OPTIMIZATION BLOCK (Apple Silicon)
# ============================================================

# 1. Verifica che R sia nativo ARM64 (non Rosetta 2)
if (R.version$arch != "aarch64") {
  warning("R non sta girando su ARM64. Reinstalla R nativo per Apple Silicon da cran.r-project.org")
} else {
  message("✓ R ARM64 nativo confermato: ", R.version$arch)
}

# 2. Configura parallelismo ottimale per macOS (usa FORK, molto più veloce di SOCKET)
library(parallel)
N_CORES <- detectCores(logical = FALSE) - 1  # core fisici, lascia 1 libero
message("✓ Core disponibili per parallelismo: ", N_CORES)

# 3. RNG riproducibile per simulazioni parallele
RNGkind("L'Ecuyer-CMRG")
set.seed(42)

# 4. Funzione helper: crea cluster FORK (macOS only)
make_mac_cluster <- function() {
  makeCluster(N_CORES, type = "FORK")  # FORK evita di riesportare variabili
}

# 5. Libera memoria dopo operazioni pesanti
gc_after <- function(expr) {
  result <- force(expr)
  gc(verbose = FALSE)
  result
}

message("✓ Blocco ottimizzazione macOS caricato.")
# ============================================================

```
Load the data and parse from excel file, weekly spread.

```{r, include=FALSE}
rm(list= ls())
getwd()
#setwd("~/Desktop")
data <- read_excel("Spread_daily_10_04.xlsx")
data <- na.omit(data)
# Cambia segno solo se la media è negativa
data$`IT10DE10=TWEB (BID)` <- as.numeric(data$`IT10DE10=TWEB (BID)`) * -1
data$Date <- as.Date(data$Date)
spread <- zoo(data$`IT10DE10=TWEB (BID)`, order.by = data$Date)
```

Inspect data

```{r}
head(data)
str(data)
sum(is.na(data))
any(is.infinite(spread))
```

Plot the data

```{r,echo=FALSE}
plot(spread, main="BTP-BUND Spread (Daily)", ylab="Spread (bps)", xlab= "Date")
```

```{r, echo=FALSE}
options(scipen = 999)
basicStats(spread)
```

The descriptive analysis of the daily BTP-BUND spread reveals a distribution with a mean of 141.05 bps and a median of 134.65 bps. The fact that the mean remains above the median confirms a persistent positive skewness, indicating that the distribution is influenced by periodic upward shocks rather than frequent downward drops.

Visual inspection of the daily time series identifies three distinct market regimes:

The 2022 Surge: A sharp escalation where the spread moved from its minimum of 59.80 bps toward the series maximum of 250.70 bps, likely reflecting heightened political or macroeconomic uncertainty.

The 2023–2024 Volatility: A period of consolidation characterized by significant fluctuations, yet largely contained within the interquartile range of 105.50 bps to 179.03 bps.

The 2025–2026 Mean Reversion: A sustained downward trend where the spread compressed significantly, approaching the lower bounds of the sample before a slight uptick at the end of the period.

Econometrically, the daily frequency highlights the "jump-diffusion" nature of sovereign spreads. While the market spends the majority of its time in a state of relative calm—as evidenced by the median and the 1st quartile of 105.50 bps—it remains susceptible to rapid "tail risk" events. These episodes of stress pull the mean upward, creating the characteristic right-skewed profile typical of risk-sensitive financial instruments.

**Forecast n°1 : naive forecast (mean)**

```{r, include=FALSE}
mean <- mean(data$`IT10DE10=TWEB (BID)`, na.rm=TRUE)
```

```{r, echo=FALSE}
plot(spread, main="BTP-BUND Spread (Weekly", ylab="Spread (bps)", xlab= "Date")
abline(h = mean, col="red", lwd=2, lty=2)
text(x = start(spread), y= mean,
     labels=paste("Mean =", round(mean, 2)),
     col="red", adj= c(0, -0.5))
```

## 2. Test for seasonality

We use seastests package

```{r}
start_date<- index(spread)[1]
library(lubridate)
start_year<-year(start_date)
start_month<-month(start_date)
start_day <-day(start_date)
spread_ts <- ts(data$`IT10DE10=TWEB (BID)`, start=c(start_year,start_month,start_day), frequency = 252)#annual frequency
isSeasonal(spread_ts, test="combined")
#We get FALSE

#Individual tests
isSeasonal(spread_ts, test="qs")
isSeasonal(spread_ts, test="seasdum")
isSeasonal(spread_ts, test="kw")
isSeasonal(spread_ts, test="fried")

#We get all FALSE

## Now plot ACF and PACF
Acf(spread_ts, lag.max= 52)
Pacf(spread_ts, lag.max = 52)

#We have 1 significant lag at 1 in the Pacf, and a Acf which gradually goes to 0
#This suggests no seasonality.

#Last point, let's fit an arima model

adf.test(spread_ts)     # stationarity test -> we have p-value = 0.546, so we fail to reject the null, so we are in the null workspace, this means no seasonality.

kpss.test(spread_ts) # stationarity test where the null hypothesis is that the series IS stationary
# p-value is 0.01 -> we REJECT the null, so the series is NOT stationary

```

## Visualization for the no-seasonality

```{r seasonal_analysis, echo=FALSE}

# 1. Seasonal Plot 
# Mostra l'andamento dello spread sovrapponendo i periodi (2021-2026)
ggseasonplot(spread_ts, year.labels = TRUE, year.labels.left = TRUE) +
  theme_minimal() +
  labs(title = "Seasonal Effect: BTP-BUND (Annual Pattern)",
       subtitle = "Daily data (2021-2026) - Annual Cycle",
       x = "Day of the Year (Trading Days)",
       y = "Spread (bps)",
       color = "Year")

# 2. Subseries Plot (Medie Mensili)
# Nota: Con freq = 252, il subseries plot potrebbe tentare di fare 252 piccoli grafici.
# Per vedere le medie mensili "storiche", è meglio aggregare prima i dati o 
# usare una frequenza mensile (12) se l'obiettivo è l'analisi macro.
# Se vuoi comunque il subseries sulle 52 settimane:
#spread_weekly_avg <- aggregate(spread_ts, nfrequency = 52)
ggsubseriesplot(spread_ts) +
  theme_minimal() +
  labs(title = "Subseries Plot: Daily data",
       subtitle = " 2021-2026",
       x = "Week of the Year",
       y = "Spread (bps)")

# 3. STL Decomposition (Multi-Year)
# Utilizziamo un s.window ampio per catturare la stagionalità annuale
decomposed <- stl(spread_ts, s.window = "periodic", robust = TRUE)

autoplot(decomposed) +
  theme_minimal() +
  labs(title = "STL Decomposition (2021-2026)",
       subtitle = "Trend, Annual Seasonality, and Daily Remainder")
```

-->Seasonal and Subseries Plot Interpretation
The seasonal plots further investigate whether specific times of the year are inherently riskier for the Italian spread:

Subseries Plot (Weekly): The blue dashed lines represent the historical mean for each specific week of the year across the 2021-2026 period. The relatively flat nature of these mean lines suggests that there is no "structural" week of the year where the spread is consistently higher. The high density of the black vertical lines reflects the daily volatility within those weeks.

Annual Pattern (Seasonal Plot): By overlaying the years, we can see that the 2022-2024 period significantly outperformed the historical averages, creating the "Surge" identified in the descriptive statistics. The 2025-2026 lines are positioned lower, confirming the transition into a period of market calm.

-->The STL (Seasonal and Trend decomposition using Loess) plot provides a granular look at the forces driving the spread:

Trend: The trend component shows a steady appreciation starting in 2021, reaching a peak during the 2024-2025 period. Notably, toward the beginning of 2026, the trend exhibits a clear mean-reversion phase, with the spread beginning to compress. This suggests that long-term macroeconomic factors were the primary drivers of the spread's expansion and subsequent cooling.

Seasonal: The annual seasonal component is present but relatively low in magnitude (oscillating between -10 and +15 bps). This indicates that while there are minor "calendar effects," they are secondary to the overall trend and unpredictable market shocks.

Remainder: This component represents "noise" or exogenous shocks. We observe significant density and volatility in the remainder during the 2024-2025 period. These spikes represent "tail risk" events—sudden political or financial news that caused the spread to jump independently of any seasonal or long-term trend.

**Econometric Summary**
From an econometric perspective, the BTP-BUND spread during this period is a trend-dominated series. The influence of seasonality is statistically marginal compared to the impact of the Remainder (shocks). This confirms that the spread is more sensitive to "news arrival" and discretionary policy changes than to cyclical or seasonal time-effects.

## 3. Model Specification: Levels vs. Logarithmic Transformations\*\*

A critical decision in modelling the BTP-BUND spread involves the choice
between raw levels (basis points) and logarithmic transformations. While
log-returns are standard for equity prices to ensure normality and
constant variance, this study follows the established econometric
literature—notably **Campbell, Lo, and MacKinlay (1997)** and **Hamilton
(1994)**—by analyzing the spread in **levels**.

As argued by **Engle (1982)** in his seminal work on ARCH models,
interest rate differentials are already expressed in percentage terms;
therefore, a log-transformation is not only redundant but can lead to
significant numerical instability as values approach the **Zero Lower
Bound (ZLB)**. Maintaining the analysis in basis points ensures the
direct economic interpretability of the results, which is essential for
assessing credit and liquidity risk premiums.

#### Literature Review on Model Specification

The choice to model the BTP-BUND spread in first differences of its
levels is supported by the foundational empirical work on Eurozone
sovereign risk. **Codogno, Favero, and Missale (2003)** and **Favero et
al. (2010)** demonstrate that interest rate differentials are best
captured in basis points to maintain linear properties between credit
risk and liquidity factors. Furthermore, the application of a **GARCH
framework**, as pioneered by **Bollerslev (1986)**, ensures that the
time-varying volatility of the spread is explicitly parameterized rather
than suppressed via logarithmic transformations.

### 3.1 Data Partitioning: Training and Test Sets

To evaluate the predictive performance of the **ARIMA model**, the
dataset is partitioned into training and testing subsets. The training
phase utilizes a **rolling window approach** with a fixed length of 10
years, corresponding to **520 weekly observations**.

This specific configuration ensures that the model is estimated on a
sufficiently long historical period while remaining sensitive to more
recent structural dynamics in the BTP-BUND spread. The architecture of
the script is designed for **parametric flexibility**: the training
window duration can be easily modified by adjusting the
`$n_train_window$` constant (e.g., replacing 520 with a different value)
to test the model's robustness across different look-back periods.

```{r}
n_test <- length(spread)*0.2
n_train_window <- length(spread)-n_test 

# Lunghezza totale della serie differenziata
total_obs <- length(spread)

# 1. Definizione degli indici per il TEST (le ultime 10)
test_start <- total_obs - n_test + 1
test_end   <- total_obs
test       <- spread[test_start:test_end]

# 2. Definizione degli indici per il TRAIN (i 10 anni precedenti al test)
# Il train deve finire esattamente dove inizia il test
train_end   <- test_start - 1
train_start <- train_end - n_train_window + 1

# Estrazione del train
train<- spread[train_start:train_end]

# Verifica coerenza
print(paste("Lunghezza Train:", length(train))) # Deve essere 520
print(paste("Lunghezza Test:", length(test)))       # Deve essere 10
# Creiamo un piccolo data frame per il plot
df_plot <- data.frame(
  Settimana = 1:length(spread),
  Valore = as.numeric(spread)
)
ggplot(df_plot, aes(x = Settimana, y = Valore)) +geom_line(color = "steelblue") +
geom_vline(xintercept = length(train), color = "red", linetype = "dashed",
linewidth = 1) +labs(title = "BTP-BUND Spread: Training vs Test Set",subtitle = paste("Punto di taglio all'osservazione:", length(train)),
    x = "Settimane",y = "Variazione Spread (bps)")+theme_minimal()

cat(sprintf("Training observations: %d\n", length(train)))
cat(sprintf("Test observations:     %d\n", length(test)))
cat(sprintf("Train period: %s -> %s\n", index(train)[1], index(train)[length(train)]))
cat(sprintf("Test period:  %s -> %s\n", index(test)[1],  index(test)[length(test)]))
```

### 3.3 Stationarity and Unit Root Testing

To ensure the statistical validity of the ARIMA-GARCH framework, we
first evaluate the integration order of the series. Financial variables
often exhibit non-stationary behavior, characterized by a "random walk".

#### 3.3.1 Augmented Dickey-Fuller (ADF) Test

We perform the ADF test to check for the presence of a unit root
($H_0$).

```{r}
par(mfrow = c(1, 2))
acf(as.numeric(train))
Pacf(as.numeric(train))
par(mfrow = c(1, 1)) 
```

### 3.4 Persistence and Autocorrelation Analysis

The initial inspection of the **Autocorrelation Function (ACF)** reveals
a very slow, near-linear decay. In time series econometrics, this
pattern is a classic diagnostic for a **"near-unit root"** process,
indicating that the AR component is highly persistent and the series is
likely non-stationary.

Regarding the optimal number of lags for the model, while empirical
rules of thumb exist, this study adopts an **automated selection
procedure** based on the minimization of Information Criteria,
specifically the **Akaike Information Criterion (AIC)** and the
**Bayesian Information Criterion (BIC)**. This approach ensures an
optimal balance between model fit and parsimony, mitigating the risks of
both underfitting (omitted variable bias) and overfitting (inflated
variance).

### 3.5 Formal Stationarity Testing: The ADF Approach

To rigorously confirm the integration order of the BTP-BUND spread, we
employ the **Augmented Dickey-Fuller (ADF) test** using the `ur.df()`
function from the `urca` package. Unlike standard implementations, this
approach requires a direct comparison between the calculated test
statistics ($\tau$) and the critical values provided by **Dickey and
Fuller (1979)**.

#### 3.5.1 Test Specification and Hypotheses

The test is conducted following a hierarchical strategy, accounting for
different deterministic components:

1.  Trend (`type = "trend"`): To check if the series is stationary
    around a deterministic time trend.

2.  Drift (`type = "drift"`): To test for stationarity around a constant
    non-zero mean (most appropriate for interest rate spreads).

3.  None (`type = "none"`): To test for stationarity around zero.

#### 3.5.2 Decision Rule

For the `ur.df()` output, we focus on the **tau statistic (**$\tau$).
The null hypothesis ($H_0$) of a unit root is rejected only if the test
statistic is **more negative** (i.e., lies in the lower tail) than the
critical value at the 5% significance level.

If the series in levels is found to be non-stationary, we proceed with
**first-differencing** ($\Delta \text{Spread}_t$) and re-apply the test.
The transformation is considered successful only when the tau statistic
for the differenced series falls significantly below the critical
threshold, confirming that the process is $I(1)$ and ready for GARCH
modelling.

```{r}
risultato_adf <- adfTest(train, lags = 1, type = "c")
print(risultato_adf)
# Equivalent with auto determination of lags

test_adf_completo <- function(serie, nome_serie = "Serie") {
  
  cat("\n==============================================\n")
  cat("ANALISI STAZIONARIETÀ (ADF):", nome_serie, "\n")
  cat("==============================================\n")
  
  # Definiamo i tre tipi di test ADF
  # "none"  = senza costante né trend
  # "drift" = con costante (intercetta)
  # "trend" = con costante e trend temporale
  tipi <- c("none", "drift", "trend")
  
  risultati_lista <- lapply(tipi, function(t) {
    # Eseguiamo il test con selezione automatica dei lag tramite AIC
    test <- ur.df(serie, type = t, selectlags = "AIC")
    
    # Estraiamo la statistica del test e il valore critico al 5%
    statistica <- test@teststat[1]
    valore_critico <- test@cval[1, "5pct"]
    
    # Verifichiamo la stazionarietà (statistica < valore critico)
    esito <- if (statistica < valore_critico) "STAZIONARIA" else "NON STAZIONARIA"
    
    return(data.frame(
      Modello = t,
      Statistica_ADF = round(statistica, 3),
      Valore_Critico_5pct = round(valore_critico, 3),
      Risultato = esito,
      stringsAsFactors = FALSE
    ))
  })
  
  # Uniamo i risultati in una tabella leggibile
  tabella_finale <- bind_rows(risultati_lista)
  print(tabella_finale)
  
  return(tabella_finale)
}
# --- COME USARLA ---

# 1. Test sullo spread originale (livelli)
risultati_livelli <- test_adf_completo(train, "Spread BTP/Bund (Livelli)")
```

###3.6 Transformation and Diagnostic of Differenced Series Following the
identification of a unit root in the levels of the spread, we proceed
with **first-differencing** to achieve stationarity, transforming the
series into an $I(0)$ process. This transformation represents the weekly
change in the BTP-BUND spread (measured in basis points).

To detect the presence of **Conditional Heteroskedasticity**—a
fundamental requirement for GARCH modelling—we analyze the
Autocorrelation Function (ACF) and Partial Autocorrelation Function
(PACF) of both the differenced series and its squared values. While the
standard ACF/PACF help identify the $ARMA$ structure for the mean
equation, the squared series analysis is crucial to identify
**volatility clustering**, where large shocks tend to be followed by
other large shocks, regardless of their sign."

```{r}
# 1. Calcolo della differenza prima (Weekly Change)
dspread <- diff(spread)
dtrain <- dspread[train_start:train_end]
dtest  <- dspread[test_start:test_end]

# 2. Controllo rapido delle dimensioni (per sicurezza)
cat("Lunghezza Train:", length(dtrain), "(4 anni )\n")
cat("Lunghezza Test:", length(dtest), "(1 anno)\n")

# 3. Verifica se ci sono NA (fondamentale prima del GARCH)
if(any(is.na(dtrain))) {
  d_train <- na.omit(dtrain)
  cat("Attenzione: Rimossi valori NA dal train\n")
}
```

**3.6.1 Identification of the Mean and Variance Equations**

## 4.4. Autocorrelation Analysis of the Differenced Spread

The diagnostic plots for the differenced spread ($\Delta \text{Spread}_t$) provide several key insights into the identification of the mean equation:

* **ACF and PACF of the Differenced Series:** the **ACF and PACF of the daily differenced spread** show that the first-order lags (Lag 1) are now **statistically insignificant**, as they fall entirely within the 95% confidence intervals (represented by the blue dashed lines). 
* **Market Efficiency:** This lack of immediate autocorrelation suggests that daily changes in the BTP-BUND spread follow a process close to **White Noise**. Econometrically, this implies that the series behaves as a **Random Walk**, where the best predictor for tomorrow's spread is simply today's value, and daily innovations are largely unpredictable.
* **Higher-Order Lags:** While isolated spikes appear at **Lag 12** and **Lag 18**, they do not correspond to standard financial cycles (such as a 5-day business week or a 21-day monthly effect). 

Following the principle of **parsimony**, these higher-order spikes are deemed statistically borderline and likely represent **noise-driven artifacts**. We exclude them to avoid **overfitting** and to ensure model stability, as including $AR(18)$ or $MA(12)$ terms would lack a clear economic or theoretical justification for daily price changes.

### Conclusion for Model Identification
The diagnostic evidence confirms that the "memory" of the spread is captured almost entirely by the **integration ($d=1$)** of the series. The absence of a clear $AR$ or $MA$ pattern in the early lags indicates that the residuals are now ready for the next stage of analysis: testing for **Heteroskedasticity (ARCH effects)** in the squared residuals to model volatility clustering.

### Summary Table: ACF/PACF Diagnostics

| Lag | Significance | Interpretation |
| :--- | :--- | :--- |
| **Lag 1** | **Insignificant** | No immediate momentum or mean-reversion effect. |
| **Lags 2–10** | **Insignificant** | High market efficiency; no short-term memory. |
| **Lags 12 & 18** | **Significant** | Likely spurious noise; excluded to maintain parsimony. |
| **Overall Result** | **White Noise** | The differenced series is suitable for volatility modeling. |

## 4.5. Analysis of Volatility Clustering (Squared Residuals)

The diagnostic plots for the **squared differenced spread** ($\Delta \text{Spread}_t^2$) provide conclusive evidence regarding the variance structure of the series:

* **ACF and PACF of Squared Differenced Spread:** In stark contrast to the linear series, the squared residuals exhibit **highly significant spikes** across multiple lags (specifically Lags 1, 2, 4, 5, and 6). These spikes far exceed the 95% confidence intervals, indicating strong serial correlation in the second moments of the distribution.
* **Volatility Clustering:** This pattern is a classic indicator of **volatility clustering**, where periods of high variance are followed by high variance, and calm periods are followed by calm. Econometrically, this confirms that the error terms are not independently and identically distributed (i.i.d.) but follow a conditional heteroskedastic process.
* **ARCH Effects:** The presence of these significant lags suggests that a simple ARIMA model is insufficient to capture the risk dynamics of the spread. The "memory" in the squared returns necessitates the formal testing of **ARCH (Autoregressive Conditional Heteroskedasticity)** effects.

### Conclusion for Model Selection
The evidence of strong autocorrelation in the squared innovations justifies the transition from a linear time-series model to a **GARCH-family model**. This will allow for the simultaneous modeling of the conditional mean (identified as a Random Walk in the previous step) and the time-varying conditional variance, accurately capturing the "fat tails" and risk bursts observed in the BTP-BUND daily data.

### Summary Table: Squared Residuals Diagnostics

| Metric | Observation | Econometric Implication |
| :--- | :--- | :--- |
| **Squared ACF Lags** | Multiple significant spikes (1-6+) | Clear presence of ARCH effects. |
| **Squared PACF Lags** | Significant initial spikes | Suggests a GARCH(1,1) or GARCH(2,1) candidate. |
| **Volatility Structure** | Non-constant over time | Heteroskedasticity must be modeled. |
| **Final Strategy** | **GARCH Modeling** | Necessary to estimate time-varying risk. |
Finally, regarding the **intercept of the mean equation**, a $t$-test
was performed to determine if the mean of the differenced series is
significantly different from zero. If the $p$-value is below 0.05, it
implies the presence of a significant **drift** component, which will be
incorporated into the $ARMA$ specification via the `include.mean = TRUE`
parameter. If the mean is statistically zero, the spread exhibits a pure
random walk behavior without a deterministic trend.

```{r, fig.width=12, fig.height=10, out.width="100%"}
par(mfrow = c(4, 1), mar = c(3, 4, 3, 2))
Acf(as.numeric(dtrain), lag.max = 30, main = "ACF of Diff Spread")
Pacf(as.numeric(dtrain), lag.max = 30, main = "PACF of Diff Spread")
Acf(as.numeric(dtrain)^2, lag.max = 30, main = "ACF of Squared Diff Spread")
Pacf(as.numeric(dtrain)^2, lag.max = 30, main = "PACF of Squared Diff Spread")
par(mfrow = c(1, 1))
```

```{r}

t.test(dtrain)
```

The $t$-test results indicate that the mean of the differenced series is
not statistically different from zero ($p$-value $> 0.05$).

##3.7 Formal Stationarity Testing on First Differences To confirm the
achievement of stationarity, we re-apply the Augmented Dickey-Fuller
(ADF) test. Given that the previous $t$-test confirmed a zero mean for
the differenced series, we utilize the "none" specification (type =
"none").Econometrically, this is a robust choice: if the spread in
levels exhibited a trend, the first difference would show a constant
drift. If we were to assume a trend in the differences, it would imply
that the spread is not just increasing, but accelerating—a scenario that
lacks long-term economic equilibrium. In financial theory, interest rate
spreads are considered "mean-reverting" and anchored to macroeconomic
fundamentals. ###3.7.1 Theoretical Framework: Key Literature This
modeling approach is supported by three foundational works:Vasicek
(1977): Introduced the mathematical framework for mean-reversion. It
justifies the exclusion of a trend in the ADF test, as rates are pulled
back toward a long-term equilibrium.Cox, Ingersoll, & Ross (1985):
Established that volatility is conditional on the level of the interest
rate, providing the theoretical predecessor to GARCH modeling.Balduzzi,
Das, & Foresi (1998): Analyzed the "Central Tendency" of bond yields,
supporting the idea that spreads oscillate around a fundamental mean.

```{r}
test_stationarity_diff<-adfTest(dtrain,lags=1,type="nc")
print(test_stationarity_diff)
#equivalent but with auto determination of lags
test_autolag <- ur.df(dtrain, type = "none", selectlags = "AIC")
summary(test_autolag)
stat_test <- test_autolag@teststat[1]
valore_critico_5pct <- test_autolag@cval[1, "5pct"]

# 3. Verifica automatica con messaggio
cat("\n--- VERIFICA RISULTATO ADF (Type = None) ---\n")
cat("Statistica del Test:", round(stat_test, 4), "\n")
cat("Valore Critico (5%):", valore_critico_5pct, "\n")

if (stat_test < valore_critico_5pct) {
  cat("ESITO: La serie è STAZIONARIA (p-value < 0.05)\n")
  cat("Significato: Puoi procedere con il modello GARCH.\n")
} else {
  cat("ESITO: La serie NON è STAZIONARIA\n")
  cat("Attenzione: Potrebbe esserci ancora una radice unitaria.\n")
}
```

The Verdict:\
We reject null hypotesysi of non stationarity

##3.8 Seasonality and ARCH Diagnostics Before finalizing the model
specification, we perform a definitive check for residual seasonality
and ARCH effects.

###3.8.1 Power Spectrum and Seasonality Tests The Periodogram is
utilized to analyze frequency strength. A pronounced peak at a specific
frequency  would indicate
a periodic component. Furthermore, we apply the seastests package for a
combined formal verification.

```{r}
# Calcola lo spettro di potenza
?spec.pgram
spec_raw <- spec.pgram(as.numeric(dtrain), spans = c(3,3), taper = 0.1, plot = TRUE,log='no')
# Trova la frequenza con il picco massimo
max_freq <- spec_raw$freq[which.max(spec_raw$spec)]
periodo_dominante <- 1 / max_freq
cat(periodo_dominante)
```

### 4.6. Spectral Analysis Interpretation

The **Smoothed Periodogram** analysis provides a frequency-domain perspective that complements the previous time-series diagnostics:

* **Frequency Distribution:** The spectrum shows multiple significant peaks rather than a single dominant frequency. This confirms that the BTP-BUND spread is a **multi-cyclical series** where seasonality is fragmented across different time horizons.
* **Short-term Dynamics:** The prominent peaks at higher frequencies (e.g., near 0.4) represent the daily "noise" and short-term volatility. This aligns with the **Maximum value of 250.70 bps**, indicating that sharp, short-lived shocks are a structural feature of the spread's behavior.
* **Low-Frequency Persistence:** The power concentrated near the origin (frequency → 0) represents the **Trend component**. This captures the long-term shifts (2022 surge and 2025 reversion) which, despite the daily noise, remain the most powerful drivers of the spread's variance.

**Conclusion:** The spectral density confirms that while no simple seasonal rule can predict the spread, the series is far from being pure white noise. The complexity of the periodogram justifies the use of non-linear models (like GARCH) to capture the clusters of energy (volatility) that appear at various frequencies.

Frequenza (s),Periodo Temporale,Obiettivo dell'analisi
5,Settimanale,Verificare pattern intra-settimanali (Trading week).
21,Mensile,Catturare la stagionalità legata ai calendari macroeconomici.
63,Trimestrale,Analizzare l'impatto dei report trimestrali e scadenze tecniche.
252,Annuale,Confermare il pattern stagionale legato alle manovre di bilancio annuali.
```{r}

autoplot(dspread)

#To use this method, we need a ts object, let's create one
dspread_ts <- ts(dtrain, start=c(start_year,start_month,start_day), frequency = 21)#try also with 12
ggseasonplot(dspread_ts)

ggsubseriesplot(dspread_ts)

#Look also at decomposed series
decomposed <- stl(dspread_ts, s.window="periodic")
autoplot(decomposed)
```
--> STL Decomposition of the Differenced Series
### 1. STL Decomposition at Monthly Resolution
The decomposition of the BTP-BUND spread at a 21-day frequency reveals a highly active cyclical structure:

* **Trend:** The trend component remains the primary driver, capturing the broad expansion of the spread from 2021 through 2024 and the subsequent compression toward 2026. At this frequency, the trend is more reactive to mid-term economic shifts compared to the annual model ($f=252$).
* **Seasonal (Monthly):** The seasonal panel displays a persistent, high-frequency oscillation. However, the amplitude is notably small, ranging between **-1 and +1.5 bps**. This suggests that while a monthly cycle exists, its statistical contribution to the total variance is marginal.(Se il tuo spread medio è di 150 bps e la volatilità giornaliera (Remainder) ha picchi di 10-20 bps, un'oscillazione di soli 3 bps è considerata trascurabile (noise-level).)
* **Remainder:** The remainder component continues to exhibit significant "bursts" or clusters of volatility. This indicates that sudden market news (stochastic shocks) has a much larger impact on daily spread changes than any predictable monthly calendar effect.



We use seastests package

```{r}
#install.packages("seastests")

isSeasonal(dspread_ts, test="combined")
#We get FALSE

#Individual tests
isSeasonal(dspread_ts, test="qs")
isSeasonal(dspread_ts, test="seasdum")
isSeasonal(dspread_ts, test="kw")
isSeasonal(dspread_ts, test="fried")

#We get all FALSE

## Now plot ACF and PACF
Acf(dspread_ts, lag.max= 52)
Pacf(dspread_ts, lag.max = 52)
## Finally, we can affirm that our data does not contain seasonality, so we can work using normal ARIMA models and ARIMA-GARCH models.
```

We conclude the diagnostic phase by applying the seastests package. All
individual tests (QS, Seasonal Dummy, Kruskal-Wallis, and Friedman)
consistently fail to reject the null hypothesis of no seasonality.

Conclusion: Having confirmed that the series is stationary, lacks
seasonality, and exhibits significant ARCH effects, we can proceed with
confidence to estimate the non-seasonal ARIMA-GARCH framework.

###3.8.3 Formal Diagnostic Tests: Ljung-Box and ARCH-LM To ensure the
necessity of both the mean and variance components of the model, we
perform rigorous diagnostic tests on the differenced training set
($dtrain$). We select a high lag (e.g., $lag = 16$) to be meticulous in
detecting long-term dependencies.

```{r}
# 1. Ljung-Box Test on differences
lb_ret <- Box.test(as.numeric(dtrain), lag = 10, type = "Ljung-Box")
cat("\nLB Test on ds p-value:\n", lb_ret$p.value, "\n")
if(lb_ret$p.value < 0.05) cat("Autocorrelation detected, need mean modelling\n") else cat("ds are White Noise\n")

# 2. Ljung-Box Test on squared differences
lb_test <- Box.test(as.numeric(dtrain)^2, lag = 10, type = "Ljung-Box")
cat(sprintf("\nLjung-Box on Squared ds at lag 10:\n %.4f\n", lb_test$p.value))
if(lb_test$p.value < 0.05) cat("Volatility clustering confirmed\n") else cat("No volatility clustering detected\n")

# 3. ARCH-LM Test (Engle's Test for ARCH effects)
arch_test <- ArchTest(dtrain, lags = 10)
cat(sprintf("\nARCH Test (5 lags) LM Test - statistic: %.4f, p-value: %.4f\n", arch_test$statistic, arch_test$p.value))
if(arch_test$p.value < 0.05) cat("ARCH effects confirmed. GARCH modelling needed\n") else cat("No ARCH effects confirmed\n")
```

Interpretation of Results:Autocorrelation: If the $p$-value for the
first Ljung-Box test is below 0.05, we reject the null hypothesis of
independence, confirming that the spread changes exhibit autocorrelation
and require an ARMA specification.ARCH Effects: A $p$-value below 0.05
in both the squared Ljung-Box and the ARCH-LM test provides definitive
evidence of volatility clustering. This confirms that the variance is
not constant (Heteroskedasticity), mandating the use of a GARCH model to
capture time-varying risk.

##3.9 Seasonality Validation: Information Criteria Approach
While previous visual diagnostics suggested no seasonality, we utilize an
automated selection procedure to determine if a seasonal frequency
(SARIMA) provides a statistically superior fit. We compare models across
plausible frequencies ($f=1$ for non-seasonal, $f=8$, and $f=12$ for
monthly proxies in weekly data) based on the Akaike Information
Criterion (AIC).

DONT'T RUN THIS!!!
```{r}
#test_freqs <- c(1,5,21, 63,252) # Le frequenze plausibili
#risultati <- data.frame(Freq = test_freqs, AIC = NA)

#for(i in 1:length(test_freqs)) {
  # Crea la TS con la frequenza del loop
#  test_ts <- ts(as.numeric(dtrain), frequency = test_freqs[i])
  
  # auto.arima testa se la stagionalità a quella frequenza aiuta il modello
#  fit <- auto.arima(test_ts, max.q = 10, max.p = 10,seasonal = TRUE,method="ML",stepwise = FALSE,approximation = FALSE)
#  risultati$AIC[i] <- fit$aic
#}

# Trova la frequenza migliore
#best_f <- risultati$Freq[which.min(risultati$AIC)]
#cat("Il verdetto statistico è: usa la frequenza", best_f)
```




##3.10 Final Mean Equation Specification Following the automated search,
we estimate the final non-seasonal $ARIMA$ model to evaluate the
significance of individual parameters.

```{r}
m1 <- auto.arima(as.numeric(dtrain),max.q = 15, max.p = 15,method="ML",
                        seasonal = FALSE, 
                        stepwise = FALSE, 
                        approximation = FALSE,
                        trace = FALSE)

summary(m1)
```
Model for daily data (train=4 years): ARIMA(2,0,2)

3.10.1
Statistical Significance of CoefficientsTo ensure model parsimony, we
evaluate the significance of each parameter. A coefficient is
statistically significant at the 95% confidence level if its t-ratio
($t = \frac{\hat{\phi}}{SE(\hat{\phi})}$) exceeds the critical value of
$|1.96|$.

```{r}
check_coef_significance <- function(model) {
  # Estraiamo coefficienti e standard errors
  coefs <- model$coef
  se <- sqrt(diag(model$var.coef))
  
  # Calcoliamo la t-statistica (z-score)
  t_stats <- coefs / se
  
  # Calcoliamo i p-value (distribuzione normale a due code)
  p_values <- (1 - pnorm(abs(t_stats))) * 2
  
  # Creiamo una tabella riassuntiva
  results <- data.frame(
    Coefficiente = round(coefs, 4),
    Std_Error = round(se, 4),
    t_stat = round(t_stats, 4),
    p_value = round(p_values, 4),
    Significativo_95 = ifelse(p_values < 0.05, "SI", "NO")
  )
  
  cat("\nAnalisi di Significatività dei Coefficienti:\n")
  print(results)
  
  cat("\nNota: Un coefficiente è significativo al 95% se |t_stat| > 1.96\n")
  print(results)
}
check_coef_significance(m1)
m1$coef
```

ALL THE COEFFICIENTS OF m1 SARE TATISTICALLY SIGNIFICANT? t ratios are
given by estimated coefficients divided by their standard errors; the
latter are given by the square root of the element on the main diagonal
(variance of coef)

```{r}
names(m1)
summary(m1)
t.ratios = m1$coef/sqrt(diag(m1$var.coef))
t.ratios
```
BUT since the Lijung Box suggests that the serie is a white noie, we specify also the model m0=ARIMA(0,0,0) to compare the performances

library(forecast)
```{r}
# Fit ARIMA(1,0,0) - equivalent to AR(1)
m0 <- Arima(as.numeric(dtrain), order = c(0, 0, 0), include.constant = FALSE)

# Summary of coefficients and information criteria
summary(m0)
```

```{r}
arma_assumptions <- function(m) {
  cat("\n============================================================\n")
  cat("ARIMA model assumptions & Cycle Analysis:\n")
  
  # Estrazione corretta per modelli auto.arima/Arima
  # Usiamo i coefficienti AR per la stazionarietà
  ar_coefs <- m$model$phi
  ma_coefs <- m$model$theta
  
  # 1. Calcolo Radici
  ar_roots <- polyroot(c(1, -ar_coefs))
  ma_roots <- polyroot(c(1, ma_coefs))
  
  is_stationary <- all(abs(ar_roots) > 1)
  is_invertible <- all(abs(ma_roots) > 1)
  
  cat("Stationary:", is_stationary, "(Min mod:", round(min(abs(ar_roots)), 3), ")\n")
  cat("Invertible:", is_invertible, "(Min mod:", round(min(abs(ma_roots)), 3), ")\n")
  
  # 2. Analisi dei Business Cycles (Radici Complesse)
  # Cerchiamo se esistono radici con parte immaginaria significativa
  complex_roots <- ar_roots[Im(ar_roots) != 0]
  
  cat("\n--- Business Cycle Analysis ---\n")
  if(length(complex_roots) > 0) {
    cat("The model IMPLIES the existence of cycles.\n")
    cat("Reason: The AR characteristic equation has complex conjugate roots.\n")
    
    # Calcolo del periodo del ciclo (se ti serve per l'analisi)
    # Periodo = 2*pi / acos(Re(root)/abs(root))
    root <- complex_roots[1]
    period <- 2 * pi / atan2(Im(root), Re(root))
    cat("Estimated average cycle period:", round(abs(period), 2), "units of time.\n")
  } else {
    cat("The model DOES NOT imply cycles.\n")
    cat("Reason: All AR roots are real; the model returns to equilibrium monotonically.\n")
  }
  
  # 3. Check for common roots
  common_roots <- FALSE
  if(length(ar_roots) > 0 && length(ma_roots) > 0) {
    for(ar in ar_roots) {
      for(ma in ma_roots) {
        if(abs(ar - ma) < 0.05) common_roots <- TRUE
      }
    }
  }
  
  cat("\n--- Root Validation ---\n")
  if(common_roots) {
    cat("Common roots detected, assumption violation\n")
  } else {
    cat("No common roots detected, assumption validated\n")
  }
  cat("============================================================\n")
}

# Esecuzione
arma_assumptions(m1)
```

#### Residual's check

NOTE: Kolmogorov-Smirnov test (KS) to test wheter they follow a
t-distribution $H_0: \text{serie follows supposed distribution}$

```{r}

cat("\n============================================================\n")
cat("\nARMA residuals check:\n")
# ARIMA residuals
n_parameters1<-sum(m1$coef!=0) 
n_parameters2<-sum(m0$coef!=0)
arima_res1<- residuals(m1)
arima_res2<- residuals(m0)                    
arma_residual <- function(res,n_parameters) {
  cat("\n============================================================\n")
  cat("ANALISI APPROFONDITA DEI RESIDUI\n")
  cat("============================================================\n")

# 1. Ljung-Box (Media e Autocorrelazione)
  lb_arima <- Box.test(res, lag = 15, type = "Ljung-Box",fitdf = n_parameters)
  cat("\n [1]LB test on ARIMA residuals (15 lags) p-value: ", round(lb_arima$p.value, 4), "\n")
  if(lb_arima$p.value >= 0.05) cat("ARIMA model cleared the mean\n") else cat("ARIMA model DID NOT clear the mean\n")
#Regola di Hyndman (autore di forecast): Per dati non stagionali si usa $h = 10$, per dati stagionali si usa $h = 2m$ (dove $m$ è la frequenza). Nel tuo caso ($m=8$), dovresti testare almeno fino a 16 lag.
  
# 2. Durbin-Watson (Autocorrelazione al lag 1)
# Nota: Il DW classico si usa su modelli lineari, sui residui ARIMA è indicativo
  dw_test <- dwtest(as.vector(res) ~ 1)
  cat("\n[2] Durbin-Watson Test p-value: ", round(dw_test$p.value, 4))
  if(dw_test$p.value >= 0.05) cat(" -> OK: No autocorrelazione lag 1\n") else cat(" -> KO: Autocorrelazione lag 1\n")

# 3. Breusch-Godfrey Test (Autocorrelazione di ordine superiore)
  bg_test <- bgtest(as.vector(res) ~ 1, order = 12)
  cat("\n[3] Breusch-Godfrey Test (order 15) p-value: ", round(bg_test$p.value, 4))
  if(bg_test$p.value >= 0.05) cat(" -> OK: No autocorrelazione superiore\n") else cat(" -> KO: Autocorrelazione superiore\n")
  
# 4. Jarque-Bera Test (Normalità dei residui)
  jb_test <- jarque.bera.test(res)
  cat("\n[4] Jarque-Bera Test p-value: ", round(jb_test$p.value, 4))
  if(jb_test$p.value >= 0.05) cat(" -> OK: Residui Normali\n") else cat(" -> KO: Residui NON Normali (code grasse)\n")
  
#5. Kolmogorov-Smirnov (KS)
  check_t <- function(residuals) {
  fit_t <- suppressWarnings(fitdistr(residuals, densfun = "t"))
  mu <- fit_t$estimate["m"]
  sigma <- fit_t$estimate["s"]
  df <- fit_t$estimate["df"]
  std_res <- (residuals - mu) / sigma
  ks <- ks.test(x = std_res, y = "pt", df = df)
  cat("\n[5] KS test assuming t-distribution:\n")
  print(c(DoF = df, KS_p_value = ks$p.value))
  if(ks$p.value > 0.05) cat("\nResiduals are t-distributed\n") else cat("\nResiduals are NOT t-distributed\n")
  }
  check_t(res)
  # 6. Effetti ARCH (Eteroschedasticità)
  lb_arima_2 <- Box.test(res^2, lag =15, type = "Ljung-Box",fitdf = n_parameters)
  cat("\n [6]LB test on ARIMA squared residuals (10 lags) p-value: ", round(lb_arima_2$p.value, 4), "\n")
  if(lb_arima_2$p.value < 0.05) cat("ARIMA model has residual variance to explain. GARCH model needed\n") else cat("ARIMA model residual variance is OK\n")

  # 7.ARCH test
  arch_test_res <- ArchTest(res, lags = 15)
  cat(sprintf("\n [7]ARCH Test on ARIMA (5 lags) residuals:\n LM Test - statistic: %.4f, p-value: %.4f\n", 
              arch_test_res$statistic, arch_test_res$p.value))
  if(arch_test_res$p.value < 0.05) cat("ARCH effects confirmed. GARCH modelling needed\n") else cat("No ARCH effects confirmed\n")
}

print(arma_residual(arima_res1,n_parameters1))
print(arma_residual(arima_res2,n_parameters2))
```
m1=ARIMA(2,0,2) pass all the tests, while m0=ARIMA(0,0,0) shows autocorrelation up to lag 15 (but if we set breush-Godfrey up to lag 11, there is not autocorrelation, so the problem is the lag 12)

ORA FACCIAMO STA BENEDETTA PREVISIONE:
IN SAMPLE PREDICTIONS

```{r}
# Estraiamo i valori fittati dal modello m1
fitted_values <- fitted(m1)
length(fitted_values)
length(train)
# Visualizzazione grafica: Reale vs Fittato
plot(as.numeric(dtrain), type="l", col="black", lwd=1, 
     main="In-Sample Evaluation: Reale vs ARIMA",
     ylab="Spread", xlab="Tempo (Osservazioni)")
lines(as.numeric(na.omit(fitted_values)), col="red")
legend("topright", legend=c("Dato Reale", "Fitted"), col=c("black", "red"), lty=1, lwd=2)
```

IN SAMPLE EVALUATION

```{r}
residuals <- dtrain - fitted_values

# Core metrics
accuracy(as.numeric(dtrain), as.numeric(fitted_values))
# R-squared (will be 0 or near 0 — expected for mean model)
SS_res <- sum(residuals^2, na.rm = TRUE)
SS_tot <- sum((dtrain - mean(train, na.rm = TRUE))^2, na.rm = TRUE)
R2 <- 1 - SS_res / SS_tot
cat("R2 del modello:", round(R2, 2), "%") 

# Calcoliamo i segni (1 per positivo, -1 per negativo)
sign_actual <- sign(dtrain)
sign_fitted <- sign(fitted_values)

# Creiamo un vettore logico: TRUE se i segni sono uguali
hits <- (sign_actual == sign_fitted)

# Calcoliamo la percentuale di successo (escludendo eventuali NA)
DA <- mean(hits, na.rm = TRUE) * 100
DA
```

OUT-SAMPLE PREDICTION AND EVALUATION

```{r}
# 1. Applichiamo il modello alla serie completa (genera tutti i forecast 1-step)
refit <- Arima(dspread, model = m1)

# 2. Estraiamo solo i valori fittati che corrispondono al periodo di TEST
# (Supponiamo che il tuo test set sia lungo 33 osservazioni)
pred_out <- tail(fitted(refit), n_test)
reali_out <- tail(as.numeric(dspread), n_test)

# 3. Calcolo metriche Out-of-Sample
accuracy(pred_out, reali_out)

# R-squared (will be 0 or near 0 — expected for mean model)
SS_res <- sum((pred_out-reali_out)^2, na.rm = TRUE)
SS_tot <- sum((test - mean(test, na.rm = TRUE))^2, na.rm = TRUE)
R2 <- 1 - SS_res / SS_tot
cat("R2 del modello:", round(R2, 2), "%") 

# Calcoliamo i segni (1 per positivo, -1 per negativo)
sign_actual <- sign(test)
sign_fitted <- sign(pred_out)
hits <- (sign_actual == sign_fitted)
# Calcoliamo la percentuale di successo (escludendo eventuali NA)
DA <- mean(hits, na.rm = TRUE) * 100
cat("DA del modello:", round(DA, 2), "%") 

```
IF WE SUBSTITUTE m1 with m0, we can see that m1 perform better both insample and out-sample

FORECAST DI domani PROSSIMA (diff_spreads)

```{r}
# 2. Generiamo la previsione per la prossima settimana
forecast_next_week <- forecast(refit, h = 1)

# 3. Visualizziamo il risultato
print(forecast_next_week)

# 4. Grafico della previsione con intervalli di confidenza
plot(forecast_next_week, main = "Previsione Spread BTP-Bund: Settimana Prossima", 
     ylab = "Spread", xlab = "Tempo")
```

Lo 80 / Hi 80 (da -21.79 a +22.18): Il modello è sicuro all'80% che la
variazione dello spread oscillerà tra una discesa di 21.8 punti e una
salita di 22.2 punti. È un range abbastanza ampio che riflette la
volatilità tipica dei mercati.

Lo 95 / Hi 95 (da -33.43 a +33.82): Il modello è sicuro al 95% (standard
statistico per la tesi) che la variazione non supererà i ±33 punti base.

Nota tecnica: Se la prossima settimana lo spread dovesse muoversi di 50
punti, sapresti che è successo un evento "fuori dal comune" (un outlier)
che il modello non poteva prevedere. TRASFORMAZIONE IN spread:

```{r}
# 1. Recuperiamo l'ultimo valore REALE dello spread (non differenziato)
ultimo_spread_reale <- tail(spread, 1) 

# 2. Otteniamo la previsione della differenza (il "delta")
previsione_diff <- forecast(m1, h = 1)$mean[1]

# 3. Calcoliamo lo spread atteso
spread_previsto <- ultimo_spread_reale + previsione_diff

cat("Lo spread previsto per domani è:", spread_previsto)
```

CONSIDERARE DI FARE UNA CROSS VALIDATION PER UNA ROBA PIU ACCURATA

############################################################################################################ 

## Volatility Analysis

We restart from our model

```{r}
summary(m1)
```

Remark that the mean is = 0 so we can work directly on residuals

```{r}
fitted_values <- fitted(m1)
res<-dtrain - fitted_values
par(mfrow = c(3, 1), mar = c(3, 4, 3, 2))

plot(index(as.xts(dspread)), coredata(as.xts(dspread)), type = "l", lwd = 0.7,
     main = "Residuals^2 ", xlab = "", ylab = "ds")
abline(h = 0, col = "black", lwd = 0.5)
Acf(as.numeric(res)^2,   lag.max = 30, main = "ACF of res^2")
Pacf(as.numeric(res)^2,  lag.max = 30, main = "PACF of res^2")
par(mfrow = c(1, 1))
```

Check for autocorrelation and ARCH effects on `train`:

```{r}
n_parameters<-sum(m1$coef!=0) 
# Ljung-Box on DS^2
lb_ds2 <- Box.test(as.numeric(res)^2, lag = 10, type = "Ljung-Box",fitdf = n_parameters)
cat(sprintf("LB Test on DS^2 p-value: %.4f\n", lb_ds2$p.value))
if (lb_ds2$p.value < 0.05) cat("Volatility clustering confirmed\n") else cat("No volatility clustering detected\n")

# ARCH test
arch_test <- ArchTest(as.numeric(res), lags = 10)
cat(sprintf("ARCH Test (10 lags) p-value: %.4f\n", arch_test$p.value))
if (arch_test$p.value < 0.05) cat("ARCH effects confirmed — GARCH modelling needed\n") else cat("No ARCH effects\n")

```

We clearly need a specification model for volatility, but which one?
Let's see if we have a unit root (i.e if the shock in volatility is persistent)
```{r}
adf.test(as.numeric(res)^2)
```
We reject the hyptosisi of unit root --> so we don't need to use IGARCH (correggetemi se sbaglio)

## Model 3: ARMA-GARCH

We try to find the best ARMA-GARCH fitting different combinations of
models, then decide considering a precise IC, kinda like an auto_arima
but hand-builded.

Create a function that simulates 1 iteration of a for cycle, then
instead of a for cycle use parrallel (mclapply) It will run the function
over $n$ times using $n$ CPU simultanealy.

Can change the critera or modify the parameters, if too much
combinations the code will run for ages.

We will keep the max possible GARCH orders to 1, that's cause vast
majority of the time a GARCH(1,1) is perfect for financial series. We
could add more GARCH model other than `sGARCH`and `eGARCH` but since
those are the most used we keep only them, also other models have
additional parameter than can increase the computational time. We also
capped max ARMA order to 3 to prevent the AIC from picking up noise (it
does, tried to run multiple times and it gives back models that do not
pass diagnostic tests).

The ugarch solvers (`gosolnp`) use random generate starting point, to
have reproducibility of results we need to impose a seed whenever a
optimization is done.

**NOTE**: in `dists` you choose the distribution you suppose the std
residuals will follow, `norm`= normal, `std`= t, `sstd`= skewed t. I
removed `norm`and cause the function would wrongly choose `norm` over
the other 2: you can see if is the distribution fit the residuals in the
printed residual plot or in the "Adjusted Pearson Goodness-of-Fit Test"
at the end of the model output, when `norm` the fit is worse. The
distribution that better fit the residuals is `sstd` (even if not
perfect).

The reason is that `sstd` and `std` have more parameters than `norm`
(number of parameter: `sstd` \> `std` \> `norm`) and since the train
data is "small" this penalize them more than it should in the eye of
AIC/BIC. `norm` will end up with better IC than `sstd` but then you have
the wrong distribution for the residuals and/or fail the diagnostic
tests.

**NOTE 2**: `mc.cores = detectCores() - 1` is a MAC thing, on Windows it
does not work, it will show that the code is running on x core but
actually will use just one, like a normal `for` cycle.

eval = FALSE --> Questo codice NON verrà eseguito con "Run All"
```{r,eval = FALSE}
start_time <- proc.time() # to see for how much the code run

best_garch <- function(y,
                       ar_lags  = 0:5,
                       ma_lags  = 0:5,
                       garch_p  = 1:1,
                       garch_q  = 1:1,
                       models   = c("sGARCH", "sGARCH-M"),
                       dists    = c("sstd", "std"),
                       criteria = "aic", 
                       timeout  = 60, 
                       seed = 42) {

  # reproducibility for the hybrid solver, it uses random starting point
  RNGkind("L'Ecuyer-CMRG")
  set.seed(seed)

  # Create every combinations possible for the parameter
  combos <- expand.grid(
    ar_lag  = ar_lags,
    ma_lag  = ma_lags,
    garch_p = garch_p,
    garch_q = garch_q,
    model   = models,
    dist    = dists,
    stringsAsFactors = FALSE
  )

  cat("Total combinations to fit:", nrow(combos), "\n")
  cat("Running on", parallel::detectCores() - 1, "cores\n")
  cat("It could take couple of minutes ...\n\n")

  fit_one <- function(i) {
    arp   <- combos$ar_lag[i]
    map   <- combos$ma_lag[i]
    gp    <- combos$garch_p[i]
    gq    <- combos$garch_q[i]
    model <- combos$model[i]
    dist  <- combos$dist[i]

    spec <- ugarchspec(
      variance.model     = list(model = model, garchOrder = c(gp, gq)),
      mean.model         = list(armaOrder = c(arp, map), include.mean = TRUE),
      distribution.model = dist
    )
    # If R encounter an error it stop the cycle, tryCatch catch the error and return NULL, then continue
    fit <- tryCatch({
      withTimeout({
        suppressWarnings(
          ugarchfit(spec = spec, data = y, solver = "gosolnp", 
          solver.control = list(n.restarts = 5, rseed = 42, trace = 0)) # for reproducibility)
        )
      },
      timeout   = timeout,
      onTimeout = "silent")
    },
    error   = function(e) NULL)

    # Store only not null + right class
    if (!is.null(fit) && inherits(fit, "uGARCHfit") && fit@fit$convergence == 0) {
      ic <- infocriteria(fit)
      return(data.frame(
        ar_p    = arp,
        ma_q    = map,
        garch_p = gp,
        garch_q = gq,
        model   = model,
        dist    = dist,
        aic     = as.numeric(round(ic["Akaike", ] * length(y), 2)), # the function calculate the aic only for that obs
        bic     = as.numeric(round(ic["Bayes",  ] * length(y), 2)),
        likelihood = as.numeric(round(likelihood(fit), 2)),
        stringsAsFactors = FALSE
      ))
    }
    return(NULL)
  }

# Run more itrations simultanely with every CPU-1, instead of for cycle
# results_list = mclapply(
#    seq_len(nrow(combos)),
#    fit_one,
#    mc.cores = detectCores() - 1
#  )
  
  
# questa è versione macOS:
 results_list <- mclapply(
  seq_len(nrow(combos)),
  fit_one,
  mc.cores = N_CORES  # usa la variabile definita sopra
)
  
  #versione per Windows:
  #  Crea il "ponte" tra i core del tuo PC (Cluster)
  #cl <- makeCluster(detectCores() - 1)
  # Esporta tutto ciò che serve ai core "figli" (fondamentale su Windows!)
  # Esportiamo i dati (y), la tabella (combos), la funzione (fit_one) e il timeout
  #clusterExport(cl, varlist = c("combos", "y", "timeout", "fit_one"), envir = environment())
  # Carica le librerie su ogni singolo core
  #clusterEvalQ(cl, {
  #  library(rugarch)
  #  library(R.utils)
  #  library(dplyr)
  #})  
  # Esegui il calcolo (parLapply sostituisce mclapply)
  #results_list <- parLapply(cl, seq_len(nrow(combos)), fit_one)
  # IMPORTANTE: Spegni i motori una volta finito
 gc(verbose = FALSE)

# Keep only non NULL model
  successful <- Filter(Negate(is.null), results_list)
  cat("Successful fits:", length(successful), "/", nrow(combos), "\n")
  cat("Models ordered by: ", criteria, "\n")

  results_df <- bind_rows(successful) |> arrange(.data[[criteria]])
  return(results_df)
}

results = best_garch(train)
cat("\nFitted models:\n")
print(head(results))

end_time <- proc.time()
execution_time <- end_time - start_time
cat("\n Total execution time: ", execution_time[3], "seconds")
```


```{r}
library(parallel)
library(R.utils)
library(rugarch)
library(dplyr)
start_time <- proc.time() # to see for how much the code run

best_garch <- function(y,
                       ar_lags = 0:5, ma_lags = 0:5,
                       garch_p = 1:1,  garch_q = 1:1,
                       # Usiamo "-M" come etichetta personalizzata
                       models  = c("sGARCH", "sGARCH-M"),
                       dists   = c("norm", "std", "sstd"),
                       criteria = "bic", timeout = 60) {

  combos <- expand.grid(
    ar_lag = ar_lags, ma_lag = ma_lags,
    garch_p = garch_p, garch_q = garch_q,
    model_label = models, # Questa è la nostra etichetta
    dist = dists,
    stringsAsFactors = FALSE
  )

  cat("Total combinations to fit:", nrow(combos), "\n")
  num_cores <- detectCores() - 1

  fit_one <- function(i) {
    # Carichiamo i parametri della riga i
    row  <- combos[i, ]
    arp  <- row$ar_lag; map <- row$ma_lag
    gp   <- row$garch_p; gq <- row$garch_q
    dist <- row$dist
    
    # --- TRADUZIONE LOGICA ---
    # 1. Controlliamo se nell'etichetta c'è "-M"
    is_archm <- grepl("-M", row$model_label)
    
    # 2. Puliamo il nome per passarlo a variance.model (es: "sGARCH-M" -> "sGARCH")
    clean_v_model <- gsub("-M", "", row$model_label) 

    # 3. Configuriamo lo spec con la distinzione tra Media e Varianza
    spec <- ugarchspec(
      variance.model = list(model = clean_v_model, garchOrder = c(gp, gq)),
      mean.model     = list(armaOrder = c(arp, map), 
                            include.mean = TRUE, 
                            archm = is_archm), # <--- Attiva il GARCH-M qui
      distribution.model = dist
    )
    fit <- tryCatch({
      withTimeout({
        suppressWarnings(
          ugarchfit(spec = spec, data = y, solver = "gosolnp", 
          solver.control = list(n.restarts = 5, rseed = 42, trace = 0)) # for reproducibility)
        )
      },
      timeout   = timeout,
      onTimeout = "silent")
    },
    error   = function(e) NULL)
    
    if (!is.null(fit) && inherits(fit, "uGARCHfit")) {
      ic <- infocriteria(fit)
      return(data.frame(
        ar_p = arp, ma_q = map, garch_p = gp, garch_q = gq,
        model = row$model_label, # Restituisce l'etichetta originale (es. sGARCH-M)
        dist = dist,
        aic = as.numeric(round(ic["Akaike", ] * length(y), 2)),
        bic = as.numeric(round(ic["Bayes",  ] * length(y), 2)),
        likelihood = as.numeric(round(likelihood(fit), 2)),
        stringsAsFactors = FALSE
      ))
    }
    return(NULL)
  }

  # Gestione Cluster (Windows richiede export esplicito)
  os_name <- Sys.info()["sysname"]
  if (os_name == "Windows") {
    cl <- makeCluster(num_cores)
    clusterExport(cl, varlist = c("y", "combos", "fit_one"), envir = environment())
    clusterEvalQ(cl, { library(rugarch); library(R.utils) })
    results_list <- parLapply(cl, seq_len(nrow(combos)), fit_one)
    stopCluster(cl)
  } else {
    results_list <- mclapply(seq_len(nrow(combos)), fit_one, mc.cores = num_cores)
  }

  successful <- Filter(Negate(is.null), results_list)
  if (length(successful) == 0) return(NULL)
  
  results_df <- bind_rows(successful) |> arrange(.data[[criteria]])
  return(results_df)
}

results <- best_garch(as.numeric(train))
cat("\nTop models by BIC:\n")
print(head(results))

end_time <- proc.time()
execution_time <- end_time - start_time
cat("\n Total execution time: ", execution_time[3], "seconds")
```
Extract the best model.

```{r, fig.width=12, fig.height=10, out.width="100%",}
arma_best_order <- c(results$ar_p[1], results$ma_q[1])
garch_best_order  <- c(results$garch_p[1], results$garch_q[1])
best_dist  <- results$dist[1]
best_mod  <- results$model[1]

cat("\nBest ARMA-GARCH model orders:\nARMA:  ", arma_best_order, "\nGARCH: ", garch_best_order,"\n") 
  garch_spec <- ugarchspec(variance.model = list(model = best_mod, garchOrder = garch_best_order),
                         mean.model = list(armaOrder =arma_best_order, include.mean = TRUE),
                         distribution.model = best_dist)

set.seed(42)
garch_fit <- ugarchfit(spec = garch_spec, data = dtrain, solver = "gosolnp", 
solver.control = list(n.restarts = 5, rseed = 42, trace = 0)) # for reproducibility


```
best model with dtrain(4 anni)(max garch par=3) = ARMA-GARCH(4,1)-(3,3),distribution=std (RANK=AIC)  agg.03_04 
best model with dtrain(4 anni)(max garch par=1) = ARMA-GARCH(4,2)-(1,1),distribution=sstd (RANK=AIC) agg.03_04  
best model with dtrain(4 anni)(max garch par=1) = ARMA-GARCH(3,2)-(1,1),distribution=norm (RANK=BIC) agg.03_04
best model with dtrain(4 anni)(max garch par=1) = ARMA-GARCH(1,0)-(1,1),distribution=norm (RANK=BIC) agg.10_04
best model with dtrain(4 anni)(max garch par=1) = ARMA-GARCH-M(1,0)-(1,1),distribution=norm (RANK=AIC) agg.10_04
Sulla base del criterio informativo **BIC**, il modello che meglio approssima la dinamica dello spread BTP-Bund è un **ARMA(3,2)-sGARCH(1,1)** con distribuzione dei residui Normale.

## Function to check the garch positivity of coefficients
```{r,}

# ============================================================
# GARCH POSITIVITY & STATIONARITY CHECK (rugarch)
# ============================================================
check_garch_positivity <- function(fit, model_name = "GARCH fit") {

  cf    <- fit@fit$coef
  # omega is always named "omega" in rugarch
  omega <- cf["omega"]
  # alpha coefficients: alpha1, alpha2, ...
  alphas <- cf[grepl("^alpha[0-9]+", names(cf))]
  # beta coefficients: beta1, beta2, ...
  betas  <- cf[grepl("^beta[0-9]+",  names(cf))]

  persistence <- sum(alphas) + sum(betas)

  positivity_ok   <- (omega > 0) & all(alphas >= 0) & all(betas >= 0)
  stationarity_ok <- persistence < 1

  cat("\n============================================================\n")
  cat("GARCH POSITIVITY & STATIONARITY CHECK:", model_name, "\n")
  cat("============================================================\n")
  cat(sprintf("  omega  = %+.6f  > 0?  %s\n",
              omega,  ifelse(omega > 0,      "YES ✓", "NO  ✗")))

  for (nm in names(alphas)) {
    cat(sprintf("  %-7s= %+.6f  >= 0? %s\n",
                nm, alphas[nm], ifelse(alphas[nm] >= 0, "YES ✓", "NO  ✗")))
  }
  for (nm in names(betas)) {
    cat(sprintf("  %-7s= %+.6f  >= 0? %s\n",
                nm, betas[nm],  ifelse(betas[nm]  >= 0, "YES ✓", "NO  ✗")))
  }

  cat(sprintf("\n  Persistence (sum α + sum β) = %.6f  < 1? %s\n",
              persistence, ifelse(stationarity_ok, "YES ✓ (covariance stationary)", "NO  ✗ (IGARCH or explosive)")))

  if (positivity_ok & stationarity_ok) {
    uncond_var <- as.numeric(omega) / (1 - persistence)
    cat(sprintf("  Unconditional variance       = %.6f\n", uncond_var))
    cat(sprintf("  Unconditional vol (daily)    = %.4f bps\n", sqrt(uncond_var)))
    cat("  --> All positivity conditions SATISFIED ✓\n")
  } else {
    if (!positivity_ok)   cat("  --> WARNING: Positivity condition VIOLATED ✗\n")
    if (!stationarity_ok) cat("  --> WARNING: Stationarity condition VIOLATED (persistence >= 1) ✗\n")
  }
  cat("============================================================\n\n")

  invisible(list(
    omega         = omega,
    alphas        = alphas,
    betas         = betas,
    persistence   = persistence,
    positivity_ok = positivity_ok,
    stationarity_ok = stationarity_ok
  ))
}

```

## Check garch positivity of coefficients

```{r,}
# --- Positivity check on the training GARCH fit ---
check_garch_positivity(garch_fit, model_name = "ARMA-GARCH (training fit)")

# check the ACF of the standardized residuals
Acf(garch_fit@fit$z)

# check the ACF of the squared residuals
Acf(garch_fit@fit$z^2,lag.max=30,main = "ACF of Squared Std. Residuals")
#Ljung-Box test
Box.test(garch_fit@fit$z^2, lag = 20, type = "Ljung-Box", fitdf = 2)
```
Dobbiamo preoccuparci di cui lag significativi al 19 e 23?
## Analisi del Modello Ottimale: ARMA(3,2)-sGARCH(1,1)

### 1. Parametri della Varianza (GARCH)

I coefficienti stimati descrivono la struttura della volatilità e la sua evoluzione temporale:

* **$\omega$ (Omega) = 0.512**: Rappresenta la costante della varianza. Essendo positiva, assicura un livello minimo di volatilità ("rumore di fondo") sempre presente nel mercato.
* **$\alpha_1$ (Alpha) = 0.201**: Misura la reattività immediata. Indica che il **20%** di uno shock inatteso (news, eventi politici) si traduce istantaneamente in nuova volatilità per il giorno successivo.
* **$\beta_1$ (Beta) = 0.777**: Indica la persistenza. Suggerisce che il **77%** della volatilità odierna viene trascinata per inerzia al giorno seguente, riflettendo una memoria lunga degli shock finanziari.

### 2. Diagnostica di Stabilità

L'affidabilità del modello è confermata dal superamento dei test di stabilità matematica:

* **Persistenza ($\alpha + \beta$) = 0.978**: Il valore è inferiore a 1, garantendo che il modello sia **stazionario**. Lo spread tende quindi a riassorbire gli shock e a convergere verso la sua media nel lungo periodo.
* **Volatility Clustering**: La prossimità del valore a 1 conferma la tendenza della volatilità a manifestarsi a "blocchi" (periodi di alta tensione seguiti da periodi di calma).

### 3. Indicatori di Lungo Periodo

* **Varianza Incondizionata**: 23.73
* **Volatilità Incondizionata (Daily)**: **4.87 bps**

> **Interpretazione**: In condizioni di equilibrio, lo spread oscilla mediamente di circa 4.9 punti base al giorno. Variazioni significativamente superiori indicano una fase di stress che eccede la norma statistica del periodo analizzato.

## Diagnostic

```{r,}
show(garch_fit)
infocriteria(garch_fit)*length(train)
```

In the ARMA(3,2)-GARCH(1,1) output look at:

-   "Weighted Ljung-Box Test on Standardized Residuals"
-   "Weighted Ljung-Box Test on Standardized Squared Residuals"
-   "Weighted ARCH LM Tests"
-   "Adjusted Pearson Goodness-of-Fit Test" --\> HO: la distribuzione
    scelta fitta bene i dati
    
```{r,}
plot(garch_fit,which=9)#does  seem too good
plot(garch_fit,which=8)

#JB test
residui_std <- garch_fit@fit$z
jb_test <- jarque.bera.test(residui_std)
print(jb_test)
```
obv normal distribution would not be a good idea but check if Shewed student t distribution would be better
    
Extract the best model with std o or sstd distribution

```{r, fig.width=12, fig.height=10, out.width="100%",}
library(rugarch)
library(tseries)

# Funzione per estrarre metriche
get_metrics <- function(fit, label) {
  ni <- infocriteria(fit)
  data.frame(
    Model = label,
    AIC = ni[1]*length(dtrain),
    BIC = ni[2]*length(dtrain),
    Likelihood = as.numeric(likelihood(fit)),
    Pearson_p_min = min(gof(fit, c(20, 30, 40, 50))[,3]), # p-value minimo tra i vari group
    stringsAsFactors = FALSE
  )
}

# --- 1. Modello con distribuzione Student-t (std) ---
garch_spec_std <- ugarchspec(variance.model = list(model = best_mod, garchOrder = garch_best_order),
                         mean.model = list(armaOrder =arma_best_order, include.mean = TRUE),
                         distribution.model = "std")

set.seed(42)
garch_fit_std <- ugarchfit(spec = garch_spec_std, data = dtrain, solver = "gosolnp", 
solver.control = list(n.restarts = 5, rseed = 42, trace = 0)) # for reproducibility

# --- 2. Modello con distribuzione Skew-Student-t (sstd) ---
garch_spec_sstd <- ugarchspec(variance.model = list(model = best_mod, garchOrder = garch_best_order),
                         mean.model = list(armaOrder =arma_best_order, include.mean = TRUE),
                         distribution.model = "sstd")

set.seed(42)
garch_fit_sstd <- ugarchfit(spec = garch_spec_sstd, data = dtrain, solver = "gosolnp", 
solver.control = list(n.restarts = 5, rseed = 42, trace = 0)) # for reproducibility
# --- Confronto Metriche ---
comp_table <- rbind(get_metrics(garch_fit_std, "Student-t (std)"),
                    get_metrics(garch_fit_sstd, "Skew-Student-t (sstd)"))

print(comp_table)

# --- Visualizzazione QQ-Plot ---
par(mfrow = c(1, 2))
plot(garch_fit_std, which = 9, main = "QQ-Plot: Student-t")
plot(garch_fit_sstd, which = 9, main = "QQ-Plot: Skew-Student-t")

```
Student t distribution wins so we don't have do change the model

```{r}
show(garch_fit)
```

**Analisi del Sign Bias Test (Engle e Ng)**

Il **Sign Bias Test** è stato condotto per verificare se il modello GARCH scelto sia in grado di catturare adeguatamente le potenziali asimmetrie nella volatilità (ovvero se shock positivi e negativi di pari entità abbiano un impatto differente sulla varianza futura).

Interpretazione dei risultati
* **Accettazione di $H_0$**: Tutti i p-value sono ampiamente superiori alla soglia critica di **0.05**. Non è possibile rifiutare l'ipotesi nulla di assenza di effetti di asimmetria nei residui.
* **Sufficienza del Modello**: Il test conferma che il modello **sGARCH(1,1)**, pur essendo una specifica simmetrica, è statisticamente **sufficiente** per descrivere la serie storica dello spread BTP-Bund analizzata.
* **Conclusione**: Non vi è prova che i rialzi dello spread generino una volatilità significativamente diversa rispetto ai ribassi (al netto di quanto già catturato dal modello). Pertanto, non è necessario ricorrere a specifiche asimmetriche più complesse (come eGARCH o GJR-GARCH), garantendo la parsimonia della stima.

N.B: È sorprendente per lo spread BTP-Bund: Di solito, lo spread reagisce in modo asimmetrico (la paura di un aumento dello spread genera più panico di quanto un calo generi euforia).
Se il test ti dà p-value così alti (0.74), significa che nel periodo che hai analizzato, la volatilità è stata guidata dall'entità degli shock (quanto grandi erano) piuttosto che dal loro segno (se erano positivi o negativi).--> CHECK LITERATURE


The model pass all the tests.

We also test if there's any breakpoint inside the residuals. We use:

-   Bau-Perron test: tests for multiple unknow structural breaks in the
    mean, it find the best breaks location that minimizes the RSS, then
    selects the number of breaks minimizing BIC. The number of breaks
    with lowest BIC is the correct one.
-   CUSUM: tests for parameter instability by tracking the cumulative
    sum of recursive residuals, a deviation of this sum from 0 means
    instability, $H_0:$ stable parameters throught the sample.


```{r, fig.width=12, fig.height=10, out.width="100%",}
par(mfrow = c(2, 1))
resid_std <- residuals(garch_fit, standardize = TRUE) |> as.numeric()

bp_resid_df <- data.frame(
  resid = resid_std,
  date  = index(dtrain) 
)
bp_resid <- breakpoints(resid ~ 1, data = bp_resid_df, h = 0.15)
summary(bp_resid)
plot(bp_resid)

cusum_resid <- efp(resid ~ 1, data = bp_resid_df, type = "Rec-CUSUM")
plot(cusum_resid)
sctest(cusum_resid)

par(mfrow = c(1, 1))
```

Checking the RSS table we can see that we have the optimal BIC at 0
breaks -\> the residuals do not need breakpoint, adding any break will
make the model worse. Also the RSS remain almost equal with differents
possibles breaks: adding breaks will not explain anything more.

We fail to reject the null in the CUSUM test since the line in plot stay
inside the confidence intervals -\> parameters are stable.

Combined both tests tell us that there's no evidence of structural
breaks in the standardize residuals, any possible change presents in the
raw series is explained by the model.

## Comment

The estimated model pass all the tests: there's no mean or volatility to
model and the residual fit the supposed distribution (t-skewed) and pass
the tests. Our data can be modelled with this ARMA-GARCH model, next we
run a validation process for its forecasts and collect some metrics. \#
Validation Now the validation phase, we take the best estimated
ARMA-GARCH and validate through the test data. At every iteration we
refit the model (re-estimate the parameter) but the structure of the
model remain the same: the goal is to calculate how good the model is at
forecasting to then forecast the real future value.

Then we compare the forecast with a Random Walk model to test it our
model is actually useful.

-   Forecast on test data using rolling forecast (window increases on 1
    week every iteration)
-   Refit the model every 1 iteration (1 week) to keep model useful
-   Perform the rolling estimation and forecast
-   refit.every = 1 means we re-estimate the model every single
    iteration
-   If window.size too small the model will fail to converge


Il grafico "Conditional SD vs Observations" sovrappone la volatilità stimata dal modello ($\sigma_t$) ai dati reali (i residui o le variazioni dello spread).   
Ecco come interpretarlo:  
1. Cosa rappresentano le linee?
--> La linea nera (o i punti):  Rappresenta i dati reali osservati (nel tuo caso, le variazioni giornaliere dello spread $\Delta \text{Spread}_t$). Mostra quanto il mercato si è mosso effettivamente in quel giorno.  
--> La linea colorata (spesso rossa o blu): Rappresenta la Conditional Standard Deviation ($\sigma_t$), ovvero la previsione del rischio fatta dal modello GARCH. Di solito viene visualizzata come un "inviluppo" (sopra e sotto i dati, o solo sopra come valore assoluto).
2. Cosa deve succedere se il modello è buono?  
Inseguimento (Tracking): La linea della volatilità deve "seguire" l'ampiezza dei dati reali. Quando vedi dei picchi nei dati neri (shocks), la linea della volatilità deve alzarsi immediatamente.Contenimento: La linea della volatilità dovrebbe agire come un "guscio". I dati reali dovrebbero quasi sempre restare all'interno della fascia definita dalla volatilità stimata (solitamente $\pm 2\sigma$ o $\pm 1.96\sigma$).Reattività: Se c'è un periodo di calma (dati neri piatti), la linea della volatilità deve scendere gradualmente (GARCH decay).3. Cosa indica un fallimento?Se vedi grandi esplosioni nei dati neri (punti molto distanti) ma la linea della volatilità rimane piatta, il modello è troppo lento (sottostima il rischio).Se la linea della volatilità rimane altissima anche quando il mercato è tornato calmo da tempo, il modello è troppo persistente (sovrastima il rischio).


```{r,eval=FALSE}
start_time <- proc.time()

# ── ugarchroll con cache ─────────────────────────────────
if (file.exists("roll.rds")) {
  roll <- readRDS("roll.rds")
  message("✓ roll caricato da cache")
} else {
  c1 <- make_mac_cluster()
  on.exit(stopCluster(c1), add = TRUE)
  set.seed(42)
  roll <- ugarchroll(garch_spec, data = dtrain,
                     n.ahead = 1,
                     forecast.length = length(dtest),
                     refit.every = 10,
                     refit.window = "moving",
                     cluster = c1)
  stopCluster(c1)
  gc(verbose = FALSE)
  saveRDS(roll, "roll.rds")
  message("✓ roll salvato in roll.rds")
}
# ────────────────────────────────────────────────────────

# Gestione non-convergenza
if (length(roll@model$nonconverged) > 0) roll <- resume(roll, solver = "nloptr")
if (length(roll@model$nonconverged) > 0) roll <- resume(roll, solver = "solnp")
if (length(roll@model$nonconverged) > 0) roll <- resume(roll, solver = "hybrid")

# Estrai risultati
roll_df <- as.data.frame(roll)
quants  <- as.data.frame(roll, which = "VaR")

actual <- as.numeric(quants[, "realized"])
VaR_95 <- as.numeric(quants[, "alpha(5%)"])
VaR_99 <- as.numeric(quants[, "alpha(1%)"])

# VaR Test 95%
cat("=== VaR Test 95% ===\n")
var_test_95 <- VaRTest(alpha = 0.05, actual = actual, VaR = VaR_95, conf.level = 0.95)
print(var_test_95)

# VaR Test 99% — Kupiec manuale
cat("=== VaR Test 99% (Kupiec only) ===\n")
n_obs    <- length(actual)
n_viol   <- sum(actual < VaR_99)
expected <- 0.01 * n_obs
lr_stat  <- -2 * (log(0.01^n_viol * 0.99^(n_obs-n_viol)) -
                   log((n_viol/n_obs)^n_viol * (1-n_viol/n_obs)^(n_obs-n_viol)))
p_val    <- 1 - pchisq(lr_stat, df = 1)

cat("Violazioni osservate:", n_viol, "| Attese:", round(expected, 1), "\n")
cat("Kupiec LR statistic:", round(lr_stat, 4), "\n")
cat("Kupiec p-value:", round(p_val, 4), "\n")
if (p_val > 0.05) cat("Esito: VaR 99% correttamente calibrato\n") else cat("Esito: VaR 99% NON calibrato\n")

# Forecast dataframe
dfc_df <- data.frame(
  pred_mu  = roll_df$Mu,
  pred_var = roll_df$Sigma^2,
  pred_vol = roll_df$Sigma,
  ci_q05   = quants[, "alpha(5%)"],
  ci_q95   = quants[, "alpha(1%)"],
  actual   = roll_df$Realized
)
rownames(dfc_df) <- rownames(roll_df)
print(dfc_df)

end_time <- proc.time()
cat("\nTotal execution time:", (end_time - start_time)[3], "seconds\n")
```


VERSIONE PER WINDOWS + CORREZIONE (HO MESSO ANCHE I QUANTILI SUPERIORI PER IL CALCOLO DELLA DISTRIBUTION COVERAGE NEL CHUNCK SUCCESSIVO )

```{r}
start_time <- proc.time()

library(parallel)
library(rugarch)

# ── ugarchroll universale (Windows & Mac) ────────────────────────
if (file.exists("roll.rds")) {
  roll <- readRDS("roll.rds")
  message("✓ roll caricato da cache")
} else {
  # 1. Rileva i core e crea il cluster
  n_cores <- detectCores() - 1
  c1 <- makeCluster(n_cores)
  
  # 2. Carica rugarch su ogni core (necessario su Windows)
  clusterEvalQ(c1, library(rugarch))
  
  on.exit(stopCluster(c1), add = TRUE)
  set.seed(42)
  
  # 3. Esecuzione con tutti i quantili (per evitare l'errore precedente)
  roll <- ugarchroll(garch_spec, data = dtrain,
                     n.ahead = 1,
                     forecast.length = length(dtest),
                     refit.every = 10,
                     refit.window = "moving",
                     VaR.alpha = c(0.01, 0.05, 0.95, 0.99), # Inclusi i quantili alti
                     cluster = c1)
  
  stopCluster(c1)
  gc(verbose = FALSE)
  saveRDS(roll, "roll.rds")
  message("✓ roll salvato in roll.rds")
}
# ────────────────────────────────────────────────────────

# Gestione non-convergenza
if (length(roll@model$nonconverged) > 0) roll <- resume(roll, solver = "nloptr")
if (length(roll@model$nonconverged) > 0) roll <- resume(roll, solver = "solnp")
if (length(roll@model$nonconverged) > 0) roll <- resume(roll, solver = "hybrid")

# Estrai risultati
roll_df <- as.data.frame(roll)
quants  <- as.data.frame(roll, which = "VaR")

actual <- as.numeric(quants[, "realized"])
VaR_95 <- as.numeric(quants[, "alpha(5%)"])
VaR_99 <- as.numeric(quants[, "alpha(1%)"])

# --- VaR Test 95% ---
cat("=== VaR Test 95% (Kupiec only) ===\n")
n_obs    <- length(actual)
n_viol_95 <- sum(actual < VaR_95)
expected_95 <- 0.05 * n_obs
# Calcolo manuale LR per coerenza
lr_stat_95 <- -2 * (log(0.05^n_viol_95 * 0.95^(n_obs-n_viol_95)) -
                    log((n_viol_95/n_obs)^n_viol_95 * (1-n_viol_95/n_obs)^(n_obs-n_viol_95)))
p_val_95 <- 1 - pchisq(lr_stat_95, df = 1)

cat("Violazioni osservate:", n_viol_95, "| Attese:", round(expected_95, 1), "\n")
cat("Kupiec p-value:", round(p_val_95, 4), "\n")
cat(if(p_val_95 > 0.05) "Esito: VaR 95% correttamente calibrato\n" else "Esito: VaR 95% NON calibrato\n")

# --- VaR Test 99% ---
cat("\n=== VaR Test 99% (Kupiec only) ===\n")
n_viol_99 <- sum(actual < VaR_99)
expected_99 <- 0.01 * n_obs
lr_stat_99 <- -2 * (log(0.01^n_viol_99 * 0.99^(n_obs-n_viol_99)) -
                    log((n_viol_99/n_obs)^n_viol_99 * (1-n_viol_99/n_obs)^(n_obs-n_viol_99)))
p_val_99 <- 1 - pchisq(lr_stat_99, df = 1)

cat("Violazioni osservate:", n_viol_99, "| Attese:", round(expected_99, 1), "\n")
cat("Kupiec p-value:", round(p_val_99, 4), "\n")
cat(if(p_val_99 > 0.05) "Esito: VaR 99% correttamente calibrato\n" else "Esito: VaR 99% NON calibrato\n")

dfc_df <- data.frame(
  pred_mu   = roll_df$Mu,
  pred_var  = roll_df$Sigma^2,
  pred_vol  = roll_df$Sigma,
  ci_q01    = quants[, "alpha(1%)"],
  ci_q05    = quants[, "alpha(5%)"],
  #ci_q95    = quants[, "alpha(95%)"],
  #ci_q99    = quants[, "alpha(99%)"],
  actual    = roll_df$Realized
)

rownames(dfc_df) <- rownames(roll_df)
print(dfc_df)

end_time <- proc.time()
cat("\nTotal execution time:", (end_time - start_time)[3], "seconds\n")

```

### Analisi dei Risultati di Backtesting

I test di copertura incondizionata (Kupiec Test) mostrano un comportamento divergente del modello ai due livelli di confidenza analizzati.

#### 1. VaR 95%: Calibrazione Ottimale
Al livello del 95%, il modello risulta **perfettamente calibrato**.
* **Analisi:** Il numero di violazioni osservate (**12**) è identico al valore atteso teorico. 
* **Statistica:** Il p-value di **0.8613** è ampiamente sopra la soglia critica di 0.05, indicando che non vi è alcuna prova statistica per rigettare l'ipotesi di correttezza del modello.

#### 2. VaR 99%: Eccesso di Conservatività
Al livello del 99%, il modello viene respinto dal test statistico ($p = 0.0244$), ma per motivi di **efficienza** piuttosto che di sicurezza.
* **Analisi:** Sono state osservate **0 violazioni** a fronte di un'attesa di **2.5**. 
* **Causa Tecnica (Student-t):** L'utilizzo della distribuzione **t di Student** introduce "code grasse" (leptocurtosi). In assenza di shock estremi nel periodo di test, la distribuzione t tende a sovrastimare la probabilità di eventi rari, spingendo la soglia del VaR molto lontano dalla media.
* **Effetto:** Il modello è "troppo prudente". Sebbene protegga completamente dalle perdite (zero sforamenti), in un contesto operativo richiederebbe un accantonamento di capitale eccessivo e inefficiente.


#### 3. Considerazioni sul Campione (Sample Size)
Il fallimento del test al 99% è influenzato anche dalla dimensione del campione (circa 250 osservazioni). Con un'attesa di sole 2.5 violazioni, la sensibilità del test di Kupiec è estremamente elevata: la differenza tra un modello "calibrato" e uno "non calibrato" è spesso legata alla presenza o assenza di una singola violazione (il cosiddetto "problema dei piccoli campioni").

---
**Conclusione:** Il modello è affidabile per la gestione del rischio standard (95%), ma risulta eccessivamente penalizzante per la gestione del rischio estremo (99%), probabilmente a causa di un parametro *shape* della distribuzione t molto basso o di una bassa volatilità nel periodo di test.

```{r,}
# --- Mean evaluation ---
# Usiamo i nomi definiti nel dataframe dfc_df
errors <- dfc_df$actual - dfc_df$pred_mu

mae  <- mean(abs(as.numeric(errors)), na.rm = TRUE)
mse  <- mean(as.numeric(errors)^2, na.rm = TRUE)
rmse <- sqrt(mse)
mape <- mean(abs(as.numeric(errors) / as.numeric(dfc_df$actual)), na.rm = TRUE)

# Directional Accuracy: il modello azzecca il segno del movimento?
dir_acc <- mean(sign(as.numeric(dfc_df$actual)) == sign(as.numeric(dfc_df$pred_mu)), na.rm = TRUE)

cat("\nMean evaluation metrics:\n")
cat(sprintf("MAE: %.4f | MSE: %.4f | RMSE: %.4f | MAPE: %.4f | Dir. Acc: %.2f%%\n",
            mae, mse, rmse, mape, dir_acc * 100))

# --- Distribution check  ---
# Coverage 90% (tra 5% e 95%)
coverage_90 <- mean(dfc_df$actual >= dfc_df$ci_q05 & dfc_df$actual <= dfc_df$ci_q95, na.rm = TRUE)

# Coverage 98% (tra 1% e 99%)
coverage_98 <- mean(dfc_df$actual >= dfc_df$ci_q01 & dfc_df$actual <= dfc_df$ci_q99, na.rm = TRUE)

cat("\nDistribution evaluation, % of forecast that fall into that interval:\n")
cat(sprintf("90%% coverage: %.4f\n", coverage_90))
cat("\nDistribution evaluation, % of forecast that fall into that interval:\n")
cat(sprintf("98%% coverage: %.4f\n", coverage_98))
```
1. Interpretazione dei Risultati
90% Coverage: 0.9008 (90.08%)

Questo è un risultato quasi perfetto. Significa che lo spread reale è caduto all'interno del tunnel previsto (tra il 5% e il 95%) nel 90% dei casi. Il tuo modello non è né troppo ottimista né troppo pessimista; descrive la realtà con una precisione chirurgica.

98% Coverage: 0.9841 (98.41%)

Anche qui siamo vicinissimi al target del 98% (l'intervallo tra l'1% e il 99%). Questo indica che anche nelle situazioni più estreme, le "code" della tua distribuzione t di Student sono calibrate correttamente.




Poiché la volatilità è la varianza, confrontiamo la Varianza Predetta ($pred\_var$) con lo Shock Reale al quadrato ($Residui^2$).
Aggiungi questo al tuo codice:
Se la correlazione è positiva e significativa, il modello sta catturando i cluster di rischio.


Compare those metrics with a Random Walk model (tomorrow's value =
today's value). To see if our model is actually useful.

```{r,}
dspread_rw <- dspread[length(dtrain):(length(dtrain) + length(dtest) - 1)] |> as.numeric()

mae_rw   <- mean(abs(dspread_rw - dfc_df$actual))
rmse_rw  <- sqrt(mean((dspread_rw - dfc_df$actual)^2))
mape_rw  <- mean(abs((dspread_rw - dfc_df$actual) / dfc_df$actual))
dir_rw   <- mean(sign(dspread_rw) == sign(diff(as.numeric(dfc_df$actual))))

cat("\nEvalutation metrics, Model vs Random Walk:\n")
cat(sprintf("\nModel  | MAE: %.4f | RMSE: %.4f | MAPE: %.2f%% | Dir: %.2f%%\n",
            mae, rmse, mape*100, dir_acc*100))
cat(sprintf("RW   | MAE: %.4f | RMSE: %.4f | MAPE: %.2f%% | Dir: %.2f%%\n",
            mae_rw, rmse_rw, mape_rw*100, dir_rw*100))
```
2. Perché il MAPE non è adatto allo Spread
Il MAPE è una metrica "traditrice" per i tassi di interesse e gli spread per due motivi:

Problema di scala: Se lo spread passa da 1 a 2 punti base, l'errore è solo di 1 bp (trascurabile), ma il MAPE segna un errore del 100%.

Asset che possono annullarsi: Per variabili che fluttuano attorno allo zero o che possono toccarlo, il MAPE perde validità statistica.

```{r, fig.width=12, fig.height=10, out.width="100%",}
dfc_df$date <- as.Date(rownames(dfc_df))

## Adding the ci_q95 and ci_q99
dfc_df$ci_q95 <- 2 * dfc_df$pred_mu - dfc_df$ci_q05  # mirror of lower bound
dfc_df$ci_q99 <- 2 * dfc_df$pred_mu - dfc_df$ci_q01

p_spreads  <- ggplot(dfc_df, aes(x = date)) +
  geom_line(aes(y = actual, color = "Actual"), linewidth = 0.8) +
  geom_line(aes(y = pred_mu, color = "Forecasted"), linewidth = 0.8) +
  geom_line(aes(y = dspread_rw, color = "Random Walk"), linewidth = 0.8) +
  geom_ribbon(aes(ymin = ci_q05, ymax = ci_q95), fill = "blue", alpha = 0.3) +
  scale_color_manual(values = c("Actual" = "black", "Forecasted" = "blue", "Random Walk" = "red")) +
  labs(title = "Spread: Actual vs Forecasted", x = "", y = "Basis points") +
  scale_x_date(date_breaks = "2 month", date_labels = "%b-%y") +
  scale_y_continuous(breaks = scales::breaks_width(20)) +
  theme_minimal() +
  theme(legend.title = element_blank(), legend.position = "bottom")

print(p_spreads)

```

# Forecast

Forecast future value for next week, use in the same structure of the
validated model.

Use the last `window_size` data (the same as validation phase)

```{r,}
# Forecast future value for next week
# Use the last window_size data

# 1. Recupero dati e data (usiamo dspread che è l'oggetto xts/zoo)
window_size <- 110  
data_fit_obj <- tail(dspread, window_size) 
data_fit     <- as.numeric(data_fit_obj)
last_date    <- index(tail(data_fit_obj, 1))

set.seed(42)
future_fit <- ugarchfit(spec = garch_spec, data = data_fit, solver = "gosolnp",
                        solver.control = list(n.restarts = 5, rseed = 42, trace = 0))

# --- Positivity check on the forecast window GARCH fit ---
check_garch_positivity(future_fit, model_name = "ARMA-GARCH (forecast window)")

future_fc_obj <- ugarchforecast(future_fit, n.ahead = 1)

# Estrazione quantili e valori
q_05_val <- as.numeric(quantile(future_fc_obj, probs = 0.05))
q_95_val <- as.numeric(quantile(future_fc_obj, probs = 0.95))

future_fc <- data.frame(
  d_spread = as.numeric(fitted(future_fc_obj)),
  vol      = as.numeric(sigma(future_fc_obj)),
  q_05     = q_05_val,
  q_95     = q_95_val
)

# 2. Funzione di conversione (Assicurati che 'data' sia un vettore numerico o xts)
spread_conversion <- function(value, q_05, q_95, data){
  # Usiamo as.numeric per forzare il calcolo matematico
  last_spread <- as.numeric(tail(data, 1))
  
  res_spread <- last_spread + value
  res_q05    <- last_spread + q_05
  res_q95    <- last_spread + q_95
  
  result_df  <- data.frame(
    spread = res_spread,
    q_05   = res_q05,
    q_95   = res_q95
  )
  return(result_df)
}

# 3. CHIAMATA CRITICA: Usa 'spread' (l'oggetto zoo dei livelli) NON 'df'
future_spread <- spread_conversion(future_fc$d_spread, future_fc$q_05, future_fc$q_95, spread)

# Output
cat("\nForecasted values:\n")
cat("\nSpread on last date (", as.character(last_date), "): ", as.numeric(tail(spread, 1)))
cat("\nSpread forecasted for next week (", as.character(last_date + 7), "):\n")
print(future_spread)
```

## Simulation Forecast (10,000 paths)

Simulating 10,000 paths accounts for the GARCH shock uncertainty, not
just the ARMA mean.

```{r,}
simulation <- ugarchsim(
  future_fit,
  n.sim       = 1,
  m.sim       = 10000,
  startMethod = "sample"
)

mean_sim <- mean(fitted(simulation))
vol_sim  <- mean(sigma(simulation))
q_90     <- quantile(fitted(simulation), probs = c(0.05, 0.95))

cat(sprintf("Mean DS:  %.4f\n", mean_sim))
cat(sprintf("Mean Vol: %.4f\n", vol_sim))
cat(sprintf("Quantiles 5%% | 95%%: %.4f | %.4f\n", q_90[1], q_90[2]))

sim_spread <- spread_conversion(mean_sim, unname(q_90[1]), unname(q_90[2]), spread)
cat("Spread forecast (simulation) for next week:", as.character(last_date + 7), ":\n")
print(sim_spread)
```

Se i risultati di sim_spread sono molto simili a quelli di future_spread
(ottenuti con forecast), significa che il tuo modello è stabile. Se sono
molto diversi, potrebbe esserci molta incertezza o asimmetria nella
distribuzione dei rendimenti.


