---
title: "Multivariate_Model_Forecast"
output: html_document
date: "2026-04-16"
---
```{r, eval=FALSE, include = FALSE}
# ============================================================
# MACOS OPTIMIZATION BLOCK (Apple Silicon)
# ============================================================
rm(list = ls())

# 1. Verifica che R sia nativo ARM64 (non Rosetta 2)
if (R.version$arch != "aarch64") {
  warning("R non sta girando su ARM64. Reinstalla R nativo per Apple Silicon da cran.r-project.org")
} else {
  message("✓ R ARM64 nativo confermato: ", R.version$arch)
}

# 2. Configura parallelismo ottimale per macOS (usa FORK, molto più veloce di SOCKET)
N_CORES <- detectCores(logical = FALSE) - 1  # core fisici, lascia 1 libero
message("✓ Core disponibili per parallelismo: ", N_CORES)

# 3. RNG riproducibile per simulazioni parallele
RNGkind("L'Ecuyer-CMRG")
seed <- 42
set.seed(seed)

# 4. Funzione helper: crea cluster FORK (macOS only)
make_mac_cluster <- function() {
  makeCluster(N_CORES, type = "FORK")  # FORK evita di riesportare variabili
}

# 5. Libera memoria dopo operazioni pesanti
gc_after <- function(expr) {
  result <- force(expr)
  gc(verbose = FALSE)
  result
}

message("✓ Blocco ottimizzazione macOS caricato.")
# ============================================================

```


```{r setup, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

getwd()
library(urca)
library(tseries)
library(quantmod)
library(forecast)
library(rugarch)
library(FinTS)
library(ggplot2)
library(dplyr)
library(httpgd)
library(R.utils)
library(xts)
library(readxl)
library(fBasics)
library(lmtest)
library(sandwich)
library(lubridate)
library(fUnitRoots)
library(stats)
library(graphics)
library(MTS)
library(MASS)
library(zoo)
library(forecast)
library(seastests)
library(strucchange)
```

## Multivariate Model
We now switch models, and we'll use a multivariate model. Ideally we are going to use a VAR model, in which there will be 2 series : one will be the BTP-Bund Spread, and the other will be one between the EURO STR series or the EuroStoxx Banks series.
If we have time, we could also use a trivariate VAR model, incorporating them all.

```{r}
data <- read_excel("Spread_daily_17_04.xlsx")
data <- data |> na.omit() |> rename(date = Date, spread = `IT10DE10=TWEB (BID)`)
data$spread <- data$spread * -1
spread  <- xts(data[, -1], order.by = as.Date(data$date, format = "%d/%m/%y"))
```

Inspect data

```{r}
head(spread)
str(spread)
sum(is.na(spread))
any(is.infinite(spread))
```

Plot the data

```{r, echo = FALSE, fig.width=12, fig.height=10, out.width="100%"}
df_plot <- data.frame(date = index(spread), spread = spread$spread)

p <- ggplot(df_plot, aes(x = date, y = spread)) +
  geom_line() +
  labs(title = "BTP-BUND Spread", y ="Basis Points", x = "") +
  theme_bw()

print(p)
```

```{r, echo=FALSE}
options(scipen = 999)
basicStats(spread)
```

```{r}
risultato_adf <- adfTest(spread, lags = 1, type = "c")
print(risultato_adf)
# Equivalent with auto determination of lags

adf_test <- function(serie, nome_serie = "Serie") {
  cat("\n==============================================\n")
  cat("Stationarity test (ADF):", nome_serie, "\n")
  cat("==============================================\n")
  types <- c("none", "drift", "trend")
  results <- lapply(types, function(t) {
    test <- ur.df(serie, type = t, selectlags = "AIC")
    stat <- test@teststat[1]
    val <- test@cval[1, "5pct"]
    result <- if (stat < val) "Stationary" else "NOT Stationary"
    return(data.frame(
      Type = t,
      ADF_stat = round(stat, 3),
      Tau_5pct = round(val, 3),
      Result = result,
      stringsAsFactors = FALSE
    ))
  })
  table <- bind_rows(results)
  print(table)
}

adf_test(spread, "Spread BTP/Bund")

```

The series is not stationary, so we need to differenciate it to use it in a VAR model.

```{r}
d_spread <- diff(spread) |> na.omit() # we loose the first observation due to differencing
#d_train <- d_spread[1:n_train]
#d_test  <- d_spread[(n_train + 1):nrow(d_spread)]

#cat("\nTraining observations:", nrow(d_train))
#cat("\nTest observations:", nrow(d_test))
#cat("\nTrain period:", as.character(index(d_train)[1]), "->", as.character(index(d_train)[nrow(d_train)]))
#cat("\nTest period:", as.character(index(d_test)[1]), "->", as.character(index(d_test)[nrow(d_test)]),"\n")
```

```{r}
adf_test(d_spread, "Differenced spread BTP/Bund")
```
# now the train series is stationary

```{r, echo=FALSE, fig.width=12, fig.height=10, out.width="100%"}
df_plot <- data.frame(date = index(d_spread), d_spread$spread)

p <- ggplot(df_plot, aes(x = date, y = spread)) +
geom_line() +
geom_hline(yintercept = 0, color = "red") +
labs(title = "Differenced Spread", y ="Basis Points", x = "") +
theme_bw()

print(p)
```

## Multicollinearity Test

We now test for multicollinearity, so now we load the 2 other series.
```{r}
# ==========================================
# PHASE 1: EXCEL DATA ACQUISITION & ALIGNMENT
# ==========================================

# Step 1: Load required libraries
#install.packages("readxl")
getwd()
library(readxl)
library(xts)
library(zoo)

# Importa l'indice VStoxx Europe Volatility

vstoxx_raw <- read_excel("vstoxx_17_04.xlsx")
vstoxx_raw <- vstoxx_raw[-1,]

# 3. Trasforma in Serie Storica
# Assumendo che la Colonna 1 sia la Data e la Colonna 3 sia il Tasso
#estr_xts <- xts(as.numeric(estr_raw[[3]]), order.by = as.Date(estr_raw[[1]]))
#colnames(estr_xts) <- "ESTR"
#estr_clean <- na.remove(estr_xts)

vstoxx_xts <- xts(as.numeric(vstoxx_raw[[2]]), order.by = as.Date(vstoxx_raw[[1]]))
colnames(vstoxx_xts) <- "VStoxx"
vstoxx_xts <- na.omit(vstoxx_xts)

plot(vstoxx_xts, type="l")

# ADF Test for vstoxx data
adf.test(vstoxx_xts)
# p-value = 0.01 -> Already stationary

# Merge the datasets
merged_data_2 <- merge(d_spread, vstoxx_xts, join = "inner")
head(merged_data_2)

clean_data_2 <- na.omit(merged_data_2)

# =========================
# 4. Split train/test 80/20
# =========================
ntest  <- round(nrow(clean_data_2) * 0.2)
ntrain <- nrow(clean_data_2) - ntest

train <- clean_data_2[1:ntrain, ]
test  <- clean_data_2[(ntrain + 1):nrow(clean_data_2), ]

cat("Training observations:", nrow(train), "\n")
cat("Test observations:", nrow(test), "\n")
cat("Train period:", as.character(index(train)[1]), "-", as.character(tail(index(train), 1)), "\n")
cat("Test period:", as.character(index(test)[1]), "-", as.character(tail(index(test), 1)), "\n")


```

Now we should have a matrix with the d_spread value and the vstoxx index value, with matching data.
We also have the train and test matrices.

## MQ Test
We have both series stationary, we now run the qm test to see if we need a VAR model.

```{r}

library(MTS)
library(vars)

# MQ test
mq(train, lag=10)

# We have all lags in which p-value is less than 0.05 ( equal to 0), so we can estimate a VAR model.
```
## VAR Model Estimation
```{r}
# 1. Determine optimal VAR lag length
lag_selection <- VARselect(train, lag.max = 10, type = "const")
lag_selection

# We have optimal lag = 1 chosen by all the models
library(vars)
# Fit the model
var_fit <- VAR(train, p = 2, type = "const")
summary(var_fit)

```
## VAR Diagnostics

```{r}

# ==========================================
# PHASE 4: VAR DIAGNOSTICS
# ==========================================

# 1. Test for leftover Serial Correlation (Portmanteau Test)
# We test up to 10 lags to be safe. We want p-value > 0.05
serial_test <- serial.test(var_fit, lags.pt = 10, type = "PT.asymptotic")
cat("\n--- Serial Correlation Test ---\n")
print(serial_test)

# 2. Test for ARCH Effects (Volatility Clustering)
# We want p-value < 0.05 to justify using DCC-GARCH next
arch_test <- arch.test(var_fit, lags.multi = 5, multivariate.only = TRUE)
cat("\n--- ARCH Effects Test ---\n")
print(arch_test)

## We need the DCC-GARCH model, as there are arch effects (p-value is near 0)

```
## DCC-GARCH Model

```{r}
# ==========================================
# PHASE 5: THE DCC-GARCH MODEL
# ==========================================
library(rugarch)
library(rmgarch)

# 1. Extract the residuals from your fitted VAR model
# The scale (* 100) is crucial! GARCH optimizers fail if the numbers are too tiny (e.g. 0.0001)
var_residuals <- residuals(var_fit) * 100 

# Convert the residuals back into a Time Series (xts) object for plotting later
# We use the dates from diff_data, but remember the VAR loses the first 'p' rows
# So we align the dates by tailing the diff_data
var_residuals_xts <- xts(var_residuals, order.by = tail(index(train), nrow(var_residuals)))
colnames(var_residuals_xts) <- c("Spread_Shock", "VStoxx_Shock")

# 2. Specify the Univariate GARCH Models
# Because the VAR already handled the mean, we set the ARMA order to (0,0) with NO mean!
uspec <- ugarchspec(
  mean.model = list(armaOrder = c(0, 0), include.mean = FALSE),
  variance.model = list(garchOrder = c(1, 1), model = "sGARCH"),
  distribution.model = "norm" # You can change to "std" (Student-t) if it fails to converge
)

# Replicate this specification for both of our variables
mspec <- multispec(replicate(2, uspec))

# 3. Specify the Multivariate DCC-GARCH model
# We set VAR = FALSE because we ALREADY did the VAR manually and are feeding it the residuals!
dcc_spec <- dccspec(
  uspec = mspec, 
  VAR = FALSE,
  dccOrder = c(1, 1), 
  distribution = "mvnorm"
)

# 4. Fit the DCC-GARCH model!
# We feed it our scaled VAR residuals
cat("\nFitting DCC-GARCH...\n")
fit_dcc <- dccfit(dcc_spec, data = var_residuals_xts)

# Print the final estimated model
show(fit_dcc)

# Dynamic conditional correlations over time
dcc_corr <- rcor(fit_dcc)        # array: 2x2xT
rho_t    <- dcc_corr[1,2,]       # time series of rho(BTP-Bund, VSTOXX)
plot(rho_t, type = "l", main = "Dynamic Conditional Correlation")

# Conditional volatilities
vol <- sigma(fit_dcc)

# DCC parameters (alpha + beta, should sum < 1)
coef(fit_dcc)

```
```{r}
library(xts)

rho_t2 <- rcor(fit_dcc)[1, 2, ]
resids <- residuals(var_fit)

plot(zoo(rho_t2, order.by = index(resids)), type = "l",
     main = "Dynamic Conditional Correlation: BTP-Bund vs VSTOXX",
     ylab = "Correlation", xlab = "")
abline(h = mean(rho_t2), col = "red", lty = 2)
```

## Model Diagnostics
```{r}
# 1. Model Diagnostics 
# Standardized residuals should have no ARCH effects remaining
infocriteria(fit_dcc)   # AIC, BIC
plot(fit_dcc, which=1)           # rmgarch gives several diagnostic plots
plot(fit_dcc, which=2)
plot(fit_dcc, which=3)
plot(fit_dcc, which=4)
plot(fit_dcc, which=5)
#colnames(var_fit$y)

# 2. VAR Impulse Response Functions (IRFs)
# This answers "how does the BTP-Bund spread respond to a VSTOXX shock?"
library(vars)
irf_result <- irf(var_fit, impulse = "VStoxx",
                  response = "spread",
                  n.ahead = 20, boot = TRUE)
plot(irf_result)

# 3. DCC-GARCH Volatility Forecast
# A short-horizon forecast of conditional variances and correlations
dcc_forecast <- dccforecast(fit_dcc, n.ahead = 20)
plot(dcc_forecast, which=1)
plot(dcc_forecast, which=2)
plot(dcc_forecast, which=3)


# Extract forecasted correlation
rcor(dcc_forecast)

# 4. Variance Decomposition (FEVD)
# Tells you how much of the BTP-Bund spread variance is explained by VSTOXX shocks

fevd_result <- fevd(var_fit, n.ahead = 20)
plot(fevd_result)
```

## Forecast
```{r}

# Re-fit with out.sample to enable rolling forecast
dcc_fit2 <- dccfit(dcc_spec, data = resids,
                   out.sample = 100)   # hold out last 100 obs

# 1. Unconditional n-ahead forecast
dcc_fore <- dccforecast(dcc_fit2, n.ahead = 20)

# Forecasted correlation
rcor(dcc_fore)   # 2x2xT array

# Forecasted volatilities
sigma(dcc_fore)

# 2. Rolling 1-step-ahead forecast (more meaningful for evaluation)
dcc_roll <- dccforecast(dcc_fit2, n.ahead = 1, n.roll = 99)
rcor(dcc_roll)

# 1-Step Ahead Forecast with VAR
library(vars)

# Forecast 1 step ahead (next trading day)
var_forecast <- predict(var_fit, n.ahead = 1, ci = 0.95)

# Print the forecast
print(var_forecast)

last_spread <- tail(spread, 1)
forecast_level <- last_spread + var_forecast$fcst$spread[1, "fcst"]
forecast_level

```
## Test the model
We now use the test portion of our data, to see the accuracy of our training model

```{r}

library(vars)
library(xts)
library(ggplot2)

# =========================
# Test set: rolling 1-step-ahead out-of-sample forecast
# =========================

train_multi <- train
test_multi  <- test

# 1. Selezione lag ottimale sul train
lag_sel <- VARselect(train_multi, lag.max = 10, type = "const")
p_opt <- as.numeric(lag_sel$selection["AIC(n)"])

cat("Optimal lag selected on train:", p_opt, "\n")

# 2. Rolling 1-step-ahead forecast sulla serie differenziata "spread"
n_test <- nrow(test_multi)

fc_dspread <- numeric(n_test)
real_dspread <- as.numeric(test_multi[, "spread"])

# Among the candidate rolling window sizes, 200 observations produced the lowest out-of-sample forecast error and  # the most plausible final spread estimate, therefore it was selected as the preferred specification.
window_size <- 250

for (i in 1:n_test) {

  if (i == 1) {
    available_data <- train_multi
  } else {
    available_data <- rbind(train_multi, test_multi[1:(i - 1), ])
  }

  if (nrow(available_data) > window_size) {
    current_train <- available_data[(nrow(available_data) - window_size + 1):nrow(available_data), ]
  } else {
    current_train <- available_data
  }

  var_roll <- VAR(current_train, p = p_opt, type = "const")
  pred_1 <- predict(var_roll, n.ahead = 1, ci = 0.95)

  fc_dspread[i] <- as.numeric(pred_1$fcst$spread[1, "fcst"])
}


# 3. Date del test
test_dates <- index(test_multi)

# 4. Ricostruzione dello spread in livelli
# ultimo valore osservato in livelli prima dell'inizio del test
test_start_date <- test_dates[1]
pos_test_start <- which(index(spread) == test_start_date)
last_level_before_test <- as.numeric(spread[pos_test_start - 1])

# forecast in livelli
fc_level <- last_level_before_test + cumsum(fc_dspread)

# valori reali in livelli
real_level <- as.numeric(spread[test_dates])

# 5. Data frame finale
df_roll <- data.frame(
  Date = as.Date(test_dates),
  RealSpread = real_level,
  ForecastSpread = fc_level,
  RealDiff = real_dspread,
  ForecastDiff = fc_dspread
)

# 6. Grafico livelli
ggplot(df_roll, aes(x = Date)) +
  geom_line(aes(y = RealSpread, color = "Real spread"), linewidth = 1) +
  geom_line(aes(y = ForecastSpread, color = "Rolling forecast"),
            linewidth = 1, linetype = "dashed") +
  labs(
    title = "Rolling 1-Step-Ahead VAR Forecast vs Real Spread",
    y = "Spread (bps)",
    x = "Date",
    color = "Legend"
  ) +
  theme_bw()

# 7. Metriche sui livelli
err_level <- df_roll$RealSpread - df_roll$ForecastSpread

mae_level <- mean(abs(err_level), na.rm = TRUE)
rmse_level <- sqrt(mean(err_level^2, na.rm = TRUE))

cat("MAE levels:", round(mae_level, 4), "\n")
cat("RMSE levels:", round(rmse_level, 4), "\n")

# 8. Benchmark naive
naive_fc_level <- rep(last_level_before_test, length(fc_level))

mae_naive <- mean(abs(real_level - naive_fc_level), na.rm = TRUE)
rmse_naive <- sqrt(mean((real_level - naive_fc_level)^2, na.rm = TRUE))

cat("Naive MAE:", round(mae_naive, 4), "\n")
cat("Naive RMSE:", round(rmse_naive, 4), "\n")

# 9. Valore finale out-of-sample 1-step-ahead
final_forecast_date <- tail(test_dates, 1)
final_forecast_diff <- tail(fc_dspread, 1)
final_forecast_level <- tail(fc_level, 1)

cat("Final forecast date:", as.character(final_forecast_date), "\n")
cat("Final out-of-sample 1-step forecast (diff):", round(final_forecast_diff, 4), "\n")
cat("Final out-of-sample 1-step forecast (level):", round(final_forecast_level, 4), "\n")





```




