---
title: "BTP-BUND Spread"
output:
  #html_document:
  #  df_print: paged
  
  pdf_document:
    latex_engine: xelatex
date: "2026-04-10"
editor_options:
  markdown:
    wrap: 72
---

# 1. Data setup

```{r, message=FALSE, warning=FALSE, include=FALSE}
getwd()
library(urca)
library(tseries)
library(quantmod)
library(forecast)
library(rugarch)
library(FinTS)
library(ggplot2)
library(dplyr)
library(httpgd)
library(R.utils)
library(xts)
library(readxl)
library(fBasics)
library(lmtest)
library(sandwich)
library(lubridate)
library(fUnitRoots)
library(stats)
library(graphics)
library(MTS)
library(MASS)
library(zoo)
library(forecast)
library(seastests)
library(strucchange)
```

```{r, eval=FALSE, include = FALSE}
# ============================================================
# MACOS OPTIMIZATION BLOCK (Apple Silicon)
# ============================================================

# 1. Verifica che R sia nativo ARM64 (non Rosetta 2)
if (R.version$arch != "aarch64") {
  warning("R non sta girando su ARM64. Reinstalla R nativo per Apple Silicon da cran.r-project.org")
} else {
  message("✓ R ARM64 nativo confermato: ", R.version$arch)
}

# 2. Configura parallelismo ottimale per macOS (usa FORK, molto più veloce di SOCKET)
N_CORES <- detectCores(logical = FALSE) - 1  # core fisici, lascia 1 libero
message("✓ Core disponibili per parallelismo: ", N_CORES)

# 3. RNG riproducibile per simulazioni parallele
RNGkind("L'Ecuyer-CMRG")
seed <- 42
set.seed(seed)

# 4. Funzione helper: crea cluster FORK (macOS only)
make_mac_cluster <- function() {
  makeCluster(N_CORES, type = "FORK")  # FORK evita di riesportare variabili
}

# 5. Libera memoria dopo operazioni pesanti
gc_after <- function(expr) {
  result <- force(expr)
  gc(verbose = FALSE)
  result
}

message("✓ Blocco ottimizzazione macOS caricato.")
# ============================================================

```
Load the data and parse from excel file, daily spread.

```{r, include=FALSE}
rm(list= ls())
getwd()
data <- read_excel("spread_daily_24_04.xlsx")
data <- data |> na.omit() |> rename(date = Date, spread = `IT10DE10=TWEB (BID)`)
data$spread <- as.numeric(data$spread) * -1
na.omit(data$date)

spread  <- xts(data[, -1], order.by = as.Date(data$date))
```

Inspect data

```{r}
head(spread)
str(spread)
sum(is.na(spread))
any(is.infinite(spread))
```

Plot the data

```{r, echo = FALSE, fig.width=12, fig.height=10, out.width="100%"}
df_plot <- data.frame(date = index(spread), spread = spread$spread)

p <- ggplot(df_plot, aes(x = date, y = spread)) +
  geom_line() +
  labs(title = "BTP-BUND Spread", y ="Basis Points", x = "") +
  theme_bw()

print(p)
```

```{r, echo=FALSE}
options(scipen = 999)
basicStats(spread)
```

The descriptive analysis of the daily BTP-BUND spread reveals a distribution with a mean of 141.05 bps and a median of 134.65 bps. The fact that the mean remains above the median confirms a persistent positive skewness, indicating that the distribution is influenced by periodic upward shocks rather than frequent downward drops.

Visual inspection of the daily time series identifies three distinct market regimes:

The 2022 Surge: A sharp escalation where the spread moved from its minimum of 59.80 bps toward the series maximum of 250.70 bps, likely reflecting heightened political or macroeconomic uncertainty.

The 2023–2024 Volatility: A period of consolidation characterized by significant fluctuations, yet largely contained within the interquartile range of 105.50 bps to 179.03 bps.

The 2025–2026 Mean Reversion: A sustained downward trend where the spread compressed significantly, approaching the lower bounds of the sample before a slight uptick at the end of the period.

Econometrically, the daily frequency highlights the "jump-diffusion" nature of sovereign spreads. While the market spends the majority of its time in a state of relative calm—as evidenced by the median and the 1st quartile of 105.50 bps—it remains susceptible to rapid "tail risk" events. These episodes of stress pull the mean upward, creating the characteristic right-skewed profile typical of risk-sensitive financial instruments.

**Forecast n°1 : naive forecast (mean)**

```{r, echo = FALSE, fig.width=12, fig.height=10, out.width="100%"}
naive <- mean(spread)

p_naive <- ggplot(df_plot, aes(x = date, y = spread)) +
  geom_line() +
  geom_hline(yintercept = naive, color = "red", linetype = "dashed", linewidth = 1) +
  annotate("text", x = end(spread) - 300, y = naive + 4, 
           label = paste("Naive Forecast (mean): ", round(naive, 2)),
           color = "red", hjust = 0) +
  labs(title = "BTP-BUND Spread with Naive Forecast", y ="Basis Points", x = "") +
  theme_bw()

print(p_naive)
```

# 2. Test for seasonality

We use seastests package

```{r}
start_date<- index(spread)[1]
start_year<-year(start_date)
start_month<-month(start_date)
start_day <-day(start_date)
spread_ts <- ts(spread, start=c(start_year,start_month,start_day), frequency = 252) #annual frequency

# Combined test
cat("\nSeasonality Test Results:\n")
cat("Combined Test: ", isSeasonal(spread_ts, test="combined"), "\n")

#Individual tests
cat("\nIndividual Test Results:\n")
cat("QS Test: ", isSeasonal(spread_ts, test="qs"), "\n")
cat("Seasdum Test: ", isSeasonal(spread_ts, test="seasdum"), "\n")
cat("KW Test: ", isSeasonal(spread_ts, test="kw"), "\n")
cat("Fried Test: ", isSeasonal(spread_ts, test="fried"), "\n")
```

Plot ACF and PACF
```{r, echo = FALSE, fig.width=12, fig.height=10, out.width="100%"}
par(mfrow = c(2, 1))
Acf(spread_ts, lag.max= 52)
Pacf(spread_ts, lag.max = 52)
par(mfrow = c(1, 1))
```
We have 1 significant lag at 1 in the Pacf, and a Acf which gradually goes to 0. This suggests no seasonality.

Last point, let's fit a ARIMA model
```{r}
adf.test(spread_ts)     
suppressWarnings(kpss.test(spread_ts)) 
```
- ADF test: we have p-value > 0.05, so we fail to reject the null, so we are in the null workspace, this means no seasonality.
- KPSS test: p-value < 0.05 -> we REJECT the null, so the series is NOT stationary

## 2.1 Visualization for the no-seasonality

```{r, echo = FALSE, fig.width=12, fig.height=10, out.width="100%"}

# 1. Seasonal Plot 
# Mostra l'andamento dello spread sovrapponendo i periodi (2021-2026)
p1 <- ggseasonplot(spread_ts, year.labels = TRUE, year.labels.left = TRUE) +
  theme_minimal() +
  labs(title = "Seasonal Effect: BTP-BUND (Annual Pattern)",
       subtitle = "Daily data (2021-2026) - Annual Cycle",
       x = "Day of the Year (Trading Days)",
       y = "Spread (bps)",
       color = "Year") +
       theme_bw()
print(p1)

# 2. Subseries Plot (Medie Mensili)
# Nota: Con freq = 252, il subseries plot potrebbe tentare di fare 252 piccoli grafici.
# Per vedere le medie mensili "storiche", è meglio aggregare prima i dati o 
# usare una frequenza mensile (12) se l'obiettivo è l'analisi macro.
# Se vuoi comunque il subseries sulle 52 settimane:
#spread_weekly_avg <- aggregate(spread_ts, nfrequency = 52)
p2 <- ggsubseriesplot(spread_ts) +
  theme_minimal() +
  labs(title = "Subseries Plot: Daily data",
       subtitle = " 2021-2026",
       x = "Week of the Year",
       y = "Spread (bps)") +
       theme_bw()
print(p2)

# 3. STL Decomposition (Multi-Year)
# Utilizziamo un s.window ampio per catturare la stagionalità annuale
decomposed <- stl(spread_ts[,1], s.window = "periodic", robust = TRUE)

p3 <- autoplot(decomposed) +
  theme_minimal() +
  labs(title = "STL Decomposition (2021-2026)",
       subtitle = "Trend, Annual Seasonality, and Daily Remainder") +
       theme_bw()
print(p3)
```

Seasonal and Subseries Plot Interpretation:
- The seasonal plots further investigate whether specific times of the year are inherently riskier for the Italian spread:
- Subseries Plot (Weekly): The blue dashed lines represent the historical mean for each specific week of the year across the 2021-2026 period. The relatively flat nature of these mean lines suggests that there is no "structural" week of the year where the spread is consistently higher. The high density of the black vertical lines reflects the daily volatility within those weeks.
- Annual Pattern (Seasonal Plot): By overlaying the years, we can see that the 2022-2024 period significantly outperformed the historical averages, creating the "Surge" identified in the descriptive statistics. The 2025-2026 lines are positioned lower, confirming the transition into a period of market calm.

The STL (Seasonal and Trend decomposition using Loess) plot provides a granular look at the forces driving the spread:
- Trend: The trend component shows a steady appreciation starting in 2021, reaching a peak during the 2024-2025 period. Notably, toward the beginning of 2026, the trend exhibits a clear mean-reversion phase, with the spread beginning to compress. This suggests that long-term macroeconomic factors were the primary drivers of the spread's expansion and subsequent cooling.
- Seasonal: The annual seasonal component is present but relatively low in magnitude (oscillating between -10 and +15 bps). This indicates that while there are minor "calendar effects," they are secondary to the overall trend and unpredictable market shocks.
- Remainder: This component represents "noise" or exogenous shocks. We observe significant density and volatility in the remainder during the 2024-2025 period. These spikes represent "tail risk" events—sudden political or financial news that caused the spread to jump independently of any seasonal or long-term trend.

**Econometric Summary**
From an econometric perspective, the BTP-BUND spread during this period is a trend-dominated series. The influence of seasonality is statistically marginal compared to the impact of the Remainder (shocks). This confirms that the spread is more sensitive to "news arrival" and discretionary policy changes than to cyclical or seasonal time-effects.

# 3. Model Specification
**Levels vs. Logarithmic Transformations**
A critical decision in modelling the BTP-BUND spread involves the choice
between raw levels (basis points) and logarithmic transformations. While
log-returns are standard for equity prices to ensure normality and
constant variance, this study follows the established econometric
literature—notably **Campbell, Lo, and MacKinlay (1997)** and **Hamilton
(1994)**—by analyzing the spread in **levels**.

As argued by **Engle (1982)** in his seminal work on ARCH models,
interest rate differentials are already expressed in percentage terms;
therefore, a log-transformation is not only redundant but can lead to
significant numerical instability as values approach the **Zero Lower
Bound (ZLB)**. Maintaining the analysis in basis points ensures the
direct economic interpretability of the results, which is essential for
assessing credit and liquidity risk premiums.

**Literature Review on Model Specification**

The choice to model the BTP-BUND spread in first differences of its
levels is supported by the foundational empirical work on Eurozone
sovereign risk. **Codogno, Favero, and Missale (2003)** and **Favero et
al. (2010)** demonstrate that interest rate differentials are best
captured in basis points to maintain linear properties between credit
risk and liquidity factors. Furthermore, the application of a **GARCH
framework**, as pioneered by **Bollerslev (1986)**, ensures that the
time-varying volatility of the spread is explicitly parameterized rather
than suppressed via logarithmic transformations.

## 3.1 Data Partitioning: Training and Test Sets

To evaluate the predictive performance of the **ARIMA model**, the
dataset is partitioned into training and testing subsets. The training
phase utilizes a **rolling window approach** with a fixed length of 

```{r}
n_test  <- round(nrow(spread) * 0.2)
n_train <- nrow(spread) - n_test

train <- spread[1:n_train]
test  <- spread[(n_train + 1):nrow(spread)]

cat("\nTraining observations:", n_train)
cat("\nTest observations:", n_test)
cat("\nTrain period:", as.character(index(train)[1]), "->", as.character(index(train)[nrow(train)]))
cat("\nTest period:", as.character(index(test)[1]), "->", as.character(index(test)[nrow(test)]),"\n")
```

```{r, echo = FALSE, fig.width=12, fig.height=10, out.width="100%"}
p_show <- ggplot(df_plot, aes(x = date, y = spread)) +
  geom_line() +
  geom_vline(xintercept = index(test)[1], color = "red", linetype = "dashed",
             linewidth = 0.8) +
  labs(title = "BTP-BUND Spread: Training vs Test Set",
       subtitle = paste("Breakpoint:", index(test)[1]),
       x = "", y = "Basis Points") +
  theme_bw()

print(p_show)
```

## 3.2 Stationarity Testing

To ensure the statistical validity of the ARIMA-GARCH framework, we
first evaluate the integration order of the series. Financial variables
often exhibit non-stationary behavior, characterized by a "random walk".

### 3.2.1 ACF and PACF
```{r, echo = FALSE, fig.width=12, fig.height=10, out.width="100%"}
par(mfrow = c(2, 1))
Acf(as.numeric(train))
Pacf(as.numeric(train))
par(mfrow = c(2, 1)) 
```

The initial inspection of the **Autocorrelation Function (ACF)** reveals
a very slow, near-linear decay. In time series econometrics, this
pattern is a classic diagnostic for a **"near-unit root"** process,
indicating that the AR component is highly persistent and the series is
likely non-stationary.

### 3.2.2 Formal Stationarity Testing: The ADF Approach

To rigorously confirm the integration order of the BTP-BUND spread, we
employ the **Augmented Dickey-Fuller (ADF) test** using the `ur.df()`
function from the `urca` package. Unlike standard implementations, this
approach requires a direct comparison between the calculated test
statistics ($\tau$) and the critical values provided by **Dickey and
Fuller (1979)**.

The test is conducted following a hierarchical strategy, accounting for
different deterministic components:

1. None (`type = "none"`): To test for stationarity around zero.
2. Drift (`type = "drift"`): To test for stationarity around a constant non-zero mean.
3. Trend (`type = "trend"`): To check if the series is stationary around a deterministic time trend.

**Decision rule**
For the `ur.df()` output, we focus on the **tau statistic (**$\tau$).
The null hypothesis ($H_0$) of a unit root is rejected only if the test
statistic is **more negative** (i.e., lies in the lower tail) than the
critical value at the 5% significance level.

If the series in levels is found to be non-stationary, we proceed with
**first-differencing** ($\Delta \text{Spread}_t$) and re-apply the test.
The transformation is considered successful only when the tau statistic
for the differenced series falls significantly below the critical
threshold, confirming that the process is $I(1)$ and ready for GARCH
modelling.

```{r}
risultato_adf <- adfTest(train, lags = 1, type = "c")
print(risultato_adf)
# Equivalent with auto determination of lags

adf_test <- function(serie, nome_serie = "Serie") {
  cat("\n==============================================\n")
  cat("Stationarity test (ADF):", nome_serie, "\n")
  cat("==============================================\n")
  types <- c("none", "drift", "trend")
  results <- lapply(types, function(t) {
    test <- ur.df(serie, type = t, selectlags = "AIC")
    stat <- test@teststat[1]
    val <- test@cval[1, "5pct"]
    result <- if (stat < val) "Stationary" else "NOT Stationary"
    return(data.frame(
      Type = t,
      ADF_stat = round(stat, 3),
      Tau_5pct = round(val, 3),
      Result = result,
      stringsAsFactors = FALSE
    ))
  })
  table <- bind_rows(results)
  print(table)
}

adf_test(train, "Spread BTP/Bund")

```

### 3.2.3 Transformation and Diagnostic of Differenced Series Following the
identification of a unit root in the levels of the spread, we proceed
with **first-differencing** to achieve stationarity, transforming the
series into an $I(0)$ process. This transformation represents the daily
change in the BTP-BUND spread.

```{r}
d_spread <- diff(spread) |> na.omit() # we loose the first observation due to differencing
d_train <- d_spread[1:n_train]
d_test  <- d_spread[(n_train + 1):nrow(d_spread)]

cat("\nTraining observations:", nrow(d_train))
cat("\nTest observations:", nrow(d_test))
cat("\nTrain period:", as.character(index(d_train)[1]), "->", as.character(index(d_train)[nrow(d_train)]))
cat("\nTest period:", as.character(index(d_test)[1]), "->", as.character(index(d_test)[nrow(d_test)]),"\n")
```

Run again the ADF test to check for stationarity of the differenced series
```{r}
adf_test(d_train, "Differenced spread BTP/Bund")
```

```{r, echo=FALSE, fig.width=12, fig.height=10, out.width="100%"}
df_plot <- data.frame(date = index(d_train), d_train$spread)

p <- ggplot(df_plot, aes(x = date, y = spread)) +
geom_line() +
geom_hline(yintercept = 0, color = "red") +
labs(title = "Differenced Spread, train data", y ="Basis Points", x = "") +
theme_bw()

print(p)
```

We can see that the differenced series is now stationary around 0 mean (`Type = none`). 
The ADF test confirms that the null hypothesis of a unit root is rejected at the 5% level, indicating that the differenced spread is now an $I(0)$ process suitable for ARIMA-GARCH modelling.

### 3.2.4 Autocorrelation Analysis of the Differenced Spread and its Squared Values
The next step is to analyze the autocorrelation structure of the differenced series and its squared values to detect the presence of conditional heteroskedasticity, which is a key feature in financial time series that justifies the use of GARCH models.

Plot ACF, PACF for differenced spread and for its square (to detect volatilit clustering)
```{r, echo=FALSE, fig.width=12, fig.height=10, out.width="100%"}
par(mfrow = c(4, 1), mar = c(3, 4, 3, 2))
Acf(d_train, lag.max = 30, main = "ACF of Diff Spread")
Pacf(d_train, lag.max = 30, main = "PACF of Diff Spread")
Acf(d_train^2, lag.max = 30, main = "ACF of Squared Diff Spread")
Pacf(d_train^2, lag.max = 30, main = "PACF of Squared Diff Spread")
par(mfrow = c(1, 1))
```

The plots shows a clear absence of significant autocorrelation in the differenced series, this suggest that the series follows a White Noise process (common in financial series).
Also that there's a significant lag 12 but is too far away, its inclusion could make the model too complex and cause overfitting.

While the squared differenced series exhibits significant spikes, indicating the presence of volatility clustering and justifying the use of GARCH models for further analysis.

## 3.3 Seasonality Diagnostics 
We perform a definitive check for seasonality effects on the differenced spread.

#### 3.3.1 Power Spectrum and Seasonality Tests 
The periodogram is utilized to analyze frequency strength. A pronounced peak at a specific frequency  would indicate
a periodic component. Furthermore, we apply the seastests package for a combined formal verification.

```{r, fig.width=12, fig.height=10, out.width="100%",}
# Calcola lo spettro di potenza
spec_raw <- spec.pgram(as.numeric(d_train), spans = c(3, 3),
                       taper = 0.1, plot = TRUE, log="no", 
                       main="Periodogram of Differenced Spread")
# Trova la frequenza con il picco massimo
max_freq <- spec_raw$freq[which.max(spec_raw$spec)]
periodo_dominante <- 1 / max_freq
cat("Dominant period: ", periodo_dominante)
```

**Spectral Analysis Interpretation**

The **Smoothed Periodogram** analysis provides a frequency-domain perspective that complements the previous time-series diagnostics:

- **Frequency Distribution:** The spectrum shows multiple significant peaks rather than a single dominant frequency. This confirms that the BTP-BUND spread is a **multi-cyclical series** where seasonality is fragmented across different time horizons.
- **Short-term Dynamics:** The prominent peaks at higher frequencies (e.g., near 0.4) represent the daily "noise" and short-term volatility. This aligns with the **Maximum value of 250.70 bps**, indicating that sharp, short-lived shocks are a structural feature of the spread's behavior.
- **Low-Frequency Persistence:** The power concentrated near the origin (frequency → 0) represents the **Trend component**. This captures the long-term shifts (2022 surge and 2025 reversion) which, despite the daily noise, remain the most powerful drivers of the spread's variance.

**Conclusion:** The spectral density confirms that while no simple seasonal rule can predict the spread, the series is far from being pure white noise. The complexity of the periodogram justifies the use of non-linear models (like GARCH) to capture the clusters of energy (volatility) that appear at various frequencies.

```{r}
start_date<- index(d_spread)[1]
start_year<-year(start_date)
start_month<-month(start_date)
start_day <-day(start_date)
d_spread_ts <- ts(d_spread, start=c(start_year,start_month,start_day), frequency = 21) 

# Combined test
cat("\nSeasonality Test Results:\n")
cat("Combined Test: ", isSeasonal(d_spread_ts, test="combined"), "\n")

#Individual tests
cat("\nIndividual Test Results:\n")
cat("QS Test: ", isSeasonal(d_spread_ts, test="qs"), "\n")
cat("Seasdum Test: ", isSeasonal(d_spread_ts, test="seasdum"), "\n")
cat("KW Test: ", isSeasonal(d_spread_ts, test="kw"), "\n")
cat("Fried Test: ", isSeasonal(d_spread_ts, test="fried"), "\n")
```

Visualization for the no-seasonality
```{r, echo = FALSE, fig.width=12, fig.height=10, out.width="100%"}
# 1. Seasonal Plot 
p1 <- ggseasonplot(d_spread_ts, year.labels = TRUE, year.labels.left = TRUE) +
  theme_minimal() +
  labs(title = "Seasonal Effect: Diff BTP-BUND (Monthly Pattern)",
       subtitle = "Daily data (2021-2026) - Monthly Cycle",
       x = "Day of the Year (Trading Days)",
       y = "DiffSpread (bps)",
       color = "Year") +
       theme_bw()
print(p1)

# 2. Subseries Plot (Medie Mensili)
p2 <- ggsubseriesplot(d_spread_ts) +
  theme_minimal() +
  labs(title = "Subseries Plot: Daily data",
       subtitle = " 2021-2026",
       x = "Week of the Year",
       y = "Diff Spread (bps)") +
       theme_bw()
print(p2)

# 3. STL Decomposition (Multi-Year)
decomposed <- stl(d_spread_ts[,1], s.window = "periodic", robust = TRUE)

p3 <- autoplot(decomposed) +
  theme_minimal() +
  labs(title = "STL Decomposition (2021-2026)",
       subtitle = "Trend, Monthly Seasonality, and Daily Remainder") +
       theme_bw()
print(p3)
```

The decomposition of the BTP-BUND spread at a 21-day frequency reveals a highly active cyclical structure:
-  **Trend:** The trend component remains the primary driver, capturing the broad expansion of the spread from 2021 through 2024 and the subsequent compression toward 2026. At this frequency, the trend is more reactive to mid-term economic shifts compared to the annual model ($f=252$).
-  **Seasonal (Monthly):** The seasonal panel displays a persistent, high-frequency oscillation. However, the amplitude is notably small, ranging between **-1 and +1.5 bps**. This suggests that while a monthly cycle exists, its statistical contribution to the total variance is marginal.(Se il tuo spread medio è di 150 bps e la volatilità giornaliera (Remainder) ha picchi di 10-20 bps, un'oscillazione di soli 3 bps è considerata trascurabile (noise-level).)
-  **Remainder:** The remainder component continues to exhibit significant "bursts" or clusters of volatility. This indicates that sudden market news (stochastic shocks) has a much larger impact on daily spread changes than any predictable monthly calendar effect.

All individual tests (QS, Seasonal Dummy, Kruskal-Wallis, and Friedman)
consistently fail to reject the null hypothesis of no seasonality.

Conclusion: Having confirmed that the series is stationary, lacks
seasonality, and exhibits significant ARCH effects, we can proceed with
confidence to estimate the non-seasonal ARIMA-GARCH framework.

## 3.4 Autocorrelation and ARCH tests
To ensure the necessity of both the mean and variance components of the model, we
perform rigorous diagnostic tests on the differenced training set
($dtrain$). We select a high lag (e.g., $lag = 16$) to be meticulous in
detecting long-term dependencies.

```{r}
# 1. Ljung-Box Test on differences
lb_ret <- Box.test(as.numeric(d_train), lag = 16, type = "Ljung-Box")
cat("\nLB Test on ds p-value:\n", lb_ret$p.value, "\n")
if(lb_ret$p.value < 0.05) cat("Autocorrelation detected, need mean modelling\n") else cat("ds are White Noise\n")

# 2. Ljung-Box Test on squared differences
lb_test <- Box.test(as.numeric(d_train)^2, lag = 16, type = "Ljung-Box")
cat(sprintf("\nLjung-Box on Squared ds at lag 16:\n %.4f\n", lb_test$p.value))
if(lb_test$p.value < 0.05) cat("Volatility clustering confirmed\n") else cat("No volatility clustering detected\n")

# 3. ARCH-LM Test (Engle's Test for ARCH effects)
arch_test <- ArchTest(d_train, lags = 16)
cat(sprintf("\nARCH Test (16 lags) LM Test - statistic: %.4f, p-value: %.4f\n", arch_test$statistic, arch_test$p.value))
if(arch_test$p.value < 0.05) cat("ARCH effects confirmed. GARCH modelling needed\n") else cat("No ARCH effects confirmed\n")
```

# 4 ARIMA estimation
We finally estimate the ARIMA model on the differenced training set, use `auto_arima()` to find the best arima according to `AIC`criteria.

Keep ARMA order as low as possible to not overfit or model noise.

```{r}
m1 <- auto.arima(d_train,max.q = 16, max.p = 16,
                        seasonal = TRUE,
                        stepwise = FALSE,
                        trace = FALSE)

cat("\nBest ARIMA model:\n")
print(summary(m1))
```
The estimated model is an ARMA(3,2), that's the best ARMA model for our train set, now we will run some diagnostic tests to see if correctly specified.

We specify also the model m0=ARIMA(0,0,0) to compare the performances
```{r}
# Fit ARIMA(1,0,0) - equivalent to AR(1)
m0 <- Arima(as.numeric(d_train), order = c(0, 0, 0))

# Summary of coefficients and information criteria
summary(m0)
```

We can see from both AIC/BIC and error measures that the models are pretty much similar, model m1 add no improvements to simple white noise.

## 4.1 Coefficients Significance
Statistical Significance of CoefficientsTo ensure model parsimony, we
evaluate the significance of each parameter. A coefficient is
statistically significant at the 95% confidence level if its t-ratio
($t = \frac{\hat{\phi}}{SE(\hat{\phi})}$) exceeds the critical value of
$|1.96|$.

```{r}
check_coef_significance <- function(model) {
  coefs <- model$coef
  se <- sqrt(diag(model$var.coef))
  t_stats <- coefs / se
  # Calcoliamo i p-value (distribuzione normale a due code)
  p_values <- (1 - pnorm(abs(t_stats))) * 2
  results <- data.frame(
    Coeff = round(coefs, 4),
    Std_Error = round(se, 4),
    t_stat = round(t_stats, 4),
    p_value = round(p_values, 4),
    Significance_95 = ifelse(p_values < 0.05, "Yes", "No")
  )
  cat("\nCoefficients significance:\n")
  print(results)
}
check_coef_significance(m1)
```

All the coefficients are significant.

## 4.2 ARMA assumptions and Cycle Analysis
```{r}
arma_assumptions <- function(m) {
  cat("\n============================================================\n")
  cat("ARIMA model assumptions & Cycle Analysis:\n")
  
  # Estrazione corretta per modelli auto.arima/Arima
  # Usiamo i coefficienti AR per la stazionarietà
  ar_coefs <- m$model$phi
  ma_coefs <- m$model$theta
  
  # 1. Calcolo Radici
  ar_roots <- polyroot(c(1, -ar_coefs))
  ma_roots <- polyroot(c(1, ma_coefs))
  
  is_stationary <- all(abs(ar_roots) > 1)
  is_invertible <- all(abs(ma_roots) > 1)
  
  cat("Stationary:", is_stationary, "(Min mod:", round(min(abs(ar_roots)), 3), ")\n")
  cat("Invertible:", is_invertible, "(Min mod:", round(min(abs(ma_roots)), 3), ")\n")
  
  # 2. Analisi dei Business Cycles (Radici Complesse)
  # Cerchiamo se esistono radici con parte immaginaria significativa
  complex_roots <- ar_roots[Im(ar_roots) != 0]
  
  cat("\n--- Business Cycle Analysis ---\n")
  if(length(complex_roots) > 0) {
    cat("The model IMPLIES the existence of cycles.\n")
    cat("Reason: The AR characteristic equation has complex conjugate roots.\n")
    
    # Calcolo del periodo del ciclo (se ti serve per l'analisi)
    # Periodo = 2*pi / acos(Re(root)/abs(root))
    root <- complex_roots[1]
    period <- 2 * pi / atan2(Im(root), Re(root))
    cat("Estimated average cycle period:", round(abs(period), 2), "units of time.\n")
  } else {
    cat("The model DOES NOT imply cycles.\n")
    cat("Reason: All AR roots are real; the model returns to equilibrium monotonically.\n")
  }
  
  # 3. Check for common roots
  common_roots <- FALSE
  if(length(ar_roots) > 0 && length(ma_roots) > 0) {
    for(ar in ar_roots) {
      for(ma in ma_roots) {
        if(isTRUE(all.equal(ar, ma, tolerance = 1e-3))) common_roots <- TRUE
      }
    }
  }
  
  cat("\n--- Root Validation ---\n")
  if(common_roots) {
    cat("Common roots detected, assumption violation\n")
  } else {
    cat("No common roots detected, assumption validated\n")
  }
  cat("============================================================\n")
}

# Esecuzione
arma_assumptions(m1)
```

## 4.3 ARMA residuals diagnostics
For distribution we use a Kolmogorov-Smirnov test (KS) to test wheter they follow a t-distribution $H_0: \text{serie follows supposed distribution}$

```{r}
arima_res <- residuals(m1)

arma_residual <- function(res) {

  lb_arima <- Box.test(res, lag = 16, type = "Ljung-Box")
  cat("\nLB test on ARIMA residuals (16 lags) p-value: ", round(lb_arima$p.value, 4), "\n")
  if(lb_arima$p.value >= 0.05) cat("ARIMA model cleared the mean\n") else cat("ARIMA model DID NOT clear the mean\n")
  
  lb_arima_2 <- Box.test(res^2, lag = 16, type = "Ljung-Box")
  cat("\nLB test on ARIMA squared residuals (16 lags) p-value: ", round(lb_arima_2$p.value, 4), "\n")
  if(lb_arima_2$p.value < 0.05) cat("ARIMA model has residual variance to explain. GARCH model needed\n") else cat("ARIMA model residual variance is OK\n")
  
  # ARCH test
  arch_test_res <- ArchTest(res, lags = 10)
  cat(sprintf("\nARCH Test on ARIMA (10 lags) residuals:\n LM Test - statistic: %.4f, p-value: %.4f\n", 
              arch_test_res$statistic, arch_test_res$p.value))
  if(arch_test_res$p.value < 0.05) cat("ARCH effects confirmed. GARCH modelling needed\n") else cat("No ARCH effects confirmed\n")

  # Distribution
  jb_test <- jarque.bera.test(arima_res)
  if(jb_test$p.value < 0.05) cat("\nResiduals are NOT normally distributed\n") else cat("\nResiduals are normally distributed\n")

  check_t <- function(residuals) {
  fit_t <- suppressWarnings(fitdistr(residuals, densfun = "t"))
  mu <- fit_t$estimate["m"]
  sigma <- fit_t$estimate["s"]
  df <- fit_t$estimate["df"]
  std_res <- (residuals - mu) / sigma
  ks <- ks.test(x = std_res, y = "pt", df = df)
  cat("\nKS test assuming t-distribution:\n")
  print(c(DoF = df, KS_p_value = ks$p.value))
  if(ks$p.value > 0.05) cat("\nResiduals are t-distributed\n") else cat("\nResiduals are NOT t-distributed\n")

}
  check_t(arima_res)
}

print(arma_residual(arima_res))
```

m1 model pass all the tests

## 4.4 ARIMA Forecast

```{r, echo = FALSE, fig.width=12, fig.height=10, out.width="100%",}
fitted_values <- fitted(m1)

plot(as.numeric(d_train), type="l", col="black", lwd=1, 
     main="In-Sample Evaluation: Real vs ARIMA",
     ylab="Spread", xlab="")
lines(as.numeric(na.omit(fitted_values)), col="red")
legend("topright", legend=c("Real data", "Fitted"), col=c("black", "red"), lty=1, lwd=2)
```

**In-Sample Evaluation**
```{r}
residuals <- as.numeric(residuals(m1))

# Core metrics
accuracy(as.numeric(d_train), as.numeric(fitted_values))
# R-squared (will be 0 or near 0 — expected for mean model)
SS_res <- sum(residuals^2, na.rm = TRUE)
SS_tot <- sum((d_train - mean(d_train, na.rm = TRUE))^2, na.rm = TRUE)
R2 <- 1 - SS_res / SS_tot
cat("\nModel R2:", round(R2, 2), "%") 

sign_actual <- sign(as.numeric(d_train))
sign_fitted <- sign(as.numeric(fitted_values))
hits <- (sign_actual == sign_fitted)

da <- mean(hits, na.rm = TRUE) * 100
cat("\nDirectional Accuracy (DA):", round(da, 2), "%")
```

**Out-of-Sample Evaluation**
```{r}
refit <- Arima(d_spread, model = m1)
pred_out <- tail(fitted(refit), n_test)
reali_out <- tail(as.numeric(d_spread), n_test)

accuracy(pred_out, reali_out)

SS_res <- sum((pred_out-reali_out)^2, na.rm = TRUE)
SS_tot <- sum((test - mean(test, na.rm = TRUE))^2, na.rm = TRUE)
R2 <- 1 - SS_res / SS_tot
cat("Model R2:", round(R2, 2), "%") 

sign_actual <- sign(as.numeric(test))
sign_fitted <- sign(as.numeric(pred_out))
hits <- (sign_actual == sign_fitted)
da <- mean(hits, na.rm = TRUE) * 100
cat("DA del modello:", round(da, 2), "%") 
```
If we substitute m1 with m0, we can see that m1 perform better both insample and out-sample

### 4.4.1 Tomorrow forecast
Compute the actual tomoroww forecast, using only this ARMA model
```{r}
forecast_next_week <- forecast(refit, h = 1)

cat("\nForecast for differenced spread tomorrow:\n")
print(forecast_next_week)

# Convert to spread
last_spread <- tail(spread, 1)
fc_diff <- forecast_next_week$mean[1]
fc_spread <- last_spread + fc_diff
fc_spread_lo95 <- last_spread + forecast_next_week$lo[1, 2]
fc_spread_hi95 <- last_spread + forecast_next_week$up[1, 2]

cat("\nForecasted spread for tomorrow:\n")
cat("Point forecast:", fc_spread, "\n")
cat("95% CI:", fc_spread_lo95, "to", fc_spread_hi95, "\n")
```

CONSIDERARE DI FARE UNA CROSS VALIDATION PER UNA ROBA PIU ACCURATA

# 5 Volatility Analysis

We restart from our model and test if the residuals need volatility modelling

```{r}
summary(m1)
```

```{r, echo = FALSE, fig.width=12, fig.height=10, out.width="100%"}
res <- as.numeric(residuals(m1))

par(mfrow = c(3, 1), mar = c(3, 4, 3, 2))
plot(index(d_spread), coredata(d_spread), type = "l", lwd = 0.7,
     main = "Residuals^2 ", xlab = "", ylab = "ds")
abline(h = 0, col = "black", lwd = 0.5)
Acf(as.numeric(res)^2,   lag.max = 30, main = "ACF of res^2")
Pacf(as.numeric(res)^2,  lag.max = 30, main = "PACF of res^2")
par(mfrow = c(1, 1))
```

Check for autocorrelation and ARCH effects on train data
```{r}
DoF <- sum(m1$coef!=0) 
# Ljung-Box on DS^2
lb_arima_res2 <- Box.test(res^2, lag = 16, type = "Ljung-Box", fitdf = DoF)
cat("\nLB test on ARIMA squared residuals (16 lags) p-value: ", round(lb_arima_res2$p.value, 4), "\n")
if (lb_arima_res2$p.value < 0.05) cat("Volatility clustering confirmed\n") else cat("No volatility clustering detected\n")

# ARCH test
arch_test <- ArchTest(as.numeric(res), lags = 16)
cat(sprintf("ARCH Test (16 lags) p-value: %.4f\n", arch_test$p.value))
if (arch_test$p.value < 0.05) cat("ARCH effects confirmed — GARCH modelling needed\n") else cat("No ARCH effects\n")

```

We clearly need a specification model for volatility

## 5.1 Best ARMA-GARCH

We try to find the best ARMA-GARCH fitting different combinations of
models, then decide considering a precise IC, kinda like an auto_arima
but hand-builded.

Create a function that simulates 1 iteration of a for cycle, then
instead of a for cycle use parrallel (mclapply) It will run the function
over $n$ times using $n$ CPU simultanealy.

Can change the critera or modify the parameters, if too much
combinations the code will run for ages.

We will keep the max possible GARCH orders to 1, that's cause vast
majority of the time a GARCH(1,1) is perfect for financial series. We
could add more GARCH model other than `sGARCH`and `eGARCH` but since
those are the most used we keep only them, also other models have
additional parameter than can increase the computational time. We also
capped max ARMA order to 3 to prevent the AIC from picking up noise (it
does, tried to run multiple times and it gives back models that do not
pass diagnostic tests).

The ugarch solvers (`gosolnp`) use random generate starting point, to
have reproducibility of results we need to impose a seed whenever a
optimization is done.

**NOTE**: in `dists` you choose the distribution you suppose the std
residuals will follow, `norm`= normal, `std`= t, `sstd`= skewed t. I
removed `norm`and cause the function would wrongly choose `norm` over
the other 2: you can see if is the distribution fit the residuals in the
printed residual plot or in the "Adjusted Pearson Goodness-of-Fit Test"
at the end of the model output, when `norm` the fit is worse. The
distribution that better fit the residuals is `sstd` (even if not
perfect).

The reason is that `sstd` and `std` have more parameters than `norm`
(number of parameter: `sstd` \> `std` \> `norm`) and since the train
data is "small" this penalize them more than it should in the eye of
AIC/BIC. `norm` will end up with better IC than `sstd` but then you have
the wrong distribution for the residuals and/or fail the diagnostic
tests.

```{r}
start_time <- proc.time() # to see for how much the code run

best_garch <- function(y,
                       ar_lags = 0:3, ma_lags = 0:3,
                       garch_p = 1:1,  garch_q = 1:1,
                       models  = c("sGARCH", "sGARCH-M"),
                       dists   = c("std", "sstd"),
                       criteria = "bic", timeout = 60) {

  combos <- expand.grid(
    ar_lag  = ar_lags,
    ma_lag  = ma_lags,
    garch_p = garch_p,
    garch_q = garch_q,
    model   = models,
    dist    = dists,
    stringsAsFactors = FALSE
  )

  cat("Total combinations to fit:", nrow(combos), "\n")
  num_cores <- detectCores() - 1
  cat("Running on", num_cores, "cores\n")

  fit_one <- function(i) {
    arp   <- combos$ar_lag[i]
    map   <- combos$ma_lag[i]
    gp    <- combos$garch_p[i]
    gq    <- combos$garch_q[i]
    model <- combos$model[i]
    dist  <- combos$dist[i]

    archm <- model == "sGARCH-M"
    true_model <- ifelse(archm, "sGARCH", model)

    spec <- ugarchspec(
      variance.model = list(model = true_model, garchOrder = c(gp, gq)),
      mean.model     = list(armaOrder = c(arp, map), include.mean = TRUE,
                            archm = archm), 
      distribution.model = dist
    )
    fit <- tryCatch({
      withTimeout({
        suppressWarnings(
          ugarchfit(spec = spec, data = y, solver = "hybrid",
          solver.control = list(n.restarts = 5, rseed = 42, trace = 0))
        )
      },
      timeout   = timeout,
      onTimeout = "silent")
    },
    error   = function(e) NULL)
    # Store only not null + right class
    if (!is.null(fit) && inherits(fit, "uGARCHfit") && fit@fit$convergence == 0) {
      ic <- infocriteria(fit)
      return(data.frame(
        ar_p    = arp,
        ma_q    = map,
        garch_p = gp,
        garch_q = gq,
        model   = model,
        dist    = dist,
        aic     = as.numeric(round(ic["Akaike", ] * length(y), 2)),
        bic     = as.numeric(round(ic["Bayes",  ] * length(y), 2)),
        likelihood = as.numeric(round(likelihood(fit), 2)),
        stringsAsFactors = FALSE
      ))
    }
    return(NULL)
  }

  # Gestione Cluster (Windows richiede export esplicito)
  os_name <- Sys.info()["sysname"]
  if (os_name == "Windows") {
    cl <- makeCluster(num_cores)
    clusterSetRNGStream(cl, iseed = 42)
    clusterExport(cl, varlist = c("y", "combos", "fit_one"), envir = environment())
    clusterEvalQ(cl, { library(rugarch); library(R.utils) })
    results_list <- parLapply(cl, seq_len(nrow(combos)), fit_one)
    stopCluster(cl)
  } else {
    results_list <- mclapply(seq_len(nrow(combos)), fit_one, mc.cores = num_cores)
  }
  # Keep only non NULL model
  successful <- Filter(Negate(is.null), results_list)
  if (length(successful) == 0) return(NULL)
  cat("Successful fits:", length(successful), "/", nrow(combos), "\n")
  cat("Models ordered by: ", criteria, "\n")
  results_df <- bind_rows(successful) |> arrange(.data[[criteria]])
  return(results_df)
}

results <- best_garch(as.numeric(d_train))
cat("\nTop models by BIC:\n")
print(head(results))

end_time <- proc.time()
execution_time <- end_time - start_time
cat("\n Total execution time: ", execution_time[3], "seconds")
```
Extract the best model.

```{r}
arma_best_order <- c(results$ar_p[1], results$ma_q[1])
garch_best_order  <- c(results$garch_p[1], results$garch_q[1])
best_dist  <- results$dist[1]
best_mod  <- results$model[1]

cat("\nBest ARMA-GARCH model orders:\nARMA:  ", arma_best_order, "\nGARCH: ", garch_best_order, 
    "\nModel: ", best_mod, "\nDistribution: ", best_dist, "\n") 
garch_spec <- ugarchspec(variance.model = list(model = best_mod, garchOrder = garch_best_order),
                         mean.model = list(armaOrder =arma_best_order, include.mean = TRUE, archm = grepl("-M", best_mod)),
                         distribution.model = best_dist)

set.seed(42)
garch_fit <- ugarchfit(spec = garch_spec, data = d_train, solver = "hybrid", 
solver.control = list(n.restarts = 5, rseed = 42, trace = 0)) # for reproducibility
```
- best model with dtrain(4 anni)(max garch par=3) = ARMA-GARCH(4,1)-(3,3),distribution=std (RANK=AIC)  agg.03_04 
- best model with dtrain(4 anni)(max garch par=1) = ARMA-GARCH(4,2)-(1,1),distribution=sstd (RANK=AIC) agg.03_04  
- best model with dtrain(4 anni)(max garch par=1) = ARMA-GARCH(3,2)-(1,1),distribution=norm (RANK=BIC) agg.03_04
- best model with dtrain(4 anni)(max garch par=1) = ARMA-GARCH-M(2,2)-(1,1),distribution=std (RANK=AIC) agg.10_04
- best model with dtrain(4 anni)(max garch par=1) = ARMA-GARCH-M(0,0)-(1,1),distribution=std (RANK=BIC) agg.10_04
-best model with dtrain(4 anni)(max garch par=5) = ARMA-GARCH(0,0)-(1,1),distribution=std (RANK=BIC) agg.17_04
-best model with dtrain(4 anni)(max garch par=5) = ARMA-GARCH(0,0)-(1,1),distribution=std (RANK=BIC) agg.20_04


## 5.2 Check the GARCH positivity of coefficients
```{r,}

# ============================================================
# GARCH POSITIVITY & STATIONARITY CHECK (rugarch)
# ============================================================
check_garch_positivity <- function(fit, model_name = "GARCH fit") {

  cf    <- fit@fit$coef
  # omega is always named "omega" in rugarch
  omega <- cf["omega"]
  # alpha coefficients: alpha1, alpha2, ...
  alphas <- cf[grepl("^alpha[0-9]+", names(cf))]
  # beta coefficients: beta1, beta2, ...
  betas  <- cf[grepl("^beta[0-9]+",  names(cf))]

  persistence <- sum(alphas) + sum(betas)

  positivity_ok   <- (omega > 0) & all(alphas >= 0) & all(betas >= 0)
  stationarity_ok <- persistence < 1

  cat("\n============================================================\n")
  cat("GARCH POSITIVITY & STATIONARITY CHECK:", model_name, "\n")
  cat("============================================================\n")
  cat(sprintf("  omega  = %+.6f  > 0?  %s\n",
              omega,  ifelse(omega > 0,      "YES ✓", "NO  ✗")))

  for (nm in names(alphas)) {
    cat(sprintf("  %-7s= %+.6f  >= 0? %s\n",
                nm, alphas[nm], ifelse(alphas[nm] >= 0, "YES ✓", "NO  ✗")))
  }
  for (nm in names(betas)) {
    cat(sprintf("  %-7s= %+.6f  >= 0? %s\n",
                nm, betas[nm],  ifelse(betas[nm]  >= 0, "YES ✓", "NO  ✗")))
  }

  cat(sprintf("\n  Persistence (sum α + sum β) = %.6f  < 1? %s\n",
              persistence, ifelse(stationarity_ok, "YES ✓ (covariance stationary)", "NO  ✗ (IGARCH or explosive)")))

  if (positivity_ok & stationarity_ok) {
    uncond_var <- as.numeric(omega) / (1 - persistence)
    cat(sprintf("  Unconditional variance       = %.6f\n", uncond_var))
    cat(sprintf("  Unconditional vol (daily)    = %.4f bps\n", sqrt(uncond_var)))
    cat("  --> All positivity conditions SATISFIED ✓\n")
  } else {
    if (!positivity_ok)   cat("  --> WARNING: Positivity condition VIOLATED ✗\n")
    if (!stationarity_ok) cat("  --> WARNING: Stationarity condition VIOLATED (persistence >= 1) ✗\n")
  }
  cat("============================================================\n\n")

  invisible(list(
    omega         = omega,
    alphas        = alphas,
    betas         = betas,
    persistence   = persistence,
    positivity_ok = positivity_ok,
    stationarity_ok = stationarity_ok
  ))
}

```

```{r, fig.width=12, fig.height=10, out.width="100%",}
# --- Positivity check on the training GARCH fit ---
check_garch_positivity(garch_fit, model_name = "ARMA-GARCH (training fit)")

par(mfrow = c(2, 1))
Acf(garch_fit@fit$z, main = "ACF of Standardized Residuals")
Acf(garch_fit@fit$z^2,lag.max=30,main = "ACF of Squared Std. Residuals")
par(mfrow = c(1, 1))

#Ljung-Box test
Box.test(garch_fit@fit$z^2, lag = 20, type = "Ljung-Box", fitdf = 2)
```

## 5.3 Best model analysis
**Variance parameters**

I coefficienti stimati descrivono la struttura della volatilità e la sua evoluzione temporale:

* **$\omega$ (Omega) = +0.638728**: Rappresenta la costante della varianza. Essendo positiva, assicura un livello minimo di volatilità ("rumore di fondo") sempre presente nel mercato.
* **$\alpha_1$ (Alpha) = +0.195306**: Misura la reattività immediata. Indica che il **18%%** di uno shock inatteso (news, eventi politici) si traduce istantaneamente in nuova volatilità per il giorno successivo.
* **$\beta_1$ (Beta) = +0.775521**: Indica la persistenza. Suggerisce che il **78%** della volatilità odierna viene trascinata per inerzia al giorno seguente, riflettendo una memoria lunga degli shock finanziari.

**Stability diagnostics**

L'affidabilità del modello è confermata dal superamento dei test di stabilità matematica:

* **Persistenza ($\alpha + \beta$) = 0.970827**: Il valore è inferiore a 1, garantendo che il modello sia **stazionario**. Lo spread tende quindi a riassorbire gli shock e a convergere verso la sua media nel lungo periodo.
* **Volatility Clustering**: La prossimità del valore a 1 conferma la tendenza della volatilità a manifestarsi a "blocchi" (periodi di alta tensione seguiti da periodi di calma).

**Long-run indicators**

* **Varianza Incondizionata**: 21.34
* **Volatilità Incondizionata (Daily)**: **4.61 bps**

**Interpretazione**: In condizioni di equilibrio, lo spread oscilla mediamente di circa 4.9 punti base al giorno. Variazioni significativamente superiori indicano una fase di stress che eccede la norma statistica del periodo analizzato.

## 5.4 Diagnostic

```{r,}
show(garch_fit)
infocriteria(garch_fit)*length(d_train)
```

In the model output look at:

-   "Weighted Ljung-Box Test on Standardized Residuals"
-   "Weighted Ljung-Box Test on Standardized Squared Residuals"
-   "Weighted ARCH LM Tests"
-   "Adjusted Pearson Goodness-of-Fit Test" --\> HO: residuals fit the supposed distribution
-   "Sign Bias Test" --\> HO: no sign bias in the standardized residuals
    
TO a visulization of the possible asymmetri effect of news we plot the NIC:

```{r, echo = FALSE, fig.width=12, fig.height=10, out.width="100%"}
# Carica la libreria
library(rugarch)

# 1. Calcola l'impatto (genera una lista)
ni <- newsimpact(garch_fit)

# 2. Estrai x e y correttamente per il plot
# ni$zx è l'asse X (lo shock)
# ni$zy è l'asse Y (la varianza prevista)
plot(ni$zx, ni$zy, 
     type = "l", 
     col = "blue", 
     lwd = 2,
     main = "News Impact Curve",
     xlab = "Shock (Ret_t-1)", 
     ylab = "Conditional Variance (Sigma_t^2)")

# Aggiungi una linea verticale sullo zero per vedere l'asimmetria
abline(v = 0, col = "red", lty = 2)
```

```{r, echo = FALSE, fig.width=12, fig.height=10, out.width="100%"}
par(mfrow = c(2, 1))
plot(garch_fit,which=9)
plot(garch_fit,which=8)
par(mfrow = c(1, 1))
```

**Assymetry test (Engle e Ng)**

From the model output, the Sign Bias Test results are as follows: all p-values are well above the 0.05 threshold, indicating that we fail to reject the null hypothesis of no sign bias in the standardized residuals.
The diagnostic tests confirm that the sGARCH specification with a Student-t distribution is sufficient to capture the volatility dynamics of the BTP-Bund spread. There is no significant evidence that positive and 
negative spread movements generate asymmetric volatility responses, suggesting that an asymmetric specification such as eGARCH is not warranted by the data.

We also test if there's any breakpoint inside the residuals. We use:

-   Bau-Perron test: tests for multiple unknow structural breaks in the
    mean, it find the best breaks location that minimizes the RSS, then
    selects the number of breaks minimizing BIC. The number of breaks
    with lowest BIC is the correct one.
-   CUSUM: tests for parameter instability by tracking the cumulative
    sum of recursive residuals, a deviation of this sum from 0 means
    instability, $H_0:$ stable parameters throught the sample.


```{r, echo = FALSE, fig.width=12, fig.height=10, out.width="100%",}
par(mfrow = c(2, 1))
resid_std <- residuals(garch_fit, standardize = TRUE) |> as.numeric()

bp_resid_df <- data.frame(
  resid = resid_std,
  date  = index(d_train) 
)
bp_resid <- breakpoints(resid ~ 1, data = bp_resid_df, h = 0.15)
summary(bp_resid)
plot(bp_resid)

cusum_resid <- efp(resid ~ 1, data = bp_resid_df, type = "Rec-CUSUM")
plot(cusum_resid)
sctest(cusum_resid)

par(mfrow = c(1, 1))
```

Checking the RSS table we can see that we have the optimal BIC at 0
breaks -\> the residuals do not need breakpoint, adding any break will
make the model worse. Also the RSS remain almost equal with differents
possibles breaks: adding breaks will not explain anything more.

We fail to reject the null in the CUSUM test since the line in plot stay
inside the confidence intervals -\> parameters are stable.

Combined both tests tell us that there's no evidence of structural
breaks in the standardize residuals, any possible change presents in the
raw series is explained by the model.

### 5.4.1 Comment

The estimated model pass all the tests: there's no mean or volatility to
model and the residual fit the supposed distribution (t-skewed) and pass
the tests. Our data can be modelled with this ARMA-GARCH model, next we
run a validation process for its forecasts and collect some metrics. 

# 6 Validation
Now the validation phase, we take the best estimated
ARMA-GARCH and validate through the test data. At every iteration we
refit the model (re-estimate the parameter) but the structure of the
model remain the same: the goal is to calculate how good the model is at
forecasting to then forecast the real future value.

Then we compare the forecast with a Random Walk model to test it our
model is actually useful.

-   Forecast on test data using rolling forecast (window increases on 1
    week every iteration)
-   Refit the model every 10 iteration (10 days) to keep model useful, if every day it sometinme doesn't converge
-   Perform the rolling estimation and forecast
-   If window.size too small the model will fail to converge
-   window.size = 252 -> last year of data

```{r}
start_time <- proc.time()

c1 <- makeCluster(detectCores() - 1)

window_size = 252

set.seed(42)

roll <- ugarchroll(spec = garch_spec, 
                   data = d_spread, 
                   n.ahead = 1, 
                   forecast.length = n_test, 
                   refit.every = 10, 
                   refit.window = "moving", 
                   window.size = window_size, 
                   solver = "hybrid",
                   solver.control = list(n.restarts = 5, rseed = 42, trace = 0),
                   calculate.VaR = TRUE,
                   VaR.alpha = c(0.01, 0.05, 0.95, 0.99), # calculate VaR using the GARCH model
                   cluster = c1)

stopCluster(c1)
# In case of non-convergence problems, re-run with other solvers
if (length(roll@model$nonconverged) > 0) {
  roll <- resume(roll, solver = "gosolnp")
}
if (length(roll@model$nonconverged) > 0) {
  roll <- resume(roll, solver = "nloptr")
}
if (length(roll@model$nonconverged) > 0) {
  roll <- resume(roll, solver = "solnp")
}

# Extract density predictions and realized spread
roll_df <- as.data.frame(roll)

# Extract theoretical quantiles
quants <- as.data.frame(roll, which = "VaR")

# Build the forecast_df for better reading
dfc_df <- data.frame(
  pred_mu  = roll_df$Mu,
  pred_var = roll_df$Sigma^2,
  pred_vol = roll_df$Sigma,
  ci_q05  = quants[, "alpha(5%)"],
  ci_q95  = quants[, "alpha(95%)"],
  ci_q01  = quants[, "alpha(1%)"],
  ci_q99  = quants[, "alpha(99%)"],
  actual   = roll_df$Realized
)
rownames(dfc_df) <- rownames(roll_df)

end_time <- proc.time()
execution_time <- end_time - start_time
cat("\n Total execution time: ", execution_time[3], "seconds\n")
```

===
VERSIONE PER WINDOWS + CORREZIONE (HO MESSO ANCHE I QUANTILI SUPERIORI PER IL CALCOLO DELLA DISTRIBUTION COVERAGE NEL CHUNCK SUCCESSIVO )

**NOTE**: OCCHIO A "ROLL.RDS", SE SI CAMBIA UN QUALCOSA NEL CODICE CHE PRECEDE (TIPO MODEL SPECIFICATION) "RICARICA" LA VECCHIA CACHE CON I VECCHI PARAMETRI
ANDREBBE AGGIUNTO UN CHECK PER VEDERE LA CACHE E COERENTE CON I NUOVI DATI O AGGIORNARE LA CACHE OGNI VOLTA (?)
===

```{r, eval = FALSE, include = FALSE}
start_time <- proc.time()

# ── ugarchroll universale (Windows & Mac) ────────────────────────
if (file.exists("roll.rds")) {
  roll <- readRDS("roll.rds")
  message("✓ roll caricato da cache")
} else {
  # 1. Rileva i core e crea il cluster
  n_cores <- detectCores() - 1
  c1 <- makeCluster(n_cores)
  
  # 2. Carica rugarch su ogni core (necessario su Windows)
  clusterEvalQ(c1, library(rugarch))
  
  on.exit(stopCluster(c1), add = TRUE)
  set.seed(42)
  
  window_size <- 252

  # 3. Esecuzione con tutti i quantili (per evitare l'errore precedente)
  roll <- ugarchroll(garch_spec,
                     data = d_spread,
                     n.ahead = 1,
                     forecast.length = n_test,
                     refit.every = 10,
                     refit.window = "moving",
                     window.size = window_size,
                     solver = "hybrid",
                     solver.control = list(n.restarts = 5, rseed = 42, trace = 0),
                     calculate.VaR = TRUE,
                     VaR.alpha = c(0.01, 0.05, 0.95, 0.99), # Inclusi i quantili alti
                     cluster = c1)
  
  stopCluster(c1)
  gc(verbose = FALSE)
  saveRDS(roll, "roll.rds")
  message("✓ roll salvato in roll.rds")
}
# ────────────────────────────────────────────────────────

# Gestione non-convergenza
if (length(roll@model$nonconverged) > 0) roll <- resume(roll, solver = "nloptr")
if (length(roll@model$nonconverged) > 0) roll <- resume(roll, solver = "solnp")
if (length(roll@model$nonconverged) > 0) roll <- resume(roll, solver = "hybrid")

# Extract density predictions and realized d_spread
roll_df <- as.data.frame(roll)
# Extract theoretical quantiles
quants  <- as.data.frame(roll, which = "VaR")

actual <- as.numeric(quants[, "realized"])
VaR_95 <- as.numeric(quants[, "alpha(5%)"])
VaR_99 <- as.numeric(quants[, "alpha(1%)"])

# --- VaR Test 95% ---
cat("=== VaR Test 95% (Kupiec only) ===\n")
n_obs    <- length(actual)
n_viol_95 <- sum(actual < VaR_95)
expected_95 <- 0.05 * n_obs
# Calcolo manuale LR per coerenza
lr_stat_95 <- -2 * (log(0.05^n_viol_95 * 0.95^(n_obs-n_viol_95)) -
                    log((n_viol_95/n_obs)^n_viol_95 * (1-n_viol_95/n_obs)^(n_obs-n_viol_95)))
p_val_95 <- 1 - pchisq(lr_stat_95, df = 1)

cat("Violazioni osservate:", n_viol_95, "| Attese:", round(expected_95, 1), "\n")
cat("Kupiec p-value:", round(p_val_95, 4), "\n")
cat(if(p_val_95 > 0.05) "Esito: VaR 95% correttamente calibrato\n" else "Esito: VaR 95% NON calibrato\n")

# --- VaR Test 99% ---
cat("\n=== VaR Test 99% (Kupiec only) ===\n")
n_viol_99 <- sum(actual < VaR_99)
expected_99 <- 0.01 * n_obs
lr_stat_99 <- -2 * (log(0.01^n_viol_99 * 0.99^(n_obs-n_viol_99)) -
                    log((n_viol_99/n_obs)^n_viol_99 * (1-n_viol_99/n_obs)^(n_obs-n_viol_99)))
p_val_99 <- 1 - pchisq(lr_stat_99, df = 1)

cat("Violazioni osservate:", n_viol_99, "| Attese:", round(expected_99, 1), "\n")
cat("Kupiec p-value:", round(p_val_99, 4), "\n")
cat(if(p_val_99 > 0.05) "Esito: VaR 99% correttamente calibrato\n" else "Esito: VaR 99% NON calibrato\n")

dfc_df2 <- data.frame(
  pred_mu  = roll_df$Mu,
  pred_var = roll_df$Sigma^2,
  pred_vol = roll_df$Sigma,
  ci_q05  = quants[, "alpha(5%)"],
  ci_q95  = quants[, "alpha(95%)"],
  ci_q01  = quants[, "alpha(1%)"],
  ci_q99  = quants[, "alpha(99%)"],
  actual   = roll_df$Realized
)

rownames(dfc_df2) <- rownames(roll_df)

end_time <- proc.time()
cat("\nTotal execution time:", (end_time - start_time)[3], "seconds\n")
print(tail(dfc_df,10))
print("============")
print(tail(dfc_df2,10))
```

I test di copertura incondizionata (Kupiec Test) mostrano un comportamento divergente del modello ai due livelli di confidenza analizzati.

--> RIFLESSIONE SUI RIOSULTATI DEI VAR

# 6.1 Forecast evaluation
First we convert back to spread to have more readable metrics
```{r}
get_spreads <- function() {
  fc_spread <- numeric(n_test)
  lower_05 <- numeric(n_test)
  upper_95 <- numeric(n_test)
  lower_01 <- numeric(n_test)
  upper_99 <- numeric(n_test)
  
  for (i in 1:n_test) {
    # precedent spread value
    prev_spread <- as.numeric(spread[n_train + i - 1]) 
    
    fc <- dfc_df$pred_mu[i] 
    fc_spread[i] <- prev_spread + fc
    
    ci_lower_05 <- dfc_df$ci_q05[i] 
    ci_upper_95 <- dfc_df$ci_q95[i]
    ci_lower_01 <- dfc_df$ci_q01[i] 
    ci_upper_99 <- dfc_df$ci_q99[i] 
    
    lower_05[i] <- prev_spread + ci_lower_05
    upper_95[i] <- prev_spread + ci_upper_95
    lower_01[i] <- prev_spread + ci_lower_01
    upper_99[i] <- prev_spread + ci_upper_99
  }
  
  act_spread <- as.numeric(spread[(n_train + 1):(n_train + n_test)])
  
  df <- data.frame(
    act_spread = act_spread,
    fc_spread = fc_spread,
    ci_q05 = lower_05,
    ci_q95 = upper_95,
    ci_q01 = lower_01,
    ci_q99 = upper_99
  )
  rownames(df) <- rownames(dfc_df)
  return(df)
}

fc_df <- get_spreads()
cat("Forecast spread vs actual spread:\n")
print(tail(fc_df))
```

Now the evaluation metrics on forecasted spread:
```{r}
errors <- fc_df$act_spread - fc_df$fc_spread

mae  <- mean(abs(as.numeric(errors)), na.rm = TRUE)
mse  <- mean(as.numeric(errors)^2, na.rm = TRUE)
rmse <- sqrt(mse)
mape <- mean(abs(as.numeric(errors) / as.numeric(dfc_df$actual)), na.rm = TRUE)
dir_acc <- mean(sign(as.numeric(dfc_df$actual)) == sign(as.numeric(dfc_df$pred_mu)), na.rm = TRUE)

cat("\nMean evaluation metrics:\n")
cat(sprintf("MAE: %.4f | MSE: %.4f | RMSE: %.4f | MAPE: %.4f | Dir. Acc: %.2f%%\n",
            mae, mse, rmse, mape, dir_acc * 100))

# --- Distribution check  ---
coverage_90 <- mean(dfc_df$actual >= dfc_df$ci_q05 & 
                    dfc_df$actual <= dfc_df$ci_q95, na.rm = TRUE)
coverage_98 <- mean(dfc_df$actual >= dfc_df$ci_q01 & 
                    dfc_df$actual <= dfc_df$ci_q99, na.rm = TRUE)

cat("\nDistribution evaluation, % of forecast that fall into that interval:\n")
cat(sprintf("90%% coverage: %.4f\n", coverage_90))
cat("\nDistribution evaluation, % of forecast that fall into that interval:\n")
cat(sprintf("98%% coverage: %.4f\n", coverage_98))
```

Compare those metrics with a Random Walk model (tomorrow's value = today's value). To see if our model is actually useful.
```{r}
spread_rw <- spread[n_train:(n_train + n_test - 1)] |> as.numeric()

fc_df$spread_rw <- spread_rw 

mae_rw   <- mean(abs(spread_rw - fc_df$act_spread))
rmse_rw  <- sqrt(mean((spread_rw - fc_df$act_spread)^2))
mape_rw  <- mean(abs((spread_rw - fc_df$act_spread) / fc_df$act_spread))
dir_rw   <- mean(sign(diff(spread_rw)) == sign(diff(fc_df$act_spread)))

cat("\nEvalutation metrics, Model vs Random Walk:\n")
cat(sprintf("\nModel  | MAE: %.4f | RMSE: %.4f | MAPE: %.2f%% | Dir: %.2f%%\n",
            mae, rmse, mape*100, dir_acc*100))
cat(sprintf("RW   | MAE: %.4f | RMSE: %.4f | MAPE: %.2f%% | Dir: %.2f%%\n",
            mae_rw, rmse_rw, mape_rw*100, dir_rw*100))
```

The metrics are similar,our model kind of act like a Random Walk. 

Positive note: the accury direction is better.

```{r, echo = FALSE, fig.width=12, fig.height=10, out.width="100%",}
fc_df$date <- as.Date(rownames(fc_df))

p_spreads  <- ggplot(fc_df, aes(x = date)) +
  geom_line(aes(y = act_spread, color = "Actual"), linewidth = 0.8) +
  geom_line(aes(y = fc_spread, color = "Forecasted"), linewidth = 0.8) +
  geom_line(aes(y = spread_rw, color = "Random Walk"), linewidth = 0.8) +
  geom_ribbon(aes(ymin = ci_q05, ymax = ci_q95), fill = "blue", alpha = 0.3) +
  scale_color_manual(values = c("Actual" = "black", "Forecasted" = "blue", "Random Walk" = "red")) +
  labs(title = "Spread: Actual vs Forecasted", x = "", y = "Basis points") +
  scale_x_date(date_breaks = "2 month", date_labels = "%b-%y") +
  scale_y_continuous(breaks = scales::breaks_width(20)) +
  theme_minimal() +
  theme(legend.title = element_blank(), legend.position = "bottom")

print(p_spreads)
```

# 7 Forecast

Forecast future value for next day, use in the same structure of the
validated model.

Use the last `window_size` data (the same as validation phase)

```{r}
data_fit = tail(d_spread, window_size) # size used to validate 
#last_date  <- last(index(data_fit))
last_date <- tail(index(data_fit), 1)

set.seed(42)
future_fit <- ugarchfit(spec = garch_spec, data = data_fit, solver = "hybrid",
solver.control = list(n.restarts = 5, rseed = 42, trace = 0)
)

future_fc  <- ugarchforecast(future_fit, n.ahead = 1)

q_05 <- quantile(future_fc, probs = 0.05)
q_95 <- quantile(future_fc, probs = 0.95)

future_fc <- data.frame(
  d_spread = future_fc  |> fitted()  |> as.numeric(),
  vol = future_fc |> sigma()  |> as.numeric(),
  q_05 = q_05 |> as.numeric(),
  q_95 = q_95 |> as.numeric()
)

spread_conversion <- function(value, q_05, q_95, data){
	last_spread <- as.numeric(tail(data, 1))
	last_date <- data |> index() |> tail(1)
	
	spread <- last_spread + value
	q_05 <- last_spread + q_05
	q_95 <- last_spread + q_95
	
	df <- data.frame(
	 spread = spread,
	 q_05 = q_05,
	 q_95 = q_95)
	
}

future_spread <- spread_conversion(future_fc$d_spread, future_fc$q_05, future_fc$q_95, spread)

cat("\nForecasted values:\n")
cat("\nSpread on last date: ", as.character(last_date), ": ", tail(spread, 1))
cat("\nSpread forecasted for tomorrow:", as.character(last_date + 1), ":\n")
print(future_spread)
```

## 7.1 Simulation Forecast (10,000 paths)

Simulating 10,000 paths accounts for the GARCH shock uncertainty, not
just the ARMA mean.

```{r}
simulation <- ugarchsim(
  future_fit,
  n.sim       = 1,
  m.sim       = 10000,
  startMethod = "sample"
)

mean_sim <- mean(fitted(simulation))
vol_sim  <- mean(sigma(simulation))
q_90     <- quantile(fitted(simulation), probs = c(0.05, 0.95))

cat(sprintf("Mean DS:  %.4f\n", mean_sim))
cat(sprintf("Mean Vol: %.4f\n", vol_sim))
cat(sprintf("Quantiles 5%% | 95%%: %.4f | %.4f\n", q_90[1], q_90[2]))

sim_spread <- spread_conversion(mean_sim, unname(q_90[1]), unname(q_90[2]), spread)
cat("Spread forecast (simulation) for tomorrow:", as.character(last_date + 1), ":\n")
print(sim_spread)
```

Se i risultati di sim_spread sono molto simili a quelli di future_spread
(ottenuti con forecast), significa che il tuo modello è stabile. Se sono
molto diversi, potrebbe esserci molta incertezza o asimmetria nella
distribuzione dei rendimenti.

# MULTIVARIATE MODEL

```{r, echo=FALSE, fig.width=12, fig.height=10, out.width="100%"}
df_plot <- data.frame(date = index(d_train), d_train$spread)
p <- ggplot(df_plot, aes(x = date, y = spread)) +
geom_line() +
geom_hline(yintercept = 0, color = "red") +
labs(title = "Differenced Spread", y ="Basis Points", x = "") +
theme_bw()

print(p)
```

## Multicollinearity Test

We now test for multicollinearity, so now we load the other series, and we make them stationary
```{r}
# ==========================================
# PHASE 1: EXCEL DATA ACQUISITION & ALIGNMENT
# ==========================================

# Step 1: Load required libraries
#install.packages("readxl")
getwd()
library(readxl)
library(xts)
library(zoo)
# 1. Convertiamo il tuo oggetto xts 'train' in un data.frame per il join
df_spread <- data.frame(Data = index(spread), core_data = as.numeric(spread))
colnames(df_spread)[2] <- "Spread"

# 2. Unione progressiva (Left Join) partendo dal TRAIN
# Questo garantisce che terrai solo ed esattamente le date del tuo train attuale
df_multi <- df_spread %>%
  
# left_join(read_excel("vstoxx_24_04.xlsx") %>% dplyr::select(Data = 1, Vstoxx = 2) %>% 
#             mutate(Data = as.Date(substr(as.character(Data), 1, 10))),
#             by = "Data") %>%
  
  left_join(read_excel("eurusd_24_04.xlsx") %>% dplyr::select(Data = 1, EURUSD = 2) %>% 
              mutate(Data = as.Date(substr(as.character(Data), 1, 10))),
              by = "Data") %>%
# left_join(read_excel("FTSEMIB_17_04.xlsx") %>% select(Data = 1, Price_MIB = 2), by = "Data") %>%
# left_join(read_excel("CDSIT10Y_17_04.xlsx") %>% select(Data = 1, CDS10_IT = 2), by = "Data") %>%
  left_join(read_excel("euribor_3M_24_04.xlsx") %>% dplyr::select(Data = 1, EURIBOR3 = 2) %>% 
              mutate(Data = as.Date(substr(as.character(Data), 1, 10))),
              by = "Data") %>%
# left_join(read_excel("EURBANKINDX_17_04.xlsx") %>% select(Data = 1, EUROABANKS = 2), by = "Data") %>%
# left_join(read_excel("ITBANKINDX_17_04.xlsx") %>% select(Data = 1, ITABANKS = 2), by = "Data") %>%
# left_join(read_excel("STOXX® Europe 600 Automobiles & Parts_17_04.xlsx") %>% select(Data = 1,
# Auto_Index = 2), by = "Data")%>%
# Ordinamento cronologico
  arrange(Data) %>%
  
  # 3. Creazione delle variabili trasformate (Rendimenti e Gap)
  mutate(across(-Data, as.numeric)) %>% # Trasforma tutto in numerico tranne la data
#   mutate(
    # Rendimenti logaritmici
#   ret_mib    = c(NA, diff(log(Price_MIB))),
#   ret_auto   = c(NA, diff(log(Auto_Index))),
   #itabanks vs eurobanks
#   bankratio= ITABANKS/ EUROABANKS,
#   ) %>%
    
    dplyr::select(
    Data,
    Spread,
    #Vstoxx,
    EURUSD,
    EURIBOR3
  )

# 2. Gestione NA con Interpolazione Lineare
# na.approx riempie i buchi (es. weekend o festività diverse tra mercati)
# senza eliminare l'intera riga.
df_multi <- df_multi %>%
  arrange(Data) %>%
  mutate(across(-Data, ~ na.approx(. , na.rm = FALSE))) %>%
  # Riempiamo eventuali NA rimasti all'inizio o alla fine con il valore più vicino
  mutate(across(-Data, ~ na.locf(., na.rm = FALSE))) %>%
  mutate(across(-Data, ~ na.locf(., fromLast = TRUE, na.rm = FALSE)))

# Verifica finale
print(paste("Dataset creato con", nrow(df_multi), "osservazioni e", ncol(df_multi), "variabili."))
sum(is.na(df_multi))

# 3. Test di Stazionarietà (ADF Test)
# Creiamo una funzione per testare ogni variabile
check_stationarity <- function(series) {
  test <- adf.test(series, alternative = "stationary")
  return(test$p.value)
}

print("--- Risultati Test ADF (Livelli) ---")
variabili <- names(df_multi)[names(df_multi) != "Data"]
stazionarieta_risultati <- sapply(df_multi[variabili], check_stationarity)
print(stazionarieta_risultati)

# 4. Trasformazione in Differenze Prime (solo dove necessario)
# Se p-value > 0.05, la serie non è stazionaria e va differenziata
df_final_staz <- df_multi


for (var in variabili) {
  if (stazionarieta_risultati[var] > 0.01) {
    message(paste("Variabile", var, "non stazionaria. Calcolo differenze prime..."))
    # Sovrascriviamo con la differenza o creiamo nuova colonna
    # Qui creiamo 'd_NomeVariabile' per chiarezza nel Granger
    df_final_staz[[paste0("d_", var)]] <- c(NA, diff(df_multi[[var]]))
  } else {
    message(paste("Variabile", var, "già stazionaria."))
    df_final_staz[[paste0("d_", var)]] <- df_multi[[var]]
  }
}

# Puliamo i nuovi NA generati dal comando diff()
df_final_staz <- df_final_staz[-1, ]
sum(is.na(df_final_staz))
df_final_staz<- df_final_staz %>%
  dplyr::select(Data, starts_with("d_"))

# 3. Create the Training set (First 80%)
n_rows <- nrow(df_final_staz)
train_var <- df_final_staz %>% slice(1:n_train)

# 4. Create the Testing set (Remaining 20%)
test_var <- df_final_staz %>% slice((n_train + 1):n_rows)
# Check the results
cat("Total observations:", n_rows, "\n")
cat("Training set rows:", nrow(train_var), "(from", min(train_var$Data), "to", max(train_var$Data), ")\n")
cat("Testing set rows:", nrow(test_var), "(from", min(test_var$Data), "to", max(test_var$Data))
  
```

Now we should have a matrix with the d_spread value and the vstoxx index value, with matching data.
We also have the train and test matrices.
```{r}
# 5. Granger Test con stampa estesa
target <- "d_Spread"
colonne_test <- names(train_var)[sapply(train_var, is.numeric)]
colonne_test <- colonne_test[colonne_test != target]

risultati_granger <- data.frame(Variabile = character(), P_Value = numeric())

cat("\n--- DETTAGLI TEST DI GRANGER (Lag = 5) ---\n")

for (var in colonne_test) {
  # Eseguiamo il test
  g_test <- grangertest(as.formula(paste(target, "~", var)), order = 5, data = train_var)
  
# Estraiamo il p-value
  p_val <- g_test$`Pr(>F)`[2]
  
  # Salviamo i risultati
  risultati_granger <- rbind(risultati_granger, data.frame(Variabile = var, P_Value = p_val))
  
  # Stampiamo l'esito a video con un po' di formattazione
  cat(paste0("\nTest: ", var, " causality on ", target, "\n"))
  print(g_test)
  cat("--------------------------------------------\n")
}

# 6. Analisi CCF: Lag Predittivi e Intervalli di Confidenza
target <- "d_Spread"
variabili_ccf <- names(train_var)[sapply(train_var, is.numeric)]
variabili_ccf <- variabili_ccf [colonne_test != target]
# Calcolo del numero di osservazioni e della soglia IC (95%)
n_obs <- nrow(train_var)
soglia_ic <- 1.96 / sqrt(n_obs) 

cat("\n========================================================\n")
cat("   ANALISI DETTAGLIATA LAG PREDITTIVI (X precede Spread)   \n")
cat("========================================================\n")
cat(paste("Numero osservazioni:", n_obs, "\n"))
cat(paste("Soglia IC (95%): +/-", round(soglia_ic, 4), "\n"))
cat("Nota: Un lag k < 0 indica che X al tempo t-|k| correla con Spread al tempo t\n")

# Tabella riassuntiva per i risultati
final_ccf_summary <- data.frame(Variabile = character(), 
                                Lag = numeric(), 
                                Correlazione = numeric())

# Setup layout: 1 grafico per riga (o puoi fare par(mfrow=c(2,2)) fuori dal ciclo)
# Setup layout: 3 righe e 2 colonne
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

for (var in variabili_ccf) {
  # 1. Calcolo della CCF
  ccf_result <- ccf(train_var[[var]], train_var[[target]], lag.max = 10, plot = FALSE)
  # 3. Generazione del PLOT
  # Usiamo il risultato già calcolato per fare il plot
  plot(ccf_result, 
       main = paste(var, "vs Spread"),
       ylab = "Corr", xlab = "Lag",
       ci.col = "blue")
  # Estrazione dati
  lags <- ccf_result$lag[,1,1]
  corrs <- ccf_result$acf[,1,1]
  
  # Filtriamo: solo lag < 0 (X predice Spread)
  idx_p <- which(lags < 0)
  lags_p <- lags[idx_p]
  corrs_p <- corrs[idx_p]
  
  # Verifichiamo quali superano l'Intervallo di Confidenza
  sig_idx <- which(abs(corrs_p) >= soglia_ic)
  
  if (length(sig_idx) > 0) {
    cat(paste0("\n>>> Variabile: ", var, "\n"))
    
    # 2. Creazione della tabella dei lag significativi
    df_temp <- data.frame(
      Lag = lags_p[sig_idx],
      Correlazione = round(corrs_p[sig_idx], 4)
    ) %>% arrange(Lag)
    
    print(df_temp)
    
    # Salvataggio nel riassunto globale
    df_temp$Variabile <- var
    final_ccf_summary <- rbind(final_ccf_summary, df_temp)
    
  } else {
    cat(paste0("\nVariabile: ", var, " - Nessun lag predittivo significativo.\n"))
  }
}

# 8. Selezione finale delle variabili basata su CCF
# Teniamo solo le variabili che hanno almeno un lag predittivo (escluso lo 0 se cerchi solo causalità)
variabili_da_usare_ccf <- unique(final_ccf_summary$Variabile)

cat("\n--- Selezione Finale basata su CCF ---\n")
print(variabili_da_usare_ccf)

```

## MQ Test
We have both series stationary, we now run the qm test to see if we need a VAR model.

```{r}

library(MTS)
library(vars)

# MQ test
mq(train_var[, -1], lag = 10)

# We have all lags in which p-value is less than 0.05 ( equal to 0), so we can estimate a VAR model.
```
## VAR Model Estimation
```{r}
library(vars)
# 1. Determine optimal VAR lag length
lag_selection <- VARselect(train_var[, -1], lag.max = 10, type = "const")
lag_selection

# We have optimal lag = 1 chosen by all the models

# Fit the model
var_bic <-vars::VAR(train_var[, -1], p = 2, type = "const")
summary(var_bic)
var_bic_mts <- MTS::VAR(train_var[, -1], p = 2)#equivalent with MTs package

var_aic<-vars::VAR(train_var[, -1], p = 3, type = "const")
summary(var_aic)
var_aic_mts <- MTS::VAR(train_var[, -1], p = 3)#equivalent with MTS package

```
BIC suggests 2 lags. #All coefficients ar significant except the one of spread itself

```{r}
var_restricted_bic = vars::restrict(var_bic, method = "ser", thresh = 1.64)
summary(var_restricted_bic)
var_ref_bic<- refVAR(var_bic_mts, thres = 1.64)
var_restricted_aic = vars::restrict(var_aic, method = "ser", thresh = 1.64)
summary(var_restricted_aic)
var_ref_aic <- refVAR(var_aic_mts, thres = 1.64)

```

```{r}
count_coeff = function(model){
  names = names(model$varresult)
  n = c()
  for (name in names){
    n = c(n, length(model$varresult[[name]]$coefficients))
  }
  sum(n)
}

cat("\nUnrestricted model bic coeff: ", count_coeff(var_bic),
    " Restricted model bic coeff: ", count_coeff(var_restricted_bic), "\n")
cat("\nUr model SC: ", BIC(var_bic), "R model SC: ", BIC(var_bic))
cat("\nUr model AIC: ", AIC(var_bic), "R model AIC: ", AIC(var_bic), "\n")


cat("\nUnrestricted model aic coeff: ", count_coeff(var_aic),
    " Restricted model aic coeff: ", count_coeff(var_restricted_aic), "\n")
cat("\nUr model SC: ", BIC(var_aic), "R model SC: ", BIC(var_aic))
cat("\nUr model AIC: ", AIC(var_aic), "R model AIC: ", AIC(var_aic))
## VAR Diagnostics
```


```{r}

# ==========================================
# PHASE 4: VAR DIAGNOSTICS
# ==========================================

# 1. Test for leftover Serial Correlation (Portmanteau Test) # equivalente dell'MQ per pacchetto vars
# We test up to 10 lags to be safe. We want p-value > 0.05
serial_test <- serial.test(var_bic, lags.pt = 10, type = "PT.asymptotic")
cat("\n--- Serial Correlation Test ---\n")
print(serial_test)

cat("\n--- MQ test ---\n")
# Calcoliamo i residui del modello raffinato
res_refined <- var_ref_bic$residuals
# Eseguiamo il test MQ fino al lag 10
mq_test <- mq(res_refined, lag = 10,adj=TRUE)


# 2. Test for ARCH Effects (Volatility Clustering)
# We want p-value < 0.05 to justify using DCC-GARCH next
arch_test <- arch.test(var_bic, lags.multi = 5, multivariate.only = TRUE)
cat("\n--- ARCH Effects Test ---\n")
print(arch_test)

## We need the DCC-GARCH model, as there are arch effects (p-value is near 0)

```
Il modello con lag come suggerito dal bic non passa la diagnostica dì--> autocorrelazione residua

```{r}
# ==========================================
# PHASE 4: VAR DIAGNOSTICS
# ==========================================

# 1. Test for leftover Serial Correlation (Portmanteau Test) # equivalente dell'MQ per pacchetto vars
# We test up to 10 lags to be safe. We want p-value > 0.05
serial_test <- serial.test(var_aic, lags.pt = 10, type = "PT.asymptotic")
cat("\n--- Serial Correlation Test ---\n")
print(serial_test)

cat("\n--- MQ test ---\n")
# Calcoliamo i residui del modello raffinato
res_refined <- var_ref_aic$residuals
# Eseguiamo il test MQ fino al lag 10
mq_test <- mq(res_refined, lag = 10,adj=TRUE)

# 2. Test for ARCH Effects (Volatility Clustering)
# We want p-value < 0.05 to justify using DCC-GARCH next
arch_test <- arch.test(var_aic, lags.multi = 5, multivariate.only = TRUE)
cat("\n--- ARCH Effects Test ---\n")
print(arch_test)

## We need the DCC-GARCH model, as there are arch effects (p-value is near 0)

```
### Stability test
We run CUSUM test to check for possible structural breaks
```{r, echo=FALSE, fig.width=12, fig.height=10, out.width="100%"}
vars::stability(var_restricted_aic)
plot(vars::stability(var_restricted_aic))
```

The `euribor` series residuals drifts outside the confidence intervals, sign that there's a structural break.
We could try to solve it using a time dummy, but first we need to understand where the break occurs
```{r}
bp_cds = breakpoints(residuals(var_restricted_aic)[,"d_EURIBOR3"] ~ 1)
summary(bp_cds)
```

```{r, echo=FALSE, fig.width=12, fig.height=10, out.width="100%"}
bp <- train_var$Data[c(302, 577, 832)]
ggplot(train_var, aes(x=Data, y=d_EURIBOR3)) +
geom_line() +
geom_vline(xintercept = bp, col = "red", linetype = "dashed",
           linewidth = 1.2) +
labs(title = paste("d_Euribor with break point at ", bp)) +
theme_bw()
```

The RSS/BIC table tells that the model is at his best with 3 breakpoints. From the plot we can also see thhem
Non so come si risolve questa cosa per cui rimandiamo

#### Normality
Check if residuals are normal distributed, if not try to fit them to a t-distribution (more likely).

Use a KS test to test wheter they follow a t-distribution $H_0: \text{serie follows supposed distribution}$
```{r, warning=FALSE}
# p_value < 0.05 -> non-normal distributed
normality.test(var_restricted_aic, multivariate.only = FALSE)

check_dist <- function(residual) {
  t = MASS::fitdistr(residuals, densfun = "t")
  n = MASS::fitdistr(residuals, densfun = "normal")
  aic = c(AIC(t), AIC(n))
  bic = c(BIC(t), BIC(n))
  df = data.frame(aic, bic, row.names = c("t", "n"))
  df
}
apply(residuals(var_restricted_aic), 2, check_dist)
```

We reject the null of the JB test for all the univariate series and the joint multivariate, thus the residuals do not follow a normal distribution.

The subsequent test, consisting in estimating a normal a t distribution on the residuals and then compute their IC, showed that a t-distribution 
is a better fit than the normal.

#### Causality
```{r}
vars::causality(var_restricted_aic, cause = c("d_EURUSD", "d_EURIBOR3"))
vars::causality(var_restricted_aic, cause = c("d_Spread", "d_EURIBOR3"))
vars::causality(var_restricted_aic, cause = c("d_Spread", "d_EURUSD"))

```

## DCC-GARCH Model

```{r}
# ==========================================
# PHASE 5: THE DCC-GARCH MODEL
# ==========================================
library(rugarch)
library(rmgarch)

# 1. Extract the residuals from your fitted VAR model
# The scale (* 100) is crucial! GARCH optimizers fail if the numbers are too tiny (e.g. 0.0001)
res_var <- resid(var_restricted_aic)*100

# Convert the residuals back into a Time Series (xts) object for plotting later
# We use the dates from diff_data, but remember the VAR loses the first 'p' rows
# So we align the dates by tailing the diff_data
# Assicurati che la colonna date sia nel formato corretto
dates_vector <- as.Date(train_var$Data)

# Crea l'oggetto xts usando le ultime 'n' date corrispondenti ai residui
var_residuals_xts <- xts(res_var, 
                         order.by = tail(dates_vector, nrow(res_var)))
colnames(var_residuals_xts) <- c("Spread_Shock", "EURUSD_Shock","EURIBOR3_Shock")

# 2. Specify the Univariate GARCH Models
# Because the VAR already handled the mean, we set the ARMA order to (0,0) with NO mean!
uspec <- ugarchspec(
  mean.model = list(armaOrder = c(0, 0), include.mean = FALSE),
  variance.model = list(garchOrder = c(1, 1), model = "sGARCH"),
  distribution.model = "norm" # You can change to "std" (Student-t) if it fails to converge
)

# Replicate this specification for all of our variables
mspec <- multispec(replicate(3, uspec))

# 3. Specify the Multivariate DCC-GARCH model
# We set VAR = FALSE because we ALREADY did the VAR manually and are feeding it the residuals!
dcc_spec <- dccspec(
  uspec = mspec, 
  VAR = FALSE,
  dccOrder = c(1, 1), 
  distribution = "mvnorm"
)

# 4. Fit the DCC-GARCH model!
# We feed it our scaled VAR residuals
cat("\nFitting DCC-GARCH...\n")
fit_dcc <- dccfit(dcc_spec, data = var_residuals_xts)

# Print the final estimated model
show(fit_dcc)

# Dynamic conditional correlations over time
dcc_corr <- rcor(fit_dcc)        # array: 2x2xT
rho_t    <- dcc_corr[1,2,]       # time series of rho(BTP-Bund, VSTOXX)
plot(rho_t, type = "l", main = "Dynamic Conditional Correlation")

# Conditional volatilities
vol <- sigma(fit_dcc)

# DCC parameters (alpha + beta, should sum < 1)
coef(fit_dcc)

```
```{r}
library(xts)

rho_t2 <- rcor(fit_dcc)[1, 2, ]
resids <- residuals(var_restricted_aic)

plot(zoo(rho_t2, order.by = index(resids)), type = "l",
     main = "Dynamic Conditional Correlation: BTP-Bund vs VSTOXX",
     ylab = "Correlation", xlab = "")
abline(h = mean(rho_t2), col = "red", lty = 2)
```

## Model Diagnostics
```{r}
# 1. Model Diagnostics 
# Standardized residuals should have no ARCH effects remaining
infocriteria(fit_dcc)   # AIC, BIC
plot(fit_dcc, which=2)           # rmgarch gives several diagnostic plots
plot(fit_dcc, which=3)

#colnames(var_fit$y)

# 2. VAR Impulse Response Functions (IRFs)
# This answers "how does the BTP-Bund spread respond to a VSTOXX shock?"
library(vars)
irf_result <- irf(var_restricted_aic, impulse = "d_EURUSD",
                  response = c("d_Spread", "d_EURIBOR3"),
                  n.ahead = 20, boot = TRUE)
plot(irf_result)

irf_result <- irf(var_restricted_aic, impulse = "d_EURIBOR3",
                  response = c("d_Spread", "d_EURUSD"),
                  n.ahead = 20, boot = TRUE)
plot(irf_result)

# 3. DCC-GARCH Volatility Forecast
# A short-horizon forecast of conditional variances and correlations
dcc_forecast <- dccforecast(fit_dcc, n.ahead = 20)
plot(dcc_forecast, which = 3)

# Extract forecasted correlation
rcor(dcc_forecast)

# 4. Variance Decomposition (FEVD)
# Tells you how much of the BTP-Bund spread variance is explained by VSTOXX shocks

fevd_result <- fevd(var_restricted_aic, n.ahead = 20)
plot(fevd_result)
```

## Forecast
```{r}

# Re-fit with out.sample to enable rolling forecast
dcc_fit2 <- dccfit(dcc_spec, data = resids,
                   out.sample = 100)   # hold out last 100 obs

# 1. Unconditional n-ahead forecast
dcc_fore <- dccforecast(dcc_fit2, n.ahead = 20)

# Forecasted correlation
rcor(dcc_fore)   # 2x2xT array

# Forecasted volatilities
sigma(dcc_fore)

# 2. Rolling 1-step-ahead forecast (more meaningful for evaluation)
dcc_roll <- dccforecast(dcc_fit2, n.ahead = 1, n.roll = 99)
rcor(dcc_roll)

# 1-Step Ahead Forecast with VAR
library(vars)

# Forecast 1 step ahead (next trading day)
var_forecast <- predict(var_restricted_aic, n.ahead = 1, ci = 0.95)

# Print the forecast
print(var_forecast)
last_spread <- tail(spread, 1)
forecast_level <- last_spread + var_forecast$fcst$d_Spread[1, "fcst"]
forecast_level

```
## Test the model
We now use the test portion of our data, to see the accuracy of our training model

```{r}

library(vars)
library(xts)
library(ggplot2)

# =========================
# Test set: rolling 1-step-ahead out-of-sample forecast
# =========================

train_multi <- train_var[,-1]
test_multi  <- test_var[,-1]
sum(is.na(train_multi))
# 1. Selezione lag ottimale sul train
lag_sel <- VARselect(train_multi, lag.max = 10, type = "const")
p_opt <- as.numeric(lag_sel$selection["AIC(n)"])

cat("Optimal lag selected on train:", p_opt, "\n")

# 2. Rolling 1-step-ahead forecast sulla serie differenziata "spread"
n_test <- nrow(test_multi)

fc_dspread <- numeric(n_test)
real_dspread <- as.numeric(test_multi[, "d_Spread"])

# Among the candidate rolling window sizes, 400 observations produced the lowest out-of-sample forecast error and  # the most plausible final spread estimate, therefore it was selected as the preferred specification.
window_size <- 400


for (i in 1:n_test) {

  if (i == 1) {
    available_data <- train_multi
  } else {
    available_data <- rbind(train_multi, test_multi[1:(i - 1), ])
  }

  if (nrow(available_data) > window_size) {
    current_train <- available_data[(nrow(available_data) - window_size + 1):nrow(available_data), ]
  } else {
    current_train <- available_data
  }

  var_roll <- VAR(current_train, p = p_opt, type = "const")
  pred_1 <- predict(var_roll, n.ahead = 1, ci = 0.95)

  fc_dspread[i] <- as.numeric(pred_1$fcst$d_Spread[1, "fcst"])
}


# 3. Date del test
test_dates <-test_var[,1]

# 4. Ricostruzione dello spread in livelli
# ultimo valore osservato in livelli prima dell'inizio del test
test_start_date <- test_dates[1]
pos_test_start <- which(index(spread) == test_start_date)
last_level_before_test <- as.numeric(spread[pos_test_start - 1])

# forecast in livelli
fc_level <- last_level_before_test + cumsum(fc_dspread)

# valori reali in livelli
real_level <- as.numeric(spread[test_dates])

# 5. Data frame finale
df_roll <- data.frame(
  Date = as.Date(test_dates),
  RealSpread = real_level,
  ForecastSpread = fc_level,
  RealDiff = real_dspread,
  ForecastDiff = fc_dspread
)

# 6. Grafico livelli
ggplot(df_roll, aes(x = Date)) +
  geom_line(aes(y = RealSpread, color = "Real spread"), linewidth = 1) +
  geom_line(aes(y = ForecastSpread, color = "Rolling forecast"),
            linewidth = 1, linetype = "dashed") +
  labs(
    title = "Rolling 1-Step-Ahead VAR Forecast vs Real Spread",
    y = "Spread (bps)",
    x = "Date",
    color = "Legend"
  ) +
  theme_bw()

# 7. Metriche sui livelli
err_level <- df_roll$RealSpread - df_roll$ForecastSpread

mae_level <- mean(abs(err_level), na.rm = TRUE)
rmse_level <- sqrt(mean(err_level^2, na.rm = TRUE))

cat("MAE levels:", round(mae_level, 4), "\n")
cat("RMSE levels:", round(rmse_level, 4), "\n")

# 8. Benchmark naive
naive_fc_level <- rep(last_level_before_test, length(fc_level))

mae_naive <- mean(abs(real_level - naive_fc_level), na.rm = TRUE)
rmse_naive <- sqrt(mean((real_level - naive_fc_level)^2, na.rm = TRUE))

cat("Naive MAE:", round(mae_naive, 4), "\n")
cat("Naive RMSE:", round(rmse_naive, 4), "\n")

# 9. Valore finale out-of-sample 1-step-ahead
final_forecast_date <- tail(test_dates, 1)
final_forecast_diff <- tail(fc_dspread, 1)
final_forecast_level <- tail(fc_level, 1)

cat("Final forecast date:", as.character(final_forecast_date), "\n")
cat("Final out-of-sample 1-step forecast (diff):", round(final_forecast_diff, 4), "\n")
cat("Final out-of-sample 1-step forecast (level):", round(final_forecast_level, 4), "\n")





```
Window ampia (es. 300): Il modello vede molta storia. I coefficienti sono più stabili perché mediati su un lungo periodo, ma il modello è "lento" a reagire a cambiamenti strutturali recenti (es. un improvviso aumento della volatilità o un cambio di politica monetaria).

Window piccola (es. 150): Il modello è molto reattivo. Si adatta rapidamente agli shock più recenti, ma i coefficienti sono più "nervosi" (instabili) perché basati su pochi dati. Questo può portare a previsioni molto diverse anche per piccoli cambiamenti nei dati.

SPiegare comportamento rol alto mae basso e viceversa


