---
title: "Spread"
output: html_document
date: "2026-03-19"
editor_options: 
  markdown: 
    wrap: 72
---

# SPREAD

## 1. Data setup

```{r setup, include=FALSE}
library(urca)
library(tseries)
library(quantmod)
library(forecast)
library(rugarch)
library(FinTS)
library(ggplot2)
library(dplyr)
library(httpgd)
library(R.utils)
library(xts)
library(readxl)
library(fBasics)
library(lmtest)
library(sandwich)
library(fUnitRoots)
library(stats)
library(graphics)
library(MTS)
library(MASS)
library(zoo)
library(forecast)
library(seastests)
library(strucchange)
```

Load the data and parse from excel file, weekly spread.

```{r, include=FALSE}
rm(list= ls())
getwd()
#setwd("~/Desktop")
data <- read_excel("Spread-weekly.xlsx")
data <- na.omit(data)
# Cambia segno solo se la media è negativa
if(mean(data$`IT10DE10=TWEB (BID)`, na.rm=TRUE) < 0) {
  data$`IT10DE10=TWEB (BID)` <- data$`IT10DE10=TWEB (BID)` * -1}
data$Date <- as.Date(data$Date)
spread <- zoo(data$`IT10DE10=TWEB (BID)`, order.by = data$Date)
```

Inspect data

```{r}
head(data)
str(data)
sum(is.na(data))
any(is.infinite(spread))
```

Plot the data

```{r,echo=FALSE}
plot(spread, main="BTP-BUND Spread (Weekly", ylab="Spread (bps)", xlab= "Date")
```

```{r, echo=FALSE}
options(scipen = 999)
basicStats(spread)
```

The descriptive analysis of the BTP-BUND weekly spread reveals a
significant **positive skewness of 1.53**, indicating a pronounced
right-skewed distribution. This coefficient suggests that while the
spread remains relatively contained for the majority of the sample
period—centered around a **median of 158 bps**—it is susceptible to
infrequent but extreme upward shocks. These outliers, culminating in a
**maximum value of 532 bps**, exert an upward pressure on the
distribution, pulling the **mean (184 bps)** notably above the median.
Econometrically, this asymmetry highlights the 'tail risk'
characteristic of sovereign spreads, where periods of relative market
calm are occasionally interrupted by severe episodes of financial or
political stress.

**Forecast n°1 : naive forecast (mean)**

```{r, include=FALSE}
mean <- mean(data$`IT10DE10=TWEB (BID)`, na.rm=TRUE)
```

```{r, echo=FALSE}
plot(spread, main="BTP-BUND Spread (Weekly", ylab="Spread (bps)", xlab= "Date")
abline(h = mean, col="red", lwd=2, lty=2)
text(x = start(spread), y= mean,
     labels=paste("Mean =", round(mean, 2)),
     col="red", adj= c(0, -0.5))
```

## 2. Test for seasonality

We use seastests package

```{r}
spread_ts <- ts(data$`IT10DE10=TWEB (BID)`, start=c(2016, 1), frequency = 52)
isSeasonal(spread_ts, test="combined")
#We get FALSE

#Individual tests
isSeasonal(spread_ts, test="qs")
isSeasonal(spread_ts, test="seasdum")
isSeasonal(spread_ts, test="kw")
isSeasonal(spread_ts, test="fried")

#We get all FALSE

## Now plot ACF and PACF
Acf(spread_ts, lag.max= 52)
Pacf(spread_ts, lag.max = 52)

#We have 1 significant lag at 1 in the Pacf, and a Acf which gradually goes to 0
#This suggests no seasonality.

#Last point, let's fit an arima model

adf.test(spread_ts)     # stationarity test -> we have p-value = 0.546, so we fail to reject the null, so we are in the null workspace, this means no seasonality.

kpss.test(spread_ts) # stationarity test where the null hypothesis is that the series IS stationary
# p-value is 0.01 -> we REJECT the null, so the series is NOT stationary

## Finally, we can affirm that our data does not contain seasonality, so we can work using normal ARIMA models and ARIMA-GARCH models.
```

## Visualization for the no-seasonality

```{r seasonal_analysis, echo=FALSE}
# 1.Plot
ggseasonplot(spread_ts, year.labels = FALSE) +
  scale_x_continuous(breaks = seq(1, 52, by = 4)) + 
  theme_minimal() +
  labs(title = "Seasonal effect BTP-BUND",
       subtitle = "Weekly data (2016-2026)",
       x = " ",
       y = "Spread (bps)",
       color = "Year")
# 2. Subseries Plot (per vedere la media di ogni settimana nel tempo)
ggsubseriesplot(spread_ts) +
  scale_x_continuous(breaks = seq(1, 52, by = 4)) +
  theme_minimal() +
  labs(title = "Subseries Plot: Medie Storiche per Settimana",
       x = "Settimana",
       y = "Spread (bps)")

# 3.STL
decomposed <- stl(spread_ts, s.window = "periodic")
autoplot(decomposed) +
  theme_minimal() +
  labs(title = "Decomposizione STL della Serie Storica",
       subtitle = "Scomposizione in Trend, Stagionalità e Residui")
```

Visual inspection of the seasonal plots confirms the results of the
formal tests (Friedman, Kruskal-Wallis). The **Subseries Plot** shows
remarkably flat weekly means, while the **STL Decomposition** reveals
that the magnitude of the seasonal component (approx. 15 bps) is
negligible compared to the overall trend and volatility of the series
(over 200 bps). This evidence justifies the use of a non-seasonal
ARIMA-GARCH framework, as the spread dynamics are driven by stochastic
shocks and structural trends rather than deterministic calendar effects.

## 3. Model Specification: Levels vs. Logarithmic Transformations\*\*

A critical decision in modelling the BTP-BUND spread involves the choice
between raw levels (basis points) and logarithmic transformations. While
log-returns are standard for equity prices to ensure normality and
constant variance, this study follows the established econometric
literature—notably **Campbell, Lo, and MacKinlay (1997)** and **Hamilton
(1994)**—by analyzing the spread in **levels**.

As argued by **Engle (1982)** in his seminal work on ARCH models,
interest rate differentials are already expressed in percentage terms;
therefore, a log-transformation is not only redundant but can lead to
significant numerical instability as values approach the **Zero Lower
Bound (ZLB)**. Maintaining the analysis in basis points ensures the
direct economic interpretability of the results, which is essential for
assessing credit and liquidity risk premiums.

#### Literature Review on Model Specification

The choice to model the BTP-BUND spread in first differences of its
levels is supported by the foundational empirical work on Eurozone
sovereign risk. **Codogno, Favero, and Missale (2003)** and **Favero et
al. (2010)** demonstrate that interest rate differentials are best
captured in basis points to maintain linear properties between credit
risk and liquidity factors. Furthermore, the application of a **GARCH
framework**, as pioneered by **Bollerslev (1986)**, ensures that the
time-varying volatility of the spread is explicitly parameterized rather
than suppressed via logarithmic transformations.

### 3.1 Data Partitioning: Training and Test Sets

To evaluate the predictive performance of the **ARIMA model**, the
dataset is partitioned into training and testing subsets. The training
phase utilizes a **rolling window approach** with a fixed length of 10
years, corresponding to **520 weekly observations**.

This specific configuration ensures that the model is estimated on a
sufficiently long historical period while remaining sensitive to more
recent structural dynamics in the BTP-BUND spread. The architecture of
the script is designed for **parametric flexibility**: the training
window duration can be easily modified by adjusting the
`$n_train_window$` constant (e.g., replacing 520 with a different value)
to test the model's robustness across different look-back periods.

```{r}
n_test <- 52
n_train_window <- 260 # 10 anni di dati weekly (52 * 10)

# Lunghezza totale della serie differenziata
total_obs <- length(spread)

# 1. Definizione degli indici per il TEST (le ultime 10)
test_start <- total_obs - n_test + 1
test_end   <- total_obs
test       <- spread[test_start:test_end]

# 2. Definizione degli indici per il TRAIN (i 10 anni precedenti al test)
# Il train deve finire esattamente dove inizia il test
train_end   <- test_start - 1
train_start <- train_end - n_train_window + 1

# Estrazione del train
train<- spread[train_start:train_end]

# Verifica coerenza
print(paste("Lunghezza Train:", length(train))) # Deve essere 520
print(paste("Lunghezza Test:", length(test)))       # Deve essere 10
# Creiamo un piccolo data frame per il plot
df_plot <- data.frame(
  Settimana = 1:length(spread),
  Valore = as.numeric(spread)
)
ggplot(df_plot, aes(x = Settimana, y = Valore)) +geom_line(color = "steelblue") +
geom_vline(xintercept = length(train), color = "red", linetype = "dashed",
linewidth = 1) +labs(title = "BTP-BUND Spread: Training vs Test Set",subtitle = paste("Punto di taglio all'osservazione:", length(train)),
    x = "Settimane",y = "Variazione Spread (bps)")+theme_minimal()

cat(sprintf("Training observations: %d\n", length(train)))
cat(sprintf("Test observations:     %d\n", length(test)))
cat(sprintf("Train period: %s -> %s\n", index(train)[1], index(train)[length(train)]))
cat(sprintf("Test period:  %s -> %s\n", index(test)[1],  index(test)[length(test)]))
```

### 3.3 Stationarity and Unit Root Testing

To ensure the statistical validity of the ARIMA-GARCH framework, we
first evaluate the integration order of the series. Financial variables
often exhibit non-stationary behavior, characterized by a "random walk".

#### 3.3.1 Augmented Dickey-Fuller (ADF) Test

We perform the ADF test to check for the presence of a unit root
($H_0$).

```{r}
par(mfrow = c(1, 2))
acf(as.numeric(train))
Pacf(as.numeric(train))
par(mfrow = c(1, 1)) 
```

### 3.4 Persistence and Autocorrelation Analysis

The initial inspection of the **Autocorrelation Function (ACF)** reveals
a very slow, near-linear decay. In time series econometrics, this
pattern is a classic diagnostic for a **"near-unit root"** process,
indicating that the AR component is highly persistent and the series is
likely non-stationary.

Regarding the optimal number of lags for the model, while empirical
rules of thumb exist, this study adopts an **automated selection
procedure** based on the minimization of Information Criteria,
specifically the **Akaike Information Criterion (AIC)** and the
**Bayesian Information Criterion (BIC)**. This approach ensures an
optimal balance between model fit and parsimony, mitigating the risks of
both underfitting (omitted variable bias) and overfitting (inflated
variance).

### 3.5 Formal Stationarity Testing: The ADF Approach

To rigorously confirm the integration order of the BTP-BUND spread, we
employ the **Augmented Dickey-Fuller (ADF) test** using the `ur.df()`
function from the `urca` package. Unlike standard implementations, this
approach requires a direct comparison between the calculated test
statistics ($\tau$) and the critical values provided by **Dickey and
Fuller (1979)**.

#### 3.5.1 Test Specification and Hypotheses

The test is conducted following a hierarchical strategy, accounting for
different deterministic components:

1.  Trend (`type = "trend"`): To check if the series is stationary
    around a deterministic time trend.

2.  Drift (`type = "drift"`): To test for stationarity around a constant
    non-zero mean (most appropriate for interest rate spreads).

3.  None (`type = "none"`): To test for stationarity around zero.

#### 3.5.2 Decision Rule

For the `ur.df()` output, we focus on the **tau statistic (**$\tau$).
The null hypothesis ($H_0$) of a unit root is rejected only if the test
statistic is **more negative** (i.e., lies in the lower tail) than the
critical value at the 5% significance level.

If the series in levels is found to be non-stationary, we proceed with
**first-differencing** ($\Delta \text{Spread}_t$) and re-apply the test.
The transformation is considered successful only when the tau statistic
for the differenced series falls significantly below the critical
threshold, confirming that the process is $I(1)$ and ready for GARCH
modelling.

```{r}
risultato_adf <- adfTest(train, lags = 1, type = "c")
print(risultato_adf)
# Equivalent with auto determination of lags

test_adf_completo <- function(serie, nome_serie = "Serie") {
  
  cat("\n==============================================\n")
  cat("ANALISI STAZIONARIETÀ (ADF):", nome_serie, "\n")
  cat("==============================================\n")
  
  # Definiamo i tre tipi di test ADF
  # "none"  = senza costante né trend
  # "drift" = con costante (intercetta)
  # "trend" = con costante e trend temporale
  tipi <- c("none", "drift", "trend")
  
  risultati_lista <- lapply(tipi, function(t) {
    # Eseguiamo il test con selezione automatica dei lag tramite AIC
    test <- ur.df(serie, type = t, selectlags = "AIC")
    
    # Estraiamo la statistica del test e il valore critico al 5%
    statistica <- test@teststat[1]
    valore_critico <- test@cval[1, "5pct"]
    
    # Verifichiamo la stazionarietà (statistica < valore critico)
    esito <- if (statistica < valore_critico) "STAZIONARIA" else "NON STAZIONARIA"
    
    return(data.frame(
      Modello = t,
      Statistica_ADF = round(statistica, 3),
      Valore_Critico_5pct = round(valore_critico, 3),
      Risultato = esito,
      stringsAsFactors = FALSE
    ))
  })
  
  # Uniamo i risultati in una tabella leggibile
  tabella_finale <- bind_rows(risultati_lista)
  print(tabella_finale)
  
  return(tabella_finale)
}
# --- COME USARLA ---

# 1. Test sullo spread originale (livelli)
risultati_livelli <- test_adf_completo(train, "Spread BTP/Bund (Livelli)")
```

###3.6 Transformation and Diagnostic of Differenced Series Following the
identification of a unit root in the levels of the spread, we proceed
with **first-differencing** to achieve stationarity, transforming the
series into an $I(0)$ process. This transformation represents the weekly
change in the BTP-BUND spread (measured in basis points).

To detect the presence of **Conditional Heteroskedasticity**—a
fundamental requirement for GARCH modelling—we analyze the
Autocorrelation Function (ACF) and Partial Autocorrelation Function
(PACF) of both the differenced series and its squared values. While the
standard ACF/PACF help identify the $ARMA$ structure for the mean
equation, the squared series analysis is crucial to identify
**volatility clustering**, where large shocks tend to be followed by
other large shocks, regardless of their sign."

```{r}
# 1. Calcolo della differenza prima (Weekly Change)
dspread <- diff(spread)
dtrain <- dspread[train_start:train_end]
dtest  <- dspread[test_start:test_end]

# 2. Controllo rapido delle dimensioni (per sicurezza)
cat("Lunghezza Train:", length(dtrain), "(",n_train_window/52 ," anni circa)\n")
cat("Lunghezza Test:", length(dtest), "(",n_test," settimane)\n")

# 3. Verifica se ci sono NA (fondamentale prima del GARCH)
if(any(is.na(dtrain))) {
  d_train <- na.omit(dtrain)
  cat("Attenzione: Rimossi valori NA dal train\n")
}
```

**3.6.1 Identification of the Mean and Variance Equations**

The diagnostic plots for the differenced spread
($\Delta \text{Spread}_t$) provide several key insights. The **ACF and
PACF of the residuals** show a highly significant spike at the first
lag, suggesting that the mean equation follows a low-order
Autoregressive ($AR$) or Moving Average ($MA$) process. Although a minor
spike appears at lag 13, it is deemed statistically borderline and
likely represents a noise-driven artifact; following the principle of
**parsimony**, we exclude it to avoid overfitting and ensure model
stability.

A critical observation arises from the **squared differenced series**.
The ACF of the squared values exhibits a persistent and significant
decay, which is the hallmark of **volatility clustering** or ARCH
effects. This indicates that the variance of the spread is not constant
over time (heteroskedastic), thus confirming the necessity of a
GARCH-type specification to correctly parameterize the risk dynamics.

Finally, regarding the **intercept of the mean equation**, a $t$-test
was performed to determine if the mean of the differenced series is
significantly different from zero. If the $p$-value is below 0.05, it
implies the presence of a significant **drift** component, which will be
incorporated into the $ARMA$ specification via the `include.mean = TRUE`
parameter. If the mean is statistically zero, the spread exhibits a pure
random walk behavior without a deterministic trend.

```{r, fig.width=12, fig.height=10, out.width="100%"}
par(mfrow = c(4, 1), mar = c(3, 4, 3, 2))
Acf(as.numeric(dtrain), lag.max = 30, main = "ACF of Diff Spread")
Pacf(as.numeric(dtrain), lag.max = 30, main = "PACF of Diff Spread")
Acf(as.numeric(dtrain)^2, lag.max = 30, main = "ACF of Squared Diff Spread")
Pacf(as.numeric(dtrain)^2, lag.max = 30, main = "PACF of Squared Diff Spread")
par(mfrow = c(1, 1))
```

```{r}

t.test(dtrain)
```

The $t$-test results indicate that the mean of the differenced series is
not statistically different from zero ($p$-value $> 0.05$).

##3.7 Formal Stationarity Testing on First Differences To confirm the
achievement of stationarity, we re-apply the Augmented Dickey-Fuller
(ADF) test. Given that the previous $t$-test confirmed a zero mean for
the differenced series, we utilize the "none" specification (type =
"none").Econometrically, this is a robust choice: if the spread in
levels exhibited a trend, the first difference would show a constant
drift. If we were to assume a trend in the differences, it would imply
that the spread is not just increasing, but accelerating—a scenario that
lacks long-term economic equilibrium. In financial theory, interest rate
spreads are considered "mean-reverting" and anchored to macroeconomic
fundamentals. ###3.7.1 Theoretical Framework: Key Literature This
modeling approach is supported by three foundational works:Vasicek
(1977): Introduced the mathematical framework for mean-reversion. It
justifies the exclusion of a trend in the ADF test, as rates are pulled
back toward a long-term equilibrium.Cox, Ingersoll, & Ross (1985):
Established that volatility is conditional on the level of the interest
rate, providing the theoretical predecessor to GARCH modeling.Balduzzi,
Das, & Foresi (1998): Analyzed the "Central Tendency" of bond yields,
supporting the idea that spreads oscillate around a fundamental mean.

```{r}
test_stationarity_diff<-adfTest(dtrain,lags=1,type="nc")
print(test_stationarity_diff)
#equivalent but with auto determination of lags
test_autolag <- ur.df(dtrain, type = "none", selectlags = "AIC")
summary(test_autolag)
stat_test <- test_autolag@teststat[1]
valore_critico_5pct <- test_autolag@cval[1, "5pct"]

# 3. Verifica automatica con messaggio
cat("\n--- VERIFICA RISULTATO ADF (Type = None) ---\n")
cat("Statistica del Test:", round(stat_test, 4), "\n")
cat("Valore Critico (5%):", valore_critico_5pct, "\n")

if (stat_test < valore_critico_5pct) {
  cat("ESITO: La serie è STAZIONARIA (p-value < 0.05)\n")
  cat("Significato: Puoi procedere con il modello GARCH.\n")
} else {
  cat("ESITO: La serie NON è STAZIONARIA\n")
  cat("Attenzione: Potrebbe esserci ancora una radice unitaria.\n")
}
```

The Verdict:\
We reject null hypotesysi of non stationarity

##3.8 Seasonality and ARCH Diagnostics Before finalizing the model
specification, we perform a definitive check for residual seasonality
and ARCH effects.

###3.8.1 Power Spectrum and Seasonality Tests The Periodogram is
utilized to analyze frequency strength. A pronounced peak at a specific
frequency (e.g., 0.083 for monthly cycles in weekly data) would indicate
a periodic component. Furthermore, we apply the seastests package for a
combined formal verification.

```{r}
# Calcola lo spettro di potenza
?spec.pgram
spec_raw <- spec.pgram(as.numeric(dtrain), spans = c(3,3), taper = 0.1, plot = TRUE,log='no')
# Trova la frequenza con il picco massimo
max_freq <- spec_raw$freq[which.max(spec_raw$spec)]
periodo_dominante <- 1 / max_freq
cat(periodo_dominante)
```

Interpretation of the Periodogram: The graph displays the power of
different frequencies within the series. A sharp peak at a frequency
corresponding to a 12-month (52-week) cycle would signal strong annual
seasonality. However, the absence of such peaks in our data further
confirms the lack of deterministic seasonal patterns.

```{r}

autoplot(dspread)

#To use this method, we need a ts object, let's create one
dspread_ts <- ts(dtrain, start=index(dtrain)[1], frequency = 8)#try also with 12
ggseasonplot(dspread_ts)

ggsubseriesplot(dspread_ts)

#Look also at decomposed series
decomposed <- stl(dspread_ts, s.window="periodic")
autoplot(decomposed)
```

### 3.8.2 Formal Verification of No-Seasonality

We use seastests package

```{r}
#install.packages("seastests")

isSeasonal(dspread_ts, test="combined")
#We get FALSE

#Individual tests
isSeasonal(dspread_ts, test="qs")
isSeasonal(dspread_ts, test="seasdum")
isSeasonal(dspread_ts, test="kw")
isSeasonal(dspread_ts, test="fried")

#We get all FALSE

## Now plot ACF and PACF
Acf(dspread_ts, lag.max= 52)
Pacf(dspread_ts, lag.max = 52)

#We have 1 significant lag at 1 in the Pacf, and a Acf which gradually goes to 0
#This suggests no seasonality.

#Last point, let's fit an arima model

adf.test(dspread_ts)     # stationarity test -> we have p-value = 0.546, so we fail to reject the null, so we are in the null workspace, this means no seasonality.

kpss.test(dspread_ts) # stationarity test where the null hypothesis is that the series IS stationary
# p-value is 0.01 -> we REJECT the null, so the series is NOT stationary

## Finally, we can affirm that our data does not contain seasonality, so we can work using normal ARIMA models and ARIMA-GARCH models.
```

We conclude the diagnostic phase by applying the seastests package. All
individual tests (QS, Seasonal Dummy, Kruskal-Wallis, and Friedman)
consistently fail to reject the null hypothesis of no seasonality.

Conclusion: Having confirmed that the series is stationary, lacks
seasonality, and exhibits significant ARCH effects, we can proceed with
confidence to estimate the non-seasonal ARIMA-GARCH framework.

###3.8.3 Formal Diagnostic Tests: Ljung-Box and ARCH-LM To ensure the
necessity of both the mean and variance components of the model, we
perform rigorous diagnostic tests on the differenced training set
($dtrain$). We select a high lag (e.g., $lag = 16$) to be meticulous in
detecting long-term dependencies.

```{r}
# 1. Ljung-Box Test on differences
lb_ret <- Box.test(as.numeric(dtrain), lag = 16, type = "Ljung-Box")
cat("\nLB Test on ds p-value:\n", lb_ret$p.value, "\n")
if(lb_ret$p.value < 0.05) cat("Autocorrelation detected, need mean modelling\n") else cat("ds are White Noise\n")

# 2. Ljung-Box Test on squared differences
lb_test <- Box.test(as.numeric(dtrain)^2, lag = 16, type = "Ljung-Box")
cat(sprintf("\nLjung-Box on Squared ds at lag 10:\n %.4f\n", lb_test$p.value))
if(lb_test$p.value < 0.05) cat("Volatility clustering confirmed\n") else cat("No volatility clustering detected\n")

# 3. ARCH-LM Test (Engle's Test for ARCH effects)
arch_test <- ArchTest(dtrain, lags = 10)
cat(sprintf("\nARCH Test (5 lags) LM Test - statistic: %.4f, p-value: %.4f\n", arch_test$statistic, arch_test$p.value))
if(arch_test$p.value < 0.05) cat("ARCH effects confirmed. GARCH modelling needed\n") else cat("No ARCH effects confirmed\n")
```

Interpretation of Results:Autocorrelation: If the $p$-value for the
first Ljung-Box test is below 0.05, we reject the null hypothesis of
independence, confirming that the spread changes exhibit autocorrelation
and require an ARMA specification.ARCH Effects: A $p$-value below 0.05
in both the squared Ljung-Box and the ARCH-LM test provides definitive
evidence of volatility clustering. This confirms that the variance is
not constant (Heteroskedasticity), mandating the use of a GARCH model to
capture time-varying risk.

##3.9 Seasonality Validation: Information Criteria Approach While
previous visual diagnostics suggested no seasonality, we utilize an
automated selection procedure to determine if a seasonal frequency
(SARIMA) provides a statistically superior fit. We compare models across
plausible frequencies ($f=1$ for non-seasonal, $f=8$, and $f=12$ for
monthly proxies in weekly data) based on the Akaike Information
Criterion (AIC).

```{r}
test_freqs <- c(1,8, 12) # Le frequenze plausibili
risultati <- data.frame(Freq = test_freqs, AIC = NA)

for(i in 1:length(test_freqs)) {
  # Crea la TS con la frequenza del loop
  test_ts <- ts(as.numeric(dtrain), frequency = test_freqs[i])
  
  # auto.arima testa se la stagionalità a quella frequenza aiuta il modello
  fit <- auto.arima(test_ts, max.q = 10, max.p = 10,seasonal = TRUE,method="ML",stepwise = FALSE,approximation = FALSE)
  risultati$AIC[i] <- fit$aic
}

# Trova la frequenza migliore
best_f <- risultati$Freq[which.min(risultati$AIC)]
cat("Il verdetto statistico è: usa la frequenza", best_f)
```

```{r}
train_ts12 <- ts(as.numeric(dtrain), frequency = 12)
m1_s12 <- auto.arima(train_ts12,max.q = 10, max.p = 10,method="ML",
                        seasonal = TRUE, 
                        stepwise = FALSE, 
                        approximation = FALSE,
                        trace = FALSE)
summary(m1_s12)
```

#### 3.9.1 Discussion of the SARIMA vs. ARIMA Trade-off

The algorithm identifies frequency 12 as the statistical optimum for the
10-year training set, suggesting a $SARIMA(4,0,0)(1,0,0)_{12}$
structure. However, it is noteworthy that for shorter estimation windows
(2 to 5 years), the optimal model reverts to a non-seasonal
$ARIMA(0,0,3)$ or simpler specifications.This discrepancy suggests that
the "seasonal" component detected in the long-term sample may not be a
deterministic calendar effect, but rather a reflection of medium-term
economic cycles or persistent shocks that mimic periodicity over long
horizons. In the interest of model parsimony and robust forecasting, the
choice between $SARIMA$ and $ARIMA$ must balance lower AIC scores with
the risk of overfitting noise.

##3.10 Final Mean Equation Specification Following the automated search,
we estimate the final non-seasonal $ARIMA$ model to evaluate the
significance of individual parameters.

```{r}
m1 <- auto.arima(as.numeric(dtrain),max.q = 10, max.p = 10,method="ML",
                        seasonal = FALSE, 
                        stepwise = FALSE, 
                        approximation = FALSE,
                        trace = FALSE)

summary(m1)
```

Model Robustness: The automated selection identifies an ARIMA(4,0,0) as
the optimal structure for the 10-year training period. It is important
to note that the model architecture exhibits sensitivity to the
estimation window: for a 2-year window, an ARIMA(0,0,3) is preferred,
while a 5-year window suggests a simpler structure. This variability
underscores the shifting short-term dynamics of sovereign risk.3.10.1
Statistical Significance of CoefficientsTo ensure model parsimony, we
evaluate the significance of each parameter. A coefficient is
statistically significant at the 95% confidence level if its t-ratio
($t = \frac{\hat{\phi}}{SE(\hat{\phi})}$) exceeds the critical value of
$|1.96|$.

```{r}
check_coef_significance <- function(model) {
  # Estraiamo coefficienti e standard errors
  coefs <- model$coef
  se <- sqrt(diag(model$var.coef))
  
  # Calcoliamo la t-statistica (z-score)
  t_stats <- coefs / se
  
  # Calcoliamo i p-value (distribuzione normale a due code)
  p_values <- (1 - pnorm(abs(t_stats))) * 2
  
  # Creiamo una tabella riassuntiva
  results <- data.frame(
    Coefficiente = round(coefs, 4),
    Std_Error = round(se, 4),
    t_stat = round(t_stats, 4),
    p_value = round(p_values, 4),
    Significativo_95 = ifelse(p_values < 0.05, "SI", "NO")
  )
  
  cat("\nAnalisi di Significatività dei Coefficienti:\n")
  print(results)
  
  cat("\nNota: Un coefficiente è significativo al 95% se |t_stat| > 1.96\n")
  print(results)
}
check_coef_significance(m1)
m1$coef
```

ARE ALL THE COEFFICIENTS OF m1 STATISTICALLY SIGNIFICANT? t ratios are
given by estimated coefficients divided by their standard errors; the
latter are given by the square root of the element on the main diagonal
(variance of coef)

```{r}
names(m1)
summary(m1)
t.ratios = m1$coef/sqrt(diag(m1$var.coef))
t.ratios
```

```{r}
# Il vettore fixed deve avere la stessa lunghezza del numero di parametri (3 MA + 1 media)
# Usiamo NA per i parametri da stimare e 0 per quelli da bloccare
m1 <- arima(as.numeric(dtrain), 
                order = c(4,0,0),
                fixed = c(0, NA, NA, NA, 0)) # 4 AR + 1 mean that summary(m1) says it is 0
checkresiduals(m1)
```

Note that ACF of redisuals present a significant spike at lag 12

```{r}
arma_assumptions <- function(m1) {
  cat("\n============================================================\n")
  cat("ARIMA model assumptions & Cycle Analysis:\n")
  
  # Estrazione corretta per modelli auto.arima/Arima
  # Usiamo i coefficienti AR per la stazionarietà
  ar_coefs <- m1$model$phi
  ma_coefs <- m1$model$theta
  
  # 1. Calcolo Radici
  ar_roots <- polyroot(c(1, -ar_coefs))
  ma_roots <- polyroot(c(1, ma_coefs))
  
  is_stationary <- all(abs(ar_roots) > 1)
  is_invertible <- all(abs(ma_roots) > 1)
  
  cat("Stationary:", is_stationary, "(Min mod:", round(min(abs(ar_roots)), 3), ")\n")
  cat("Invertible:", is_invertible, "(Min mod:", round(min(abs(ma_roots)), 3), ")\n")
  
  # 2. Analisi dei Business Cycles (Radici Complesse)
  # Cerchiamo se esistono radici con parte immaginaria significativa
  complex_roots <- ar_roots[Im(ar_roots) != 0]
  
  cat("\n--- Business Cycle Analysis ---\n")
  if(length(complex_roots) > 0) {
    cat("The model IMPLIES the existence of cycles.\n")
    cat("Reason: The AR characteristic equation has complex conjugate roots.\n")
    
    # Calcolo del periodo del ciclo (se ti serve per l'analisi)
    # Periodo = 2*pi / acos(Re(root)/abs(root))
    root <- complex_roots[1]
    period <- 2 * pi / atan2(Im(root), Re(root))
    cat("Estimated average cycle period:", round(abs(period), 2), "units of time.\n")
  } else {
    cat("The model DOES NOT imply cycles.\n")
    cat("Reason: All AR roots are real; the model returns to equilibrium monotonically.\n")
  }
  
  # 3. Check for common roots
  common_roots <- FALSE
  if(length(ar_roots) > 0 && length(ma_roots) > 0) {
    for(ar in ar_roots) {
      for(ma in ma_roots) {
        if(abs(ar - ma) < 0.05) common_roots <- TRUE
      }
    }
  }
  
  cat("\n--- Root Validation ---\n")
  if(common_roots) {
    cat("Common roots detected, assumption violation\n")
  } else {
    cat("No common roots detected, assumption validated\n")
  }
  cat("============================================================\n")
}

# Esecuzione
arma_assumptions(m1)
```

#### Residual's check

NOTE: Kolmogorov-Smirnov test (KS) to test wheter they follow a
t-distribution $H_0: \text{serie follows supposed distribution}$

```{r}

cat("\n============================================================\n")
cat("\nARMA residuals check:\n")
# ARIMA residuals
n_parameters<-sum(m1$coef!=0) 
arima_res <- residuals(m1)
arma_residual <- function(res) {
  cat("\n============================================================\n")
  cat("ANALISI APPROFONDITA DEI RESIDUI\n")
  cat("============================================================\n")

# 1. Ljung-Box (Media e Autocorrelazione)
  lb_arima <- Box.test(res, lag = 15, type = "Ljung-Box",fitdf = n_parameters)
  cat("\n [1]LB test on ARIMA residuals (15 lags) p-value: ", round(lb_arima$p.value, 4), "\n")
  if(lb_arima$p.value >= 0.05) cat("ARIMA model cleared the mean\n") else cat("ARIMA model DID NOT clear the mean\n")
#Regola di Hyndman (autore di forecast): Per dati non stagionali si usa $h = 10$, per dati stagionali si usa $h = 2m$ (dove $m$ è la frequenza). Nel tuo caso ($m=8$), dovresti testare almeno fino a 16 lag.
  
# 2. Durbin-Watson (Autocorrelazione al lag 1)
# Nota: Il DW classico si usa su modelli lineari, sui residui ARIMA è indicativo
  dw_test <- dwtest(as.vector(res) ~ 1)
  cat("\n[2] Durbin-Watson Test p-value: ", round(dw_test$p.value, 4))
  if(dw_test$p.value >= 0.05) cat(" -> OK: No autocorrelazione lag 1\n") else cat(" -> KO: Autocorrelazione lag 1\n")

# 3. Breusch-Godfrey Test (Autocorrelazione di ordine superiore)
  bg_test <- bgtest(as.vector(res) ~ 1, order = 52)
  cat("\n[3] Breusch-Godfrey Test (order 52) p-value: ", round(bg_test$p.value, 4))
  if(bg_test$p.value >= 0.05) cat(" -> OK: No autocorrelazione superiore\n") else cat(" -> KO: Autocorrelazione superiore\n")
  
# 4. Jarque-Bera Test (Normalità dei residui)
  jb_test <- jarque.bera.test(res)
  cat("\n[4] Jarque-Bera Test p-value: ", round(jb_test$p.value, 4))
  if(jb_test$p.value >= 0.05) cat(" -> OK: Residui Normali\n") else cat(" -> KO: Residui NON Normali (code grasse)\n")
  
#5. Kolmogorov-Smirnov (KS)
  check_t <- function(residuals) {
  fit_t <- suppressWarnings(fitdistr(residuals, densfun = "t"))
  mu <- fit_t$estimate["m"]
  sigma <- fit_t$estimate["s"]
  df <- fit_t$estimate["df"]
  std_res <- (residuals - mu) / sigma
  ks <- ks.test(x = std_res, y = "pt", df = df)
  cat("\n[5] KS test assuming t-distribution:\n")
  print(c(DoF = df, KS_p_value = ks$p.value))
  if(ks$p.value > 0.05) cat("\nResiduals are t-distributed\n") else cat("\nResiduals are NOT t-distributed\n")
  }
  check_t(res)
  # 6. Effetti ARCH (Eteroschedasticità)
  lb_arima_2 <- Box.test(res^2, lag =10, type = "Ljung-Box",fitdf = n_parameters)
  cat("\n [6]LB test on ARIMA squared residuals (10 lags) p-value: ", round(lb_arima_2$p.value, 4), "\n")
  if(lb_arima_2$p.value < 0.05) cat("ARIMA model has residual variance to explain. GARCH model needed\n") else cat("ARIMA model residual variance is OK\n")

  # 7.ARCH test
  arch_test_res <- ArchTest(res, lags = 5)
  cat(sprintf("\n [7]ARCH Test on ARIMA (5 lags) residuals:\n LM Test - statistic: %.4f, p-value: %.4f\n", 
              arch_test_res$statistic, arch_test_res$p.value))
  if(arch_test_res$p.value < 0.05) cat("ARCH effects confirmed. GARCH modelling needed\n") else cat("No ARCH effects confirmed\n")
}

print(arma_residual(arima_res))
```

ORA FACCIAMO STA BENEDETTA PREVISIONE:

IN SAMPLE PREDICTIONS

```{r}
# Estraiamo i valori fittati dal modello m1
fitted_values <- fitted(m1)
length(fitted_values)
length(train)
# Visualizzazione grafica: Reale vs Fittato
plot(as.numeric(dtrain), type="l", col="black", lwd=1, 
     main="In-Sample Evaluation: Reale vs ARIMA",
     ylab="Spread", xlab="Tempo (Osservazioni)")
lines(as.numeric(na.omit(fitted_values)), col="red")
legend("topright", legend=c("Dato Reale", "Fitted"), col=c("black", "red"), lty=1, lwd=2)
```

IN SAMPLE EVALUATION

```{r}
residuals <- dtrain - fitted_values

# Core metrics
accuracy(as.numeric(dtrain), as.numeric(fitted_values))
# R-squared (will be 0 or near 0 — expected for mean model)
SS_res <- sum(residuals^2, na.rm = TRUE)
SS_tot <- sum((dtrain - mean(train, na.rm = TRUE))^2, na.rm = TRUE)
R2 <- 1 - SS_res / SS_tot
cat("R2 del modello:", round(R2, 2), "%") 

# Calcoliamo i segni (1 per positivo, -1 per negativo)
sign_actual <- sign(dtrain)
sign_fitted <- sign(fitted_values)

# Creiamo un vettore logico: TRUE se i segni sono uguali
hits <- (sign_actual == sign_fitted)

# Calcoliamo la percentuale di successo (escludendo eventuali NA)
DA <- mean(hits, na.rm = TRUE) * 100
DA
```

OUT-SAMPLE PREDICTION AND EVALUATION

```{r}
# 1. Applichiamo il modello alla serie completa (genera tutti i forecast 1-step)
refit <- Arima(dspread, model = m1)

# 2. Estraiamo solo i valori fittati che corrispondono al periodo di TEST
# (Supponiamo che il tuo test set sia lungo 33 osservazioni)
pred_out <- tail(fitted(refit), n_test)
reali_out <- tail(as.numeric(dspread), n_test)

# 3. Calcolo metriche Out-of-Sample
accuracy(pred_out, reali_out)

# R-squared (will be 0 or near 0 — expected for mean model)
SS_res <- sum((pred_out-reali_out)^2, na.rm = TRUE)
SS_tot <- sum((test - mean(test, na.rm = TRUE))^2, na.rm = TRUE)
R2 <- 1 - SS_res / SS_tot
cat("R2 del modello:", round(R2, 2), "%") 

# Calcoliamo i segni (1 per positivo, -1 per negativo)
sign_actual <- sign(test)
sign_fitted <- sign(pred_out)
hits <- (sign_actual == sign_fitted)
# Calcoliamo la percentuale di successo (escludendo eventuali NA)
DA <- mean(hits, na.rm = TRUE) * 100
cat("DA del modello:", round(DA, 2), "%") 

```

FORECAST DI SETTIMANA PROSSIMA (diff_spreads)

```{r}
# 2. Generiamo la previsione per la prossima settimana
forecast_next_week <- forecast(refit, h = 1)

# 3. Visualizziamo il risultato
print(forecast_next_week)

# 4. Grafico della previsione con intervalli di confidenza
plot(forecast_next_week, main = "Previsione Spread BTP-Bund: Settimana Prossima", 
     ylab = "Spread", xlab = "Tempo")
```

Lo 80 / Hi 80 (da -21.79 a +22.18): Il modello è sicuro all'80% che la
variazione dello spread oscillerà tra una discesa di 21.8 punti e una
salita di 22.2 punti. È un range abbastanza ampio che riflette la
volatilità tipica dei mercati.

Lo 95 / Hi 95 (da -33.43 a +33.82): Il modello è sicuro al 95% (standard
statistico per la tesi) che la variazione non supererà i ±33 punti base.

Nota tecnica: Se la prossima settimana lo spread dovesse muoversi di 50
punti, sapresti che è successo un evento "fuori dal comune" (un outlier)
che il modello non poteva prevedere. TRASFORMAZIONE IN spread:

```{r}
# 1. Recuperiamo l'ultimo valore REALE dello spread (non differenziato)
ultimo_spread_reale <- tail(spread, 1) 

# 2. Otteniamo la previsione della differenza (il "delta")
previsione_diff <- forecast(m1_s12, h = 1)$mean[1]

# 3. Calcoliamo lo spread atteso
spread_previsto <- ultimo_spread_reale + previsione_diff

cat("Lo spread previsto per settimana prossima è:", spread_previsto)
```

CONSIDERARE DI FARE UNA CROSS VALIDATION PER UNA ROBA PIU ACCURATA

############################################################################################################ 

## Volatility Analysis

We restart from our model

```{r}
summary(m1)
```

Remark that the mean is = 0 so we can work directly on residuals

```{r}
fitted_values <- fitted(m1)
res<-train - fitted_values
par(mfrow = c(4, 1), mar = c(3, 4, 3, 2))

plot(index(as.xts(dspread)), coredata(as.xts(dspread)), type = "l", lwd = 0.7,
     main = "Residuals^2 ", xlab = "", ylab = "ds")
abline(h = 0, col = "black", lwd = 0.5)
Acf(as.numeric(res)^2,   lag.max = 30, main = "ACF of res^2")
Pacf(as.numeric(res)^2,  lag.max = 30, main = "PACF of res^2")
par(mfrow = c(1, 1))
```

Check for autocorrelation and ARCH effects on `train`:

```{r}
n_parameters<-sum(m1$coef!=0) 
# Ljung-Box on DS^2
lb_ds2 <- Box.test(as.numeric(res)^2, lag = 10, type = "Ljung-Box",fitdf = n_parameters)
cat(sprintf("LB Test on DS^2 p-value: %.4f\n", lb_ds2$p.value))
if (lb_ds2$p.value < 0.05) cat("Volatility clustering confirmed\n") else cat("No volatility clustering detected\n")

# ARCH test
arch_test <- ArchTest(as.numeric(res), lags = 10)
cat(sprintf("ARCH Test (10 lags) p-value: %.4f\n", arch_test$p.value))
if (arch_test$p.value < 0.05) cat("ARCH effects confirmed — GARCH modelling needed\n") else cat("No ARCH effects\n")

```

## Model 3: ARMA-GARCH

We try to find the best ARMA-GARCH fitting different combinations of
models, then decide considering a precise IC, kinda like an auto_arima
but hand-builded.

Create a function that simulates 1 iteration of a for cycle, then
instead of a for cycle use parrallel (mclapply) It will run the function
over $n$ times using $n$ CPU simultanealy.

Can change the critera or modify the parameters, if too much
combinations the code will run for ages.

We will keep the max possible GARCH orders to 1, that's cause vast
majority of the time a GARCH(1,1) is perfect for financial series. We
could add more GARCH model other than `sGARCH`and `eGARCH` but since
those are the most used we keep only them, also other models have
additional parameter than can increase the computational time. We also
capped max ARMA order to 3 to prevent the AIC from picking up noise (it
does, tried to run multiple times and it gives back models that do not
pass diagnostic tests).

The ugarch solvers (`gosolnp`) use random generate starting point, to
have reproducibility of results we need to impose a seed whenever a
optimization is done.

**NOTE**: in `dists` you choose the distribution you suppose the std
residuals will follow, `norm`= normal, `std`= t, `sstd`= skewed t. I
removed `norm`and cause the function would wrongly choose `norm` over
the other 2: you can see if is the distribution fit the residuals in the
printed residual plot or in the "Adjusted Pearson Goodness-of-Fit Test"
at the end of the model output, when `norm` the fit is worse. The
distribution that better fit the residuals is `sstd` (even if not
perfect).

The reason is that `sstd` and `std` have more parameters than `norm`
(number of parameter: `sstd` \> `std` \> `norm`) and since the train
data is "small" this penalize them more than it should in the eye of
AIC/BIC. `norm` will end up with better IC than `sstd` but then you have
the wrong distribution for the residuals and/or fail the diagnostic
tests.

**NOTE 2**: `mc.cores = detectCores() - 1` is a MAC thing, on Windows it
does not work, it will show that the code is running on x core but
actually will use just one, like a normal `for` cycle.

```{r,}
start_time <- proc.time() # to see for how much the code run

best_garch <- function(y,
                       ar_lags  = 0:5,
                       ma_lags  = 0:5,
                       garch_p  = 1:3,
                       garch_q  = 1:3,
                       models   = c("sGARCH", "eGARCH"),
                       dists    = c("sstd", "std"),
                       criteria = "aic", 
                       timeout  = 60, 
                       seed = 42) {

  # reproducibility for the hybrid solver, it uses random starting point
  RNGkind("L'Ecuyer-CMRG")
  set.seed(seed)

  # Create every combinations possible for the parameter
  combos <- expand.grid(
    ar_lag  = ar_lags,
    ma_lag  = ma_lags,
    garch_p = garch_p,
    garch_q = garch_q,
    model   = models,
    dist    = dists,
    stringsAsFactors = FALSE
  )

  cat("Total combinations to fit:", nrow(combos), "\n")
  cat("Running on", parallel::detectCores() - 1, "cores\n")
  cat("It could take couple of minutes ...\n\n")

  fit_one <- function(i) {
    arp   <- combos$ar_lag[i]
    map   <- combos$ma_lag[i]
    gp    <- combos$garch_p[i]
    gq    <- combos$garch_q[i]
    model <- combos$model[i]
    dist  <- combos$dist[i]

    spec <- ugarchspec(
      variance.model     = list(model = model, garchOrder = c(gp, gq)),
      mean.model         = list(armaOrder = c(arp, map), include.mean = TRUE),
      distribution.model = dist
    )
    # If R encounter an error it stop the cycle, tryCatch catch the error and return NULL, then continue
    fit <- tryCatch({
      withTimeout({
        suppressWarnings(
          ugarchfit(spec = spec, data = y, solver = "gosolnp", 
          solver.control = list(n.restarts = 5, rseed = 42, trace = 0)) # for reproducibility)
        )
      },
      timeout   = timeout,
      onTimeout = "silent")
    },
    error   = function(e) NULL)

    # Store only not null + right class
    if (!is.null(fit) && inherits(fit, "uGARCHfit") && fit@fit$convergence == 0) {
      ic <- infocriteria(fit)
      return(data.frame(
        ar_p    = arp,
        ma_q    = map,
        garch_p = gp,
        garch_q = gq,
        model   = model,
        dist    = dist,
        aic     = as.numeric(round(ic["Akaike", ] * length(y), 2)), # the function calculate the aic only for that obs
        bic     = as.numeric(round(ic["Bayes",  ] * length(y), 2)),
        likelihood = as.numeric(round(likelihood(fit), 2)),
        stringsAsFactors = FALSE
      ))
    }
    return(NULL)
  }

# Run more itrations simultanely with every CPU-1, instead of for cycle
# results_list = mclapply(
#    seq_len(nrow(combos)),
#    fit_one,
#    mc.cores = detectCores() - 1
#  )
  #versione per Windows:
  #  Crea il "ponte" tra i core del tuo PC (Cluster)
  cl <- makeCluster(detectCores() - 1)
  # Esporta tutto ciò che serve ai core "figli" (fondamentale su Windows!)
  # Esportiamo i dati (y), la tabella (combos), la funzione (fit_one) e il timeout
  clusterExport(cl, varlist = c("combos", "y", "timeout", "fit_one"), envir = environment())
  # Carica le librerie su ogni singolo core
  clusterEvalQ(cl, {
    library(rugarch)
    library(R.utils)
    library(dplyr)
  })  
  # Esegui il calcolo (parLapply sostituisce mclapply)
  results_list <- parLapply(cl, seq_len(nrow(combos)), fit_one)
  # IMPORTANTE: Spegni i motori una volta finito
  stopCluster(cl)  

# Keep only non NULL model
  successful <- Filter(Negate(is.null), results_list)
  cat("Successful fits:", length(successful), "/", nrow(combos), "\n")
  cat("Models ordered by: ", criteria, "\n")

  results_df <- bind_rows(successful) |> arrange(.data[[criteria]])
  return(results_df)
}

results = best_garch(train)
cat("\nFitted models:\n")
print(head(results))

end_time <- proc.time()
execution_time <- end_time - start_time
cat("\n Total execution time: ", execution_time[3], "seconds")
```

best model with dtrain(10 anni) = ARMA-GARCH(5,4)-(1,3)

#Computed using the last 5 years from the 10 year data, 
#[1] "Lunghezza Train: 260"
#[1] "Lunghezza Test: 52"
#Training observations: 260
#Test observations:     52
#Train period: 2020-04-06 -> 2025-03-24
#Test period:  2025-03-31 -> 2026-03-23
best model with dtrain (5 anni) + dtest (1 anno) = ARMA-GARCH(5,5)-(1,1)

#TODO Try to use 8years train + 2 years test -> see which is the best

## OR use daily data, using 3 years?
Extract the best model.

```{r, fig.width=12, fig.height=10, out.width="100%",}
arma_best_order <- c(results$ar_p[1], results$ma_q[1])
garch_best_order  <- c(results$garch_p[1], results$garch_q[1])
best_dist  <- results$dist[1]
best_mod  <- results$model[1]

cat("\nBest ARMA-GARCH model orders:\nARMA:  ", arma_best_order, "\nGARCH: ", garch_best_order,"\n") 
  garch_spec <- ugarchspec(variance.model = list(model = best_mod, garchOrder = garch_best_order),
                         mean.model = list(armaOrder =arma_best_order, include.mean = TRUE),
                         distribution.model = best_dist)

set.seed(42)
garch_fit <- ugarchfit(spec = garch_spec, data = dtrain, solver = "gosolnp", 
solver.control = list(n.restarts = 5, rseed = 42, trace = 0)) # for reproducibility


```

## Diagnostic

```{r,}
show(garch_fit)
```

In the ARMA(0,3)-GARCH(1,1) output look at:

-   "Weighted Ljung-Box Test on Standardized Residuals"
-   "Weighted Ljung-Box Test on Standardized Squared Residuals"
-   "Weighted ARCH LM Tests"
-   "Adjusted Pearson Goodness-of-Fit Test" --\> HO: la distribuzione
    scelta fitta bene i dati

In addition in the last part of output:Tutti i p-value sono molto
superiori a 0.05. Questo è un ottimo risultato per il tuo modello
attuale: significa che non c'è bias del segno significativo. Il modello
GARCH che hai stimato sembra gestire correttamente sia gli shock
positivi che quelli negativi senza mostrare distorsioni sistematiche.

The model pass all the tests.

Let's add some visualization prove hof the distribution chosen

```{r,}
plot(garch_fit,which=9)#does not seem too good
plot(garch_fit,which=8)
```

We also test if there's any breakpoint inside the residuals. We use:

-   Bau-Perron test: tests for multiple unknow structural breaks in the
    mean, it find the best breaks location that minimizes the RSS, then
    selects the number of breaks minimizing BIC. The number of breaks
    with lowest BIC is the correct one.
-   CUSUM: tests for parameter instability by tracking the cumulative
    sum of recursive residuals, a deviation of this sum from 0 means
    instability, $H_0:$ stable parameters throught the sample.

```{r, fig.width=12, fig.height=10, out.width="100%",}
par(mfrow = c(2, 1))
resid_std <- residuals(garch_fit, standardize = TRUE) |> as.numeric()

bp_resid_df <- data.frame(
  resid = resid_std,
  date  = index(dtrain) 
)
bp_resid <- breakpoints(resid ~ 1, data = bp_resid_df, h = 0.15)
summary(bp_resid)
plot(bp_resid)

cusum_resid <- efp(resid ~ 1, data = bp_resid_df, type = "Rec-CUSUM")
plot(cusum_resid)
sctest(cusum_resid)

par(mfrow = c(1, 1))
```

Checking the RSS table we can see that we have the optimal BIC at 0
breaks -\> the residuals do not need breakpoint, adding any break will
make the model worse. Also the RSS remain almost equal with differents
possibles breaks: adding breaks will not explain anything more.

We fail to reject the null in the CUSUM test since the line in plot stay
inside the confidence intervals -\> parameters are stable.

Combined both tests tell us that there's no evidence of structural
breaks in the standardize residuals, any possible change presents in the
raw series is explained by the model.

## Comment

The estimated model pass all the tests: there's no mean or volatility to
model and the residual fit the supposed distribution (t-skewed) and pass
the tests. Our data can be modelled with this ARMA-GARCH model, next we
run a validation process for its forecasts and collect some metrics. \#
Validation Now the validation phase, we take the best estimated
ARMA-GARCH and validate through the test data. At every iteration we
refit the model (re-estimate the parameter) but the structure of the
model remain the same: the goal is to calculate how good the model is at
forecasting to then forecast the real future value.

Then we compare the forecast with a Random Walk model to test it our
model is actually useful.

-   Forecast on test data using rolling forecast (window increases on 1
    week every iteration)
-   Refit the model every 1 iteration (1 week) to keep model useful
-   Perform the rolling estimation and forecast
-   refit.every = 1 means we re-estimate the model every single
    iteration
-   If window.size too small the model will fail to converge

```{r,}
start_time <- proc.time()

c1 <- makeCluster(detectCores() - 1)

window_size = 110

set.seed(42)

roll <- ugarchroll(spec = garch_spec, 
                   data = dspread, 
                   n.ahead = 1, 
                   forecast.length = length(test), 
                   refit.every = 1, 
                   refit.window = "moving", 
                   window.size = window_size, 
                   solver = "gosolnp",
                   solver.control = list(n.restarts = 5, rseed = 42, trace = 0),
                   calculate.VaR = TRUE,
                   VaR.alpha = c(0.05, 0.95), # calculate VaR using the GARCH model
                   cluster = c1)

stopCluster(c1)
# In case of non-convergence problems, re-run with other solvers
if (length(roll@model$nonconverged) > 0) {
  roll <- resume(roll, solver = "nloptr")
}
if (length(roll@model$nonconverged) > 0) {
  roll <- resume(roll, solver = "solnp")
}
if (length(roll@model$nonconverged) > 0) {
  roll <- resume(roll, solver = "hybrid")
}

# Extract density predictions and realized spread
roll_df <- as.data.frame(roll)

# Extract theoretical quantiles
quants <- as.data.frame(roll, which = "VaR")

# Build the forecast_df for better reading
dfc_df <- data.frame(
  pred_mu  = roll_df$Mu,
  pred_var = roll_df$Sigma^2,
  pred_vol = roll_df$Sigma,
  ci_q05  = quants[, "alpha(5%)"],
  ci_q95  = quants[, "alpha(95%)"],
  actual   = roll_df$Realized
)
rownames(dfc_df) <- rownames(roll_df)
print(dfc_df)
end_time <- proc.time()
execution_time <- end_time - start_time
cat("\n Total execution time: ", execution_time[3], "seconds\n")
```

Convert back to spread

```{r,}
get_spreads_fixed <- function(original_series, forecast_results) {
  # 1. Recupera le date delle previsioni (sono i rownames di dfc_df)
  forecast_dates <- as.Date(rownames(forecast_results))
  n <- length(forecast_dates)
  
  # Preparazione vettori
  fc_spread    <- numeric(n)
  lower_spread <- numeric(n)
  upper_spread <- numeric(n)
  act_spread   <- numeric(n)
  
  # Tutte le date disponibili nella serie originale
  all_dates <- index(original_series)
  
  for (i in 1:n) {
    curr_date <- forecast_dates[i]
    
    # Trova la posizione esatta della data corrente nella serie originale
    pos <- which(all_dates == curr_date)
    
    if (length(pos) > 0) {
      # Valore REALE oggi
      act_spread[i] <- as.numeric(original_series[pos])
      
      # Valore REALE del periodo precedente (per sommare il rendimento previsto)
      # Questo risolve lo sfasamento di "5 osservazioni" o di "1 osservazione"
      prev_val <- as.numeric(original_series[pos - 1])
      
      fc_spread[i]    <- prev_val + forecast_results$pred_mu[i]
      lower_spread[i] <- prev_val + forecast_results$ci_q05[i]
      upper_spread[i] <- prev_val + forecast_results$ci_q95[i]
    }
  }
  
  # Costruzione dataframe finale
  res_df <- data.frame(
    act_spread = act_spread,
    fc_spread  = fc_spread,
    ci_q05     = lower_spread,
    ci_q95     = upper_spread,
    row.names  = forecast_dates
  )
  return(res_df)
}

# ESECUZIONE
fc_df <- get_spreads_fixed(spread, dfc_df)
print(tail(fc_df))
```

Now the metrics

```{r,}
# Mean evaluation
errors <- fc_df$act_spread - fc_df$fc_spread
mae <- mean(abs(errors))
mse <- mean(errors^2)
rmse <- sqrt(mse)
mape <- mean(abs(errors / fc_df$act_spread))
dir_acc <- mean(sign(fc_df$act_spread) == sign(fc_df$fc_spread))

cat("\nMean evaluation metrics:\n")
cat(sprintf("MAE: %.4f  |  MSE: %.4f  |  RMSE: %.4f  |  MAPE: %.4f  |  Dir. Acc: %.2f%%\n", 
            mae, mse, rmse, mape, dir_acc * 100))

## Distribution check ##
# Check if actual value fall into 90%
# 90% interval: q05 to q95
coverage_90 <- mean(fc_df$act_spread >= fc_df$ci_q05 & 
                    fc_df$act_spread <= fc_df$ci_q95)

cat("\nDistribution evaluation, % of forecast that fall into that interval:\n")
cat(sprintf("90%% coverage: %.4f\n", coverage_90)) 
```

Compare those metrics with a Random Walk model (tomorrow's value =
today's value). To see if our model is actually useful.

```{r,}
spread_rw <- spread[length(train):(length(train) + length(test) - 1)] |> as.numeric()

fc_df$spread_rw <- spread_rw 

mae_rw   <- mean(abs(spread_rw - fc_df$act_spread))
rmse_rw  <- sqrt(mean((spread_rw - fc_df$act_spread)^2))
mape_rw  <- mean(abs((spread_rw - fc_df$act_spread) / fc_df$act_spread))
dir_rw   <- mean(sign(diff(spread_rw)) == sign(diff(fc_df$act_spread)))

cat("\nEvalutation metrics, Model vs Random Walk:\n")
cat(sprintf("\nModel  | MAE: %.4f | RMSE: %.4f | MAPE: %.2f%% | Dir: %.2f%%\n",
            mae, rmse, mape*100, dir_acc*100))
cat(sprintf("RW   | MAE: %.4f | RMSE: %.4f | MAPE: %.2f%% | Dir: %.2f%%\n",
            mae_rw, rmse_rw, mape_rw*100, dir_rw*100))
```

The metrics are identical -\> our model is useless, it acts like a
Random Walk.

```{r, fig.width=12, fig.height=10, out.width="100%",}
fc_df$date <- as.Date(rownames(fc_df))

p_spreads  <- ggplot(fc_df, aes(x = date)) +
  geom_line(aes(y = act_spread, color = "Actual"), linewidth = 0.8) +
  geom_line(aes(y = fc_spread, color = "Forecasted"), linewidth = 0.8) +
  geom_line(aes(y = spread_rw, color = "Random Walk"), linewidth = 0.8) +
  geom_ribbon(aes(ymin = ci_q05, ymax = ci_q95), fill = "blue", alpha = 0.3) +
  scale_color_manual(values = c("Actual" = "black", "Forecasted" = "blue", "Random Walk" = "red")) +
  labs(title = "Spread: Actual vs Forecasted", x = "", y = "Basis points") +
  scale_x_date(date_breaks = "2 month", date_labels = "%b-%y") +
  scale_y_continuous(breaks = scales::breaks_width(20)) +
  theme_minimal() +
  theme(legend.title = element_blank(), legend.position = "bottom")

print(p_spreads)
```

# Forecast

Forecast future value for next week, use in the same structure of the
validated model.

Use the last `window_size` data (the same as validation phase)

```{r,}
# Forecast future value for next week
# Use the last window_size data

# 1. Recupero dati e data (usiamo dspread che è l'oggetto xts/zoo)
data_fit_obj <- tail(dspread, window_size) 
data_fit     <- as.numeric(data_fit_obj)
last_date    <- index(tail(data_fit_obj, 1))

set.seed(42)
future_fit <- ugarchfit(spec = garch_spec, data = data_fit, solver = "gosolnp",
                        solver.control = list(n.restarts = 5, rseed = 42, trace = 0))

future_fc_obj <- ugarchforecast(future_fit, n.ahead = 1)

# Estrazione quantili e valori
q_05_val <- as.numeric(quantile(future_fc_obj, probs = 0.05))
q_95_val <- as.numeric(quantile(future_fc_obj, probs = 0.95))

future_fc <- data.frame(
  d_spread = as.numeric(fitted(future_fc_obj)),
  vol      = as.numeric(sigma(future_fc_obj)),
  q_05     = q_05_val,
  q_95     = q_95_val
)

# 2. Funzione di conversione (Assicurati che 'data' sia un vettore numerico o xts)
spread_conversion <- function(value, q_05, q_95, data){
  # Usiamo as.numeric per forzare il calcolo matematico
  last_spread <- as.numeric(tail(data, 1))
  
  res_spread <- last_spread + value
  res_q05    <- last_spread + q_05
  res_q95    <- last_spread + q_95
  
  result_df  <- data.frame(
    spread = res_spread,
    q_05   = res_q05,
    q_95   = res_q95
  )
  return(result_df)
}

# 3. CHIAMATA CRITICA: Usa 'spread' (l'oggetto zoo dei livelli) NON 'df'
future_spread <- spread_conversion(future_fc$d_spread, future_fc$q_05, future_fc$q_95, spread)

# Output
cat("\nForecasted values:\n")
cat("\nSpread on last date (", as.character(last_date), "): ", as.numeric(tail(spread, 1)))
cat("\nSpread forecasted for next week (", as.character(last_date + 7), "):\n")
print(future_spread)
```

## Simulation Forecast (10,000 paths)

Simulating 10,000 paths accounts for the GARCH shock uncertainty, not
just the ARMA mean.

```{r,}
simulation <- ugarchsim(
  future_fit,
  n.sim       = 1,
  m.sim       = 10000,
  startMethod = "sample"
)

mean_sim <- mean(fitted(simulation))
vol_sim  <- mean(sigma(simulation))
q_90     <- quantile(fitted(simulation), probs = c(0.05, 0.95))

cat(sprintf("Mean DS:  %.4f\n", mean_sim))
cat(sprintf("Mean Vol: %.4f\n", vol_sim))
cat(sprintf("Quantiles 5%% | 95%%: %.4f | %.4f\n", q_90[1], q_90[2]))

sim_spread <- spread_conversion(mean_sim, unname(q_90[1]), unname(q_90[2]), spread)
cat("Spread forecast (simulation) for next week:", as.character(last_date + 7), ":\n")
print(sim_spread)
```

Se i risultati di sim_spread sono molto simili a quelli di future_spread
(ottenuti con forecast), significa che il tuo modello è stabile. Se sono
molto diversi, potrebbe esserci molta incertezza o asimmetria nella
distribuzione dei rendimenti.

Ti sono usciti dei numeri questa volta o hai ancora il problema degli
NA? (Se hai usato spread nella funzione come suggerito, dovresti vedere
i risultati corretti).
