---
title: "Spread: ARMA-GARCH Model"
output:
  html_document: default
  pdf_document: default
date: "2026-04-30"
editor_options:
  markdown:
    wrap: 72
---

# Data setup
```{r, include = FALSE}
rm(list = ls())
gc()
```

```{r, message=FALSE, warning=FALSE, include=FALSE}
library(ggplot2)
library(patchwork)
library(tidyverse)
library(vars)
library(rmgarch)
library(rugarch)
library(MTS)
library(urca)
library(xts)
library(readxl)
library(lmtest)
library(MASS)
library(scales)
library(strucchange)
library(httpgd)
```

## Load data
```{r}
spread <- read_excel("/Users/mattia/Desktop/github/BTP-BUND_Spread_R/Forecast_8_final/var_data/spread_daily_24_04.xlsx")
eur1m <- read_excel("/Users/mattia/Desktop/github/BTP-BUND_Spread_R/Forecast_8_final/var_data/EURDOLL1M_24_04.xlsx")
ois3m <- read_excel("/Users/mattia/Desktop/github/BTP-BUND_Spread_R/Forecast_8_final/var_data/OIS3M_24_04.xlsx")
euribor3m <- read_excel("/Users/mattia/Desktop/github/BTP-BUND_Spread_R/Forecast_8_final/var_data/euribor_3M_24_04.xlsx")
inflexp <- read_excel("/Users/mattia/Desktop/github/BTP-BUND_Spread_R/Forecast_8_final/var_data/INFLEXP_24_04.xlsx")

data <- spread |>
  left_join(eur1m, by = "Date") |>
  left_join(ois3m, by = "Date") |>
  left_join(euribor3m, by = "Date") |>
  left_join(inflexp, by = "Date")

data <- data |> rename(spread = names(data)[2],
                       eur1m = names(data)[3],
                       ois3m = names(data)[4],
                       euribor3m = names(data)[5],
                       inflexp = names(data)[6])

data$spread <- data$spread * -1
data$eurois3m <- data$euribor3m - data$ois3m
data <- data |> dplyr::select(Date, spread, eur1m, eurois3m)
data <- xts(data[-1], order.by = as.Date(data$Date, format = "%Y-%m-%d"))
data <- xts::last(data, "5 years") |> na.omit()
```

```{r, echo=FALSE, fig.width=12, fig.height=10, out.width="100%"}
df_plot <- data.frame(date = index(data), coredata(data))

p_spread <- ggplot(df_plot, aes(x=date, y=spread)) +
  geom_line() +
  labs(title="Spread", x="", y="") +
  theme_bw()

p_eur1m <- ggplot(df_plot, aes(x=date, y=eur1m)) +
  geom_line() +
  labs(title="Euro/Dollar 1 month expectations", x="", y="") +
  theme_bw()

p_eurois3m <- ggplot(df_plot, aes(x=date, y=eurois3m)) +
  geom_line() +
  labs(title="Euribor 3M - OIS3M", x="", y="") +
  theme_bw()

(p_spread + p_eur1m + p_eurois3m) + plot_layout(nrow=3, ncol=1)
```

# Data preparation
## Train/Test
Use only the last `year` years of data, then split.
```{r}
n_test  <- round(nrow(data) * 0.2)
n_train <- nrow(data) - n_test

train <- data[1:n_train]
test  <- data[(n_train + 1):nrow(data)]

cat("\nTraining observations:", n_train)
cat("\nTest observations:", n_test)
cat("\nTrain period:", as.character(index(train)[1]), "->", as.character(index(train)[nrow(train)]))
cat("\nTest period:", as.character(index(test)[1]), "->", as.character(index(test)[nrow(test)]),"\n")
```

## Stationarity
Check if raw data are stationary around a drift with ADF test, use `ur.df()`, then first-difference and check again
```{r}
check_stat <- function(series, type = c("adf", "kpss")) {
  # To run both just don't select the type
  df_list <- list()
  # ADF test
  if ("adf" %in% type) {
     for (test in c("none", "drift", "trend")) {
      adf <- ur.df(series, type = test, selectlags = "AIC")
      row <- data.frame(
        test_type = paste("ADF: ", test),
        teststat = adf@teststat[1],
        cval = adf@cval[1, "5pct"],
        result = ifelse(adf@teststat[1] < adf@cval[1, "5pct"], "Stationary", "Non-Stationary")
      )
      df_list[[length(df_list) + 1]] <- row
    }
  }
  # KPSS test
  if ("kpss" %in% type) {
    for (test in c("mu", "tau")) {
      kpss <- ur.kpss(series, type = test, lags = "short")
      row <- data.frame(
        test_type = paste("KPSS: ", ifelse(test == "mu", "const", "trend")),
        teststat = kpss@teststat,
        cval = kpss@cval[1, "5pct"],
        result = ifelse(kpss@teststat < kpss@cval[1, "5pct"], "Stationary", "Non-Stationary")
      )
      df_list[[length(df_list) + 1]] <- row
    }
  }
  df <- bind_rows(df_list)
  return(df)
}

cat("\nRaw data:\n")
for (name in names(train)){
  test_stat <- check_stat(train[, name], type = "adf")
  cat(name, ":\n")
  print(test_stat)
}

var_dif = c("spread", "eur1m")
d_data = data
d_data[, var_dif] = d_data[, var_dif] |> diff()
d_data = d_data |> na.omit()
d_train <- d_data[1:(n_train - 1)]
d_test  <- d_data[n_train:nrow(d_data)]

cat("\nTransformed data:\n")
for (name in names(d_train)){
  test_stat <- check_stat(d_train[, name], type = "adf")
  cat(name, ":\n")
  print(test_stat)
}
```

- `spread` series showed non-stationary behaviour, it does says "trend stationary" but looking at the series plot we can see that the supposed trend isn't linear,
for semplicity we decided to difference it. 
- We differenced ` eur1m` becasu, even if it pass the test, if not differenced it will cause problems in the model, 
the differenced series works better, also looking at the plot we can see that its stationarity is kinda false since it doesn't look like it.
- On the other side, and `eurois3m` is perfectly stationary. 

```{r, echo=FALSE, fig.width=12, fig.height=10, out.width="100%"}
d_df_plot <- data.frame(date = index(d_train), coredata(d_train))

p_spread <- ggplot(d_df_plot, aes(x=date, y=spread)) +
  geom_line() +
  labs(title="Spread", x="", y="") +
  theme_bw()

p_eur1m <- ggplot(d_df_plot, aes(x=date, y=eur1m)) +
  geom_line() +
  labs(title="Euro/Dollar 1 month expectations", x="", y="") +
  theme_bw()

p_eurois3m <- ggplot(d_df_plot, aes(x=date, y=eurois3m)) +
  geom_line() +
  labs(title="Euribor 3M - OIS3M", x="", y="") +
  theme_bw() 

(p_spread + p_eur1m + p_eurois3m) + plot_layout(nrow = 3, ncol = 1)
```

# VAR
## Estimation
Check if there's autocorrelation between variables
```{r}
mq(d_train, lag = 10)
```

Test shows there's autocorrelation present across variables

Check if variables granger-cause `spread`

```{r, warning=FALSE}
var_names = names(d_train)

print(var_names)

for (v in var_names[var_names != "spread"]) {
  cat("\n", v, "\n")
  print(lmtest::grangertest(d_train[,v], d_train[,"spread"], order = 5))
}
```

Both granger-cause `spread`, even if `eur1m` slighty

Find the optimal lag
```{r}
selection = vars::VARselect(y = d_train, lag.max = 10,
                            type = "const")

print(t(selection$criteria))
print(selection$selection)
```

Following the AIC we will estimate a VAR(5).
```{r}
model = vars::VAR(d_train, lag.max = 10, ic = "AIC", type = "const")
summary(model)
```

Now we simplify the model eliminating non-significant parameters, with t-ratio < 1.64 $(\alpha = 10\%)$
```{r}
model_s = vars::restrict(model, method = "ser", thresh = 1.64)
fixed_p = model_s$p
fixed_params = model_s$restrictions
summary(model_s)
```

```{r}
count_coeff = function(model){
  names = names(model$varresult)
  n = c()
  for (name in names){
    n = c(n, length(model$varresult[[name]]$coefficients))
  }
  sum(n)
}

cat("\nUnrestricted model coeff: ", count_coeff(model),
    " Restricted model coeff: ", count_coeff(model_s), "\n")
cat("\nUr model SC: ", BIC(model), "R model SC: ", BIC(model_s))
cat("\nUr model AIC: ", AIC(model), "R model AIC: ", AIC(model_s))
```

### Diagnostic test
#### Residuals autocorrelation
```{r, fig.width=12, fig.height=10, out.width="100%"}
par(ask = FALSE)
plot(vars::serial.test(model_s, lags.pt = 16, type = "PT.adjusted"))
```

Run a Portmanteau test $H_0: \text{No cross-correlation}$ to test for autocorrelatino in residuals
```{r}
adj_var = (nrow(coef(model_s)$spread) + nrow(coef(model_s)$eur1m) + nrow(coef(model_s)$eurois3m))

res = residuals(model_s)
mq(res, adj = adj_var)
```

We fail reject $H_0$, indicating the presence of no autocorrelation in residuals, apart from some rejection in the further away lags

Then we test for ARCH effects

### ARCH test
```{r}
vars::arch.test(model_s, lags.single = 16, lags.multi = 5, multivariate.only = FALSE)
```

We strongly reject the null of no ARCH effects, a GARCH modelling is required. 

Test for stability in the residuals series with the CUSUM test, if the series cross the 5% significance level
the parameters are unstable

### Stability test
We run CUSUM test to check for possible structural breaks
```{r, echo=FALSE, fig.width=12, fig.height=10, out.width="100%"}
par(ask = FALSE)
plot(vars::stability(model_s))
```

All variables residuals lay around 0, even if they slighty drift. There's no evidence of structural breaks

#### Normality
Check if residuals are normal distributed, if not try to fit them to a t-distribution (more likely).

Use a KS test to test wheter they follow a t-distribution $H_0: \text{serie follows supposed distribution}$
```{r, warning=FALSE}
# p_value < 0.05 -> non-normal distributed
normality.test(model_s, multivariate.only = FALSE)

check_dist <- function(residuals) {
  t = suppressWarnings(MASS::fitdistr(residuals, densfun = "t",
        start = list(m = mean(residuals), s = sd(residuals), df = 5)))  # <- add this
  n = MASS::fitdistr(residuals, densfun = "normal")
  aic = c(AIC(t), AIC(n))
  bic = c(BIC(t), BIC(n))
  df = data.frame(aic, bic, row.names = c("t", "n"))
  df
}

apply(residuals(model_s), 2, check_dist)
```

We reject the null of the JB test for all the univariate series and the joint multivariate, thus the residuals do not follow a normal distribution.

The subsequent code, consisting in estimating a normal a t distribution on the residuals and then compute their IC, showed that a t-distribution 
is a better fit than the normal.

#### Causality
```{r}
vars::causality(model_s, cause = c("eur1m", "eurois3m"))
vars::causality(model_s, cause = c("spread", "eurois3m"))
vars::causality(model_s, cause = c("eur1m", "spread"))
```

In the restricted model the granger-causality disappear ???

### Impulse Response Functions IRF
```{r, echo=FALSE, fig.width=12, fig.height=10, out.width="100%"}
par(ask = FALSE)
irf_result <- vars::irf(model_s,
                        impulse   = c("eur1m", "eurois3m"),
                        response  = "spread",
                        n.ahead   = 20,
                        boot      = TRUE,
                        ci        = 0.95,
                        runs      = 500)
plot(irf_result)
```

### Forecast Error Variance Decomposition FEVD
```{r, echo=FALSE, fig.width=12, fig.height=10, out.width="100%"}
par(ask = FALSE)
fevd_result <- vars::fevd(model_s, n.ahead = 20)
plot(fevd_result)
```

# GARCH
For the mulivariate GARCH we use the DCC way

Then i re-run the same tests to see it the model is better.
```{r}
res = residuals(model_s)

uni_spec = ugarchspec(
  variance.model = list(model = "sGARCH", garchOrder = c(1,1)),
  mean.model = list(armaOrder = c(0,0), include.mean = FALSE),
  distribution.model = "std"
)

multi_spec = multispec(replicate(ncol(res), uni_spec))

dcc_spec = dccspec(uspec = multi_spec, dccOrder = c(1,1), distribution = "mvt")

dcc_fit = dccfit(dcc_spec, data = res)
```

```{r, echo=FALSE, fig.width=12, fig.height=10, out.width="100%"}
dcc_corr = rcor(dcc_fit)

corr_df = data.frame(date = as.Date(dimnames(dcc_corr)[[3]]),
                     spread_eur1m = dcc_corr[1, 2, ],
                     spread_eurois3m = dcc_corr[1, 3, ],
                     eur1m_eurois3m = dcc_corr[2, 3, ],
                     row.names = NULL)

ggplot(corr_df, aes(x=date)) +
  geom_line(aes(y=spread_eur1m, col="spread : eur1m")) +
  geom_line(aes(y=spread_eurois3m, col="spread : eurois3m")) +
  geom_line(aes(y=eur1m_eurois3m, col="eur1m : eurois3m")) +
  labs(title="Dynamic Conditional Correlations", x="", y="", col = "Pairs") +
  theme_bw() +
  theme(legend.position = "bottom") 
```

Check for residuals ARCH effect on DCC standard residuals
```{r}
std_res <- residuals(dcc_fit) / sigma(dcc_fit)
MTS::MarchTest(std_res, lag = 10)
```

Fail to reject the null, the series do not have any volatility clustering anymore.

Check for distributional assumptions. 
```{r, warning=FALSE}
apply(std_res, 2, check_dist)
```

They appear to better follow a t-distribution

No ARCH effects and right distributional assumptions -> Model is correctly specified.

# Validation
Validate the model via rolling forecast on test data, the window dimension is set to the same window used for 
the diagnostic, we keep the same structure we tested but re-estimating the coefficients at every iteration, 
we can do that by fixing the parameter when restricting the VAR model.
Then transform back to spread and evaluate
```{r}
start_time <- proc.time()

invert_diff <- function(d, last) last + d

window_size <- nrow(d_train)
len         <- nrow(d_test)

df <- data.frame(
  date = rep(as.Date(NA), len),
  act_spread  = numeric(len), act_eur1m      = numeric(len), act_eurois3m   = numeric(len),
  fc_spread   = numeric(len), fc_spread_lo   = numeric(len), fc_spread_up   = numeric(len),
  fc_eur1m    = numeric(len), fc_eur1m_lo    = numeric(len), fc_eur1m_up    = numeric(len),
  fc_eurois3m = numeric(len), fc_eurois3m_lo = numeric(len), fc_eurois3m_up = numeric(len)
)

for (i in 1:len) {
  start <- n_train - window_size + i - 1
  end   <- start + window_size - 1
  window_data <- d_data[start:end, ]

  # Last value, d_data, end_data lost one row
  orig_idx <- end + 1
  last_spread  <- as.numeric(data[orig_idx, "spread"])
  last_eur1m <- as.numeric(data[orig_idx, "eur1m"])

  df$act_spread[i]  <- as.numeric(data[orig_idx + 1, "spread"])
  df$act_eur1m[i] <- as.numeric(data[orig_idx + 1, "eur1m"])
  df$act_eurois3m[i] <- as.numeric(data[orig_idx + 1, "eurois3m"])

  # VAR
  roll_var <- vars::VAR(window_data, p = fixed_p, type = "const")
  roll_var_s <- vars::restrict(roll_var, method = "manual", resmat = fixed_params)
  fc <- predict(roll_var_s, n.ahead = 1)

  # DCC 
  roll_res <- residuals(roll_var_s)
  roll_dcc_fit <- dccfit(dcc_spec, data = roll_res, solver = c("hybrid", "nlminb"),
                         fit.control = list(eval.se = FALSE))
  dcc_fc  <- dccforecast(roll_dcc_fit, n.ahead = 1)
  dof  <- roll_dcc_fit@mfit$coef["[Joint]mshape"]
  t_sc <- qt(p = 0.975, df = dof)

  # Extract
  d_spread  <- fc$fcst$spread[1,  "fcst"]
  d_eur1m  <- fc$fcst$eur1m[1,  "fcst"]
  d_eurois3m  <- fc$fcst$eurois3m[1,  "fcst"]

  sigma <- rcov(dcc_fc)
  se    <- sqrt(diag(sigma[[1]][, , 1]))

  # Point forecasts
  df$date[i]       <- index(d_test)[i]
  df$fc_spread[i]  <- invert_diff(d_spread, last_spread)
  df$fc_eur1m[i]  <- invert_diff(d_eur1m, last_eur1m)
  df$fc_eurois3m[i]  <- d_eurois3m

  # Confidence intervals
  df$fc_spread_lo[i]  <- invert_diff(d_spread  - t_sc * se[1], last_spread)
  df$fc_spread_up[i]  <- invert_diff(d_spread  + t_sc * se[1], last_spread)
  df$fc_eur1m_lo[i]  <- invert_diff(d_eur1m  - t_sc * se[2], last_eur1m)
  df$fc_eur1m_up[i]  <- invert_diff(d_eur1m  + t_sc * se[2], last_eur1m)
  df$fc_eurois3m_lo[i]  <- d_eurois3m  - t_sc * se[3]
  df$fc_eurois3m_up[i]  <- d_eurois3m  + t_sc * se[3]

  if (i %% 25 == 0 || i == len) cat("Step", i, "/", len, "\n")
}

end_time <- proc.time()
execution_time <- end_time - start_time
cat("\n Total execution time: ", execution_time[3], "seconds")
```

Metrics
```{r}
rmse <- function(actual, forecast) sqrt(mean((actual - forecast)^2, na.rm = TRUE))

acc_dir <- function(actual, forecast) {
  d_actual = diff(actual)
  d_forecast = forecast[-1] - actual[-length(actual)]
  dir = mean(sign(d_forecast) == sign(d_actual), na.rm = TRUE) * 100
}

ci_coverage <- function(actual, lo, up) mean(actual >= lo & actual <= up, na.rm = TRUE) * 100

results <- data.frame(
  variable    = var_names,
  rmse        = c(
    rmse(df$act_spread,  df$fc_spread),
    rmse(df$act_eur1m,  df$fc_eur1m),
    rmse(df$act_eurois3m, df$fc_eurois3m)
  ),
  acc_dir    = c(
    acc_dir(df$act_spread, df$fc_spread),
    acc_dir(df$act_eur1m, df$fc_eur1m),
    acc_dir(df$act_eurois3m, df$fc_eurois3m)
  ),
  ci_coverage_0.95 = c(
    ci_coverage(df$act_spread,  df$fc_spread_lo,  df$fc_spread_up),
    ci_coverage(df$act_eur1m,  df$fc_eur1m_lo,  df$fc_eur1m_up),
    ci_coverage(df$act_eurois3m, df$fc_eurois3m_lo, df$fc_eurois3m_up)
  )
)

print(results)
```

## Compare
I compare the forecast evaluation metrics of the model with a Random Walk model (tomorrow's value = today's value).
```{r}
spread_rw <- numeric(len)

for (i in 1:len) {
  end <- n_train + i
  spread_rw[i] <- as.numeric(data[n_train + i - 1, "spread"])
}

df$spread_rw <- spread_rw

rmse_rw = rmse(df$act_spread, df$spread_rw)
acc_dir_rw = acc_dir(df$act_spread, df$spread_rw)
cat("\nRMSE RW: ", rmse_rw)
cat("\nDirection accuracy RW: ", acc_dir_rw)
```

Model predict the spread like a Random Walk
```{r, fig.width=12, fig.height=10, out.width="100%"}
p_spreads  <- ggplot(df, aes(x = date)) +
  geom_line(aes(y = act_spread, color = "Actual"), linewidth = 0.8) +
  geom_line(aes(y = fc_spread, color = "Forecasted"), linewidth = 0.8) +
  geom_line(aes(y = spread_rw, color = "Random Walk"), linewidth = 0.8) +
  geom_ribbon(aes(ymin = fc_spread_lo, ymax = fc_spread_up), fill = "blue", alpha = 0.3) +
  scale_color_manual(values = c("Actual" = "black", "Forecasted" = "blue", "Random Walk" = "red")) +
  labs(title = "Spread: Actual vs Forecasted", x = "", y = "Basis points") +
  scale_x_date(date_breaks = "2 month", date_labels = "%b-%y") +
  scale_y_continuous(breaks = scales::breaks_width(20)) +
  theme_minimal() +
  theme(legend.title = element_blank(), legend.position = "bottom")

print(p_spreads)
```

# Forecast
Next week forecast
```{r}
data_fit = tail(d_data, window_size) 
last_date  <- last(index(data_fit))
last_spread <- last(data$spread)

fc_var <- vars::VAR(window_data, p = fixed_p, type = "const")
fc_var_s <- vars::restrict(roll_var, method = "manual", resmat = fixed_params)
fc <- predict(roll_var_s, n.ahead = 1)

fc_res <- residuals(roll_var_s)
fc_dcc_fit <- dccfit(dcc_spec, data = roll_res)
fc_dcc  <- dccforecast(roll_dcc_fit, n.ahead = 1)
dof  <- fc_dcc_fit@mfit$coef["[Joint]mshape"]
t_sc <- qt(p = 0.975, df = dof)
sigma <- rcov(fc_dcc)
se    <- sqrt(diag(sigma[[1]][, , 1]))

d_spread  <- fc$fcst$spread[1,  "fcst"]
fc_spread  <- invert_diff(d_spread, last_spread)
fc_spread_lo  <- invert_diff(d_spread  - t_sc * se[1], last_spread)
fc_spread_up  <- invert_diff(d_spread  + t_sc * se[1], last_spread)

forecast <- data.frame(
  spread = fc_spread |> as.numeric(),
  q_05 = fc_spread_lo |> as.numeric(),
  q_95 = fc_spread_up |> as.numeric()
)

cat("\nForecasted values:\n")
cat("\nSpread on last date: ", as.character(last_date), ": ", last_spread)
cat("\nSpread forecasted for tomorrow:", as.character(last_date + 1), ":\n")
print(forecast)
```